---
title: "Overview of discovery and validation cohorts"
author: "Tiangen Chang <tiangen.chang@nih.gov>"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Overview of discovery and validation cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "markup", warning = FALSE, message = FALSE)
```


####### This script makes the below displays for the paper
## Figure 1a-e
## SFigure 1a-b

# Package Loading
```{r}
library(readr)
library(readxl)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(survival)
library(survminer)
```

# Parameters and shared functions
```{r}
result_figure_dir <- "./Figures/"
data_dir <- "./SourceData/"

color_clusters <- c(
  "#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72", "#B17BA6",
  "#FF7F00", "#FDB462", "#E7298A", "#E78AC3", "#33A02C", "#B2DF8A",
  "#55A1B1", "#8DD3C7", "#A6761D", "#E6AB02", "#7570B3", "#BEAED4",
  "#666666", "#999999", "#aa8282", "#d4b7b7", "#8600bf", "#ba5ce3",
  "#808000", "#aeae5c", "#1e90ff", "#00bfff", "#56ff0d", "#ffff00",
  "#8B3800", "#4E8500", "#D70131", "#E64B35B2", "#4DBBD5B2",
  "#00A087B2", "#3C5488B2", "#F39B7FB2", "#4E8500", "#4C00FFFF",
  "#000DFFFF", "#0068FFFF", "#00C1FFFF", "#00FF24FF", "#42FF00FF",
  "#A8FF00FF", "#FFF90CFF", "#FFE247FF", "#FFDB83FF", "#4a5b67",
  "#91D1C2B2", "#DC0000B2", "#7E6148B2", "#B09C85B2", "#15b2d3",
  "#236e96", "#ffd700", "#f3872f", "#ff598f", "#4B0082", "#00FA9A",
  "#FFD700", "#9400D3", "#FF4500", "#DA70D6", "#8B0000", "#5F9EA0",
  "#7FFF00", "#6495ED", "#DB7093", "#FF6347", "#4682B4", "#FF1493",
  "#FFDAB9", "#ADFF2F", "#006400", "#00CED1", "#800080", "#FF69B4",
  "#1E90FF", "#B22222", "#8B4513", "#A52A2A", "#0000FF", "#00FF7F",
  "#FFB6C1", "#32CD32", "#808080", "#D2691E", "#4169E1", "#DAA520",
  "#FF4500", "#98FB98", "#CD5C5C", "#2F4F4F", "#00008B", "#8A2BE2",
  "#CD853F", "#20B2AA", "#DC143C"
)

custom_colors <- c(
  "#1c4f27", "#81144e", "#79831e", "#00305d", "#C0C0C0",
  "#9c1915", "black", "#404040", "#808080", "#D3D3D3"
)

theme_clean_axis <- function(base_size = 11, x_angle = 0) {
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    panel.border = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black", size = base_size),
    axis.text.x = element_text(angle = x_angle, hjust = ifelse(x_angle == 0, 0.5, 1)),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key = element_rect(fill = NA, color = NA)
  )
}

save_pdf <- function(filename, plot, width, height) {
  pdf(file.path(result_figure_dir, filename), width = width, height = height, onefile = FALSE)
  print(plot)
  dev.off()
}

plot_pie <- function(data, suffix, colors = custom_colors) {
  p <- ggplot(data, aes(x = "", y = Count, fill = Category)) +
    geom_col(width = 1) +
    coord_polar(theta = "y") +
    scale_fill_manual(values = colors) +
    theme_void() +
    theme(legend.position = "right") +
    guides(fill = guide_legend(ncol = 1, byrow = FALSE)) +
    labs(title = NULL, fill = NULL)

  save_pdf(paste0("piePlot_", suffix, ".pdf"), p, width = 2 * 0.6 * 3, height = 2 * 0.4 * 3)
}

plot_binary_rate <- function(data, group_var, group_labels, filename, y_limit = c(0, 50)) {
  plot_df <- data %>%
    select(group = all_of(group_var), irAE) %>%
    tidyr::drop_na() %>%
    mutate(group = factor(group, levels = names(group_labels), labels = group_labels))

  ratio_df <- plot_df %>%
    group_by(group) %>%
    summarise(irAE_Ratio = mean(irAE) * 100, .groups = "drop")

  p <- ggplot(ratio_df, aes(x = group, y = irAE_Ratio, fill = group)) +
    geom_col(width = 0.6, alpha = 0.8, color = "black") +
    coord_cartesian(ylim = y_limit) +
    labs(x = NULL, y = "Severe-irAE rate (%)", title = NULL) +
    theme_clean_axis(x_angle = 45) +
    theme(legend.position = "none")

  save_pdf(filename, p, width = 1.1, height = 2.5)
  print(chisq.test(table(plot_df$group, plot_df$irAE)))
}

plot_continuous_by_irae <- function(data, value_var, y_label, filename) {
  plot_df <- data %>%
    select(value = all_of(value_var), irAE = IRAEactionable01) %>%
    tidyr::drop_na() %>%
    mutate(irAE = factor(ifelse(irAE == 1, "Severe irAE", "Non-severe irAE"),
                         levels = c("Non-severe irAE", "Severe irAE")))

  p <- ggplot(plot_df, aes(x = irAE, y = value, fill = irAE)) +
    geom_jitter(width = 0.15, size = 1, alpha = 0.8, color = "grey") +
    geom_boxplot(width = 0.2, outlier.shape = NA, fill = NA, color = "black") +
    labs(x = NULL, y = y_label, title = NULL) +
    theme_clean_axis(x_angle = 45) +
    theme(legend.position = "none")

  save_pdf(filename, p, width = 1.2, height = 2.5)
  print(wilcox.test(value ~ irAE, data = plot_df))
}

```

# Figure 1a-b (middle and right panels). UMAP and heatmap.
```{r}
set.seed(1)

file_prefix <- "HCI001" # HCI001 or HCI002_baseline 

umap_df <- readRDS(file.path(data_dir, paste0("Umap_", file_prefix, ".rds")))
umap_df <- umap_df[sample(nrow(umap_df), min(10000, nrow(umap_df))), ]

centroids <- umap_df %>%
  group_by(cell_type) %>%
  summarise(Umap1 = median(Umap1), Umap2 = median(Umap2), .groups = "drop")

p_umap <- ggplot(umap_df, aes(Umap1, Umap2, color = cell_type)) +
  geom_point(size = 0.1) +
  geom_text_repel(data = centroids, aes(label = cell_type), color = "black") +
  scale_color_manual(values = color_clusters) +
  labs(title = NULL, x = "UMAP1", y = "UMAP2") +
  guides(color = guide_legend(ncol = 1, override.aes = list(size = 3), title = NULL)) +
  theme_clean_axis() +
  theme(legend.key.height = unit(0.8, "lines"))

save_pdf(paste0(file_prefix, "_UMAP_CellTypes.pdf"), p_umap, width = 14 * 0.4, height = 10 * 0.4)

scaled_data <- read.csv(file.path(data_dir, paste0("scaled_expr_", file_prefix, ".csv")), row.names = 1)

pdf(file.path(result_figure_dir, paste0(file_prefix, "_heatmap_CellTypes.pdf")),
    width = 14 * 0.4, height = 8 * 0.4)
pheatmap(
  scaled_data,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  fontsize_row = 9,
  fontsize_col = 9,
  main = ""
)
dev.off()

```

# Load data (and sample size statistics for Figure 1a-b left panels)
```{r}
HCI002_all_info0 <- read_csv(file.path(data_dir, "HCI002_all_info.csv"), name_repair = "minimal")
HCI002_map_df <- read.csv(file.path(data_dir, "HCI002_OlinkID_ProteinName_map.csv"))

HCI001_all_info0 <- read_csv(file.path(data_dir, "HCI001_all_info.csv"), name_repair = "minimal")
HCI001_map_df <- read.csv(file.path(data_dir, "HCI001_OlinkID_ProteinName_map.csv"))

hci001_samples <- sub("_.*", "", HCI001_all_info0$`RedCap ID`)
HCI001_all_info0$patient_therapy <- paste(
  hci001_samples,
  HCI001_all_info0$`Current Therapy`,
  HCI001_all_info0$`Type of therapy`,
  sep = "_"
)

summarise_available_samples <- function(data, sample_col, marker_cols) {
  df <- data.frame(sample = data[[sample_col]], data[marker_cols], check.names = FALSE)
  df <- df[rowSums(is.na(df)) < length(marker_cols), ]
  list(
    n_unique_samples = length(unique(df$sample)),
    n_non_missing = colSums(!is.na(df[marker_cols])),
    samples = unique(df$sample)
  )
}

hci001_summary <- summarise_available_samples(
  HCI001_all_info0,
  "patient_therapy",
  c(
    "CD4_nonTREG > CD38_Baseline",
    "OID21336_OLINK_baseline",
    "CD4_nonTREG > CD38_irAEmax",
    "OID21336_OLINK_irAEmax"
  )
)

hci002_summary <- summarise_available_samples(
  HCI002_all_info0,
  "patient_therapy",
  c(
    "CD4_nonTREG_FC_baselineL2",
    "OID21336_OLINK_baseline",
    "CD4_nonTREG_FC_ontreatL2",
    "OID21336_OLINK_ontreat"
  )
)

HCI001_uniqueSamples <- hci001_summary$samples
HCI002_uniqueSamples <- hci002_summary$samples

length(unique(hci001_samples))
length(unique(HCI001_all_info0$patient_therapy))
hci001_summary$n_unique_samples
hci001_summary$n_non_missing

length(unique(HCI002_all_info0$patient_therapy))
length(unique(HCI002_all_info0$`Record ID`))
hci002_summary$n_unique_samples
hci002_summary$n_non_missing

```

# Figure 1c. Pie plots.
```{r}
empty_names <- is.na(names(HCI002_all_info0)) | names(HCI002_all_info0) == ""
names(HCI002_all_info0)[empty_names] <- paste0("unnamed_col_", which(empty_names))

all_info <- HCI002_all_info0 %>%
  filter(patient_therapy %in% HCI002_uniqueSamples) %>%
  select(1:60) %>%
  distinct(patient_therapy, .keep_all = TRUE)

plot_pie(data.frame(Category = c("Response (n = 86)", "Non-response (n = 71)"), Count = c(86, 71)), "Response")
plot_pie(data.frame(Category = c("Severe-irAE (n = 55)", "Non-severe-irAE (n = 102)"), Count = c(55, 102)), "irAE")
plot_pie(data.frame(Category = c("ICB naive (n = 97)", "ICB experienced (n = 62)"), Count = c(97, 62)), "priorICB")

ctla4_therapies <- c(
  "Nivolumab/ Ipilimumab", "PD1-CTLA4 bispecific", "Ipi 10 / temozolomide",
  "Nivolumab/ Ipilimumab flipped dose", "Ipi 3 / temozolomide",
  "Pembrolizumab / CTLA-4-LAG3 bispecific", "CTLA-4-LAG3 bispecific",
  "Pembrolizumab/ anti-CTLA4", "anti-CTLA4"
)

all_info <- all_info %>%
  mutate(
    Drug = ifelse(`Systemic Therapy Name` %in% ctla4_therapies, "CTLA4-containing", "PD1/PDL1"),
    Stage = case_when(
      grepl("Stage I", `Baseline Stage / TNM`) ~ "Stage I",
      grepl("Stage II", `Baseline Stage / TNM`) ~ "Stage II",
      grepl("Stage III", `Baseline Stage / TNM`) ~ "Stage III",
      grepl("Stage IV", `Baseline Stage / TNM`) ~ "Stage IV",
      TRUE ~ `Baseline Stage / TNM`
    )
  )

plot_pie(data.frame(Category = c("anti-PD1/PDL1 (n = 118)", "anti-CTLA4-containing (n = 41)"), Count = c(118, 41)), "therapy")
plot_pie(data.frame(Category = c("Stage II (n = 5)", "Stage III (n = 66)", "Stage IV (n = 88)"), Count = c(5, 66, 88)), "stage")
plot_pie(data.frame(Category = c("Female (n = 63)", "Male (n = 96)"), Count = c(63, 96)), "sex")

```

# Figure 1d. Severe-irAE subtype bar plots.
```{r}
all_info <- all_info %>%
  mutate(
    IRAE_other01 = case_when(
      is.na(IRAEactionable01) ~ NA_real_,
      IRAEactionable01 == 1 &
        (is.na(IRAE_GI01) | IRAE_GI01 == 0) &
        (is.na(IRAE_HB01) | IRAE_HB01 == 0) &
        (is.na(IRAE_skin01) | IRAE_skin01 == 0) &
        (is.na(IRAE_MSK01) | IRAE_MSK01 == 0) ~ 1,
      TRUE ~ 0
    )
  )

subtype_df <- data.frame(
  group = c("Gastrointestinal", "Hepatobiliary", "Musculoskeletal", "Skin", "Other"),
  phenotype = c(
    mean(all_info$IRAE_GI01, na.rm = TRUE),
    mean(all_info$IRAE_HB01, na.rm = TRUE),
    mean(all_info$IRAE_MSK01, na.rm = TRUE),
    mean(all_info$IRAE_skin01, na.rm = TRUE),
    mean(all_info$IRAE_other01, na.rm = TRUE)
  ) * 100
)

subtype_df$group <- factor(subtype_df$group, levels = subtype_df$group)

p_subtype <- ggplot(subtype_df, aes(group, phenotype, fill = group)) +
  geom_col(color = "black", width = 0.6) +
  coord_cartesian(ylim = c(0, 10)) +
  labs(x = NULL, y = "Severe-irAE rate (%)", title = NULL) +
  theme_clean_axis(x_angle = 45) +
  theme(legend.position = "none")

save_pdf("barplot_severe_irAE_subtype_ratio.pdf", p_subtype, width = 1.5, height = 2.5)

```

# Figure 1d. Categorical associations with severe irAE.
```{r}
clinical_df <- data.frame(
  PDL1 = all_info$`PD-L1 CPS`,
  Response = all_info$Response01,
  Sex = all_info$Gender,
  irAE = all_info$IRAEactionable01,
  check.names = FALSE
) %>%
  mutate(
    PDL1_group = ifelse(PDL1 < 1, "PDL1 CPS-", "PDL1 CPS+"),
    Response_group = ifelse(Response < 1, "Non-response", "Response"),
    Sex_group = ifelse(Sex == "Male", "Male", "Female")
  )

plot_binary_rate(
  clinical_df,
  "PDL1_group",
  c("PDL1 CPS-" = "PDL1 CPS-", "PDL1 CPS+" = "PDL1 CPS+"),
  "barplot_PDL1_vs_irAE.pdf"
)

plot_binary_rate(
  clinical_df,
  "Response_group",
  c("Non-response" = "Non-response", "Response" = "Response"),
  "barplot_Response_vs_irAE.pdf"
)

plot_binary_rate(
  clinical_df,
  "Sex_group",
  c("Male" = "Male", "Female" = "Female"),
  "barplot_Sex_vs_irAE.pdf"
)

```

# Figure 1d. Age/TMB vs severe irAE group box plots.
```{r}
plot_continuous_by_irae(all_info, "Age at C1", "Age (years)", "boxplot_age_vs_irAE.pdf")
plot_continuous_by_irae(all_info, "TMB (mutation /mb)", "TMB (mut. / Mb)", "boxplot_TMB_vs_irAE.pdf")

```

# Figure 1e and SFigure 1a-b. Kaplan–Meier survival curves.
```{r}
survival_info_df0 <- read_excel(file.path(data_dir, "HCI002_clinical_info.xlsx")) %>%
  select(
    `Vial ID`, `Record ID`, `Systemic Therapy Name`, `Systemic Therapy Type`,
    OS_time, OS_event, PFS_time, PFS_event,
    Nonsevere_irAE_time, Nonsevere_irAE_event,
    `irAE Overall - Actionable`, `Best/ Worst Response`
  ) %>%
  mutate(
    uniqueID = paste(`Record ID`, `Systemic Therapy Name`, `Systemic Therapy Type`, sep = "_"),
    IRAEactionable01 = case_when(
      `irAE Overall - Actionable` == "Severe" ~ 1,
      `irAE Overall - Actionable` %in% c("Mild", "Moderate", "None") ~ 0,
      TRUE ~ NA_real_
    ),
    Response01 = case_when(
      `Best/ Worst Response` %in% c("CR", "PR", "NED") ~ 1,
      `Best/ Worst Response` %in% c("SD", "PD") ~ 0,
      TRUE ~ NA_real_
    ),
    OS_time = OS_time / 30,
    PFS_time = PFS_time / 30,
    Nonsevere_irAE_time = Nonsevere_irAE_time / 30
  ) %>%
  filter(uniqueID %in% HCI002_uniqueSamples) %>%
  distinct(uniqueID, .keep_all = TRUE)

plot_survival_curve <- function(data, therapy_type, phenotype, group_var) {
  therapy_data <- switch(
    therapy_type,
    active = filter(data, `Systemic Therapy Type` == "active treatment"),
    adjuvant = filter(data, `Systemic Therapy Type` == "adjuvant"),
    neoadjuvant = filter(data, `Systemic Therapy Type` == "neoadjuvant"),
    All = data
  )

  y_label <- switch(
    phenotype,
    OS = "Overall \nsurvival probability",
    PFS = "Progression-free \nsurvival probability",
    Nonsevere_irAE = "Severe-irAE-free \nsurvival probability"
  )

  groups <- if (group_var == "IRAEactionable01") {
    c(case = "Severe-irAE", control = "Non-severe-irAE")
  } else {
    c(case = "Response", control = "Non-response")
  }

  cancer_data <- data.frame(
    Score = ifelse(therapy_data[[group_var]] > 0.5, groups["case"], groups["control"]),
    time = therapy_data[[paste0(phenotype, "_time")]],
    event = therapy_data[[paste0(phenotype, "_event")]]
  ) %>%
    tidyr::drop_na() %>%
    mutate(Score = relevel(factor(Score), ref = groups["control"]))

  sfit <- survfit(Surv(time, event) ~ Score, data = cancer_data)
  scox <- coxph(Surv(time, event) ~ Score, data = cancer_data)
  cox_summary <- summary(scox)

  hr <- cox_summary$coefficients[2]
  p_value <- cox_summary$coefficients[5]
  hr_ci <- cox_summary$conf.int[3:4]

  surv_plot <- ggsurvplot(
    sfit,
    data = cancer_data,
    size = 1,
    palette = c("#00305d", "#9c1915"),
    conf.int = FALSE,
    pval = FALSE,
    xlim = c(-3, 70),
    ylim = c(0, 1),
    xlab = "Time (months)",
    ylab = y_label,
    break.time.by = 10,
    risk.table = TRUE,
    risk.table.height = 0.25,
    risk.table.pos = "out",
    risk.table.col = "black",
    risk.table.y.text = F,
    tables.y.text = F,
    tables.theme = theme_cleantable(),
    legend.labs = c(groups["control"], groups["case"]),
    legend.title = NULL,
    legend = c(0.65, 0.975),
    font.main = 12,
    font.caption = 12,
    font.legend = 12,
    font.tickslab = 12,
    font.x = 12,
    font.y = 12,
    ggtheme = theme_clean_axis()
  )

  surv_plot$plot <- surv_plot$plot +
    annotate(
      "text",
      x = 0,
      y = 0.15,
      hjust = 0,
      size = 5,
      label = paste0(
        "HR = ", round(hr, 2),
        " (", round(hr_ci[1], 2), "-", round(hr_ci[2], 2), ")",
        "\np = ", sprintf("%.3f", p_value)
      )
    )

  filename <- paste0(
    "KM_curve_", therapy_type, "_", phenotype, "_vs_",
    groups["case"], "_", groups["control"], ".pdf"
  )

  pdf(file.path(result_figure_dir, filename), width = 3.3, height = 3, onefile = FALSE)
  print(surv_plot, newpage = FALSE)
  dev.off()
}

for (group_var in c("Response01", "IRAEactionable01")) {
  for (therapy_type in c("active", "All", "adjuvant")) {
    for (phenotype in c("OS", "PFS", "Nonsevere_irAE")) {
      plot_survival_curve(survival_info_df0, therapy_type, phenotype, group_var)
    }
  }
}

```

