---
title: "Correlate CyTOF and OLINK features on HCI002"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

####### This script makes the below displays for the paper
## Figure 1F-G
## Figure 3F
## SFigure 1C-D
## SFigure 5H

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, results = 'markup')
```

# load required package
```{r}

library(data.table)
library(ggplot2)
library(readxl)
library(viridis)
library(tidyr)
library(pROC)
library(verification)
library(reshape2)
library(ggpubr)

library(dplyr)
library(survival)
library(survminer)

library(ggrepel) 
library(glmnet)
library(caret)
library(pROC)
library(ranger)
library(readr)

library(AnnotationDbi)
library(org.Hs.eg.db) 
library(ReactomePA)
library(enrichplot)
library(gridExtra)
library(clusterProfiler)
library(msigdbr)

library(VennDiagram)
library(tidyverse)
library(igraph)
library(ggraph)
library(corrr)

library(RColorBrewer)
library(msigdbr)
library(ComplexHeatmap)
library(circlize)
library(patchwork)

```

# common functions and parameters
```{r}

# Create a function to make olink gene names unique
make_unique <- function(names) {
  # Initialize an empty vector to store unique names
  unique_names <- character(length(names))
  # Create a named vector to count occurrences
  name_count <- table(names)
  # Track the current count of each name
  name_tracker <- setNames(rep(0, length(name_count)), names(name_count))
  # Iterate over the names and create unique ones
  for (i in seq_along(names)) {
    name <- names[i]
    # If this name is duplicated, append the appropriate number
    if (name_tracker[name] > 0) {
      new_name <- paste0(name, "_", name_tracker[name] + 1)
    } else {
      new_name <- name
    }
    # Assign the unique name
    unique_names[i] <- new_name
    # Increment the tracker for this name
    name_tracker[name] <- name_tracker[name] + 1
  }
  return(unique_names)
}

convert_label_genes <- function(label_genes) {
  label_genes_show <- sapply(label_genes, function(x) {
    parts <- unlist(strsplit(x, " > "))
    part1 <- gsub("_", "", parts[1])
    if (!(grepl("neg$", parts[2]))){
      part2 <- paste0(gsub("_", "+", parts[2]), "+")
    }else{
      part2 <- paste0(gsub("_", "+", parts[2]))
    }
    
    paste(part2, part1)
  })
  label_genes_show <- gsub("^NA\\+\\s*", "", label_genes_show)
  label_genes_show <- gsub("neg", "-", label_genes_show)
  label_genes_show <- gsub("TCM", "Tcm", label_genes_show)
  label_genes_show <- gsub("TEMRA", "Temra", label_genes_show)
  label_genes_show <- gsub("TEM", "Tem", label_genes_show)
  label_genes_show <- gsub("NAIVE", "naive", label_genes_show)
  label_genes_show <- gsub("TFH", "Tfh", label_genes_show)
  label_genes_show <- gsub("MONO", "Mono", label_genes_show)
  label_genes_show <- gsub("TSCM", "Tscm", label_genes_show)
  label_genes_show <- gsub("TREG", "Treg", label_genes_show)
  label_genes_show <- gsub("BREG", "Breg", label_genes_show)
  return(label_genes_show)
}


# Define the Venn plot function
plot_venn_three_sets <- function(vec1, vec2, vec3, fig_name) {
  # Create named list
  venn_list <- list(
    AUC = vec1,
    `Odds ratio`  = vec2,
    `Fold Change`  = vec3
  )
  # Generate Venn plot object
  venn.plot <- venn.diagram(
    x = venn_list,
    filename = NULL,            # Create object, not file
    output = TRUE,
    log.filename = NULL,        # Suppresses .log file creation
    imagetype = "pdf",
    compression = "lzw",
    fill = c("#E41A1C", "#377EB8", "#4DAF4A"),
    alpha = 0.5,
    cat.names = NULL,    # No names
    cat.cex = 0,              # No category text size cat.cex = 1.5,
    cat.pos = 0,              # No label position
    cex = 1.5,
    fontfamily = "sans",
    cat.fontfamily = "sans",
    cat.dist = 0.05
  )
  # Save to PDF
  pdf(file = fig_name, width = 3.5, height = 3.5)
  grid.draw(venn.plot)
  dev.off()
}

# Function to create multiple box plots comparing cell abundance in irAE vs non-irAE patients
plot_cell_boxplots <- function(df, cell_pattern, cell_display_name, ymax = 2, ylabel_height = 1.7, ytickstep = 0.5, labeltextadd = 0.25, phenotype = "Severe irAE") {
  if (phenotype == "Severe irAE"){
    plot_df <- data.frame(
      Pre = df[[paste0(cell_pattern, "baseline")]]/100,
      On = df[[paste0(cell_pattern, "ontreat")]]/100,
      FC = df[[paste0(cell_pattern, "FC_on_pre")]],
      IRAE = factor(df$IRAEactionable01)
    ) %>% melt(id.vars = "IRAE", variable.name = "Group", value.name = "Value")
  }else if (phenotype == "Response"){
    plot_df <- data.frame(
      Pre = df[[paste0(cell_pattern, "baseline")]]/100,
      On = df[[paste0(cell_pattern, "ontreat")]]/100,
      FC = df[[paste0(cell_pattern, "FC_on_pre")]],
      IRAE = factor(df$Response01)
    ) %>% melt(id.vars = "IRAE", variable.name = "Group", value.name = "Value")
  }
  
  plot_df <- na.omit(plot_df)
  
  # Ensure the x-axis factor is properly ordered
  plot_df <- plot_df %>%
    mutate(
      Group = factor(Group, levels = c("Pre", "On", "FC")),
      Combo = interaction(Group, IRAE, sep = "."),
      Combo = factor(Combo, levels = c("Pre.0", "Pre.1", "On.0", "On.1", "FC.0", "FC.1")),
      x_pos = as.numeric(Combo)  # numeric x-axis for positioning
    )
  # Define comparisons for p-values
  comparisons <- list(c("Pre.0", "Pre.1"), c("On.0", "On.1"), c("FC.0", "FC.1"))
  
  # Function to convert p-value to symbol
  p_to_symbol <- function(p) {
    if (is.na(p)) return("ns")
    if (p < 0.001) return("***")
    else if (p < 0.01) return("**")
    else if (p < 0.05) return("*")
    else return("ns")
  }
  
  # Calculate p-values and significance symbols
  pvals <- sapply(comparisons, function(pair) {
    data1 <- plot_df$Value[plot_df$Combo == pair[1]]
    data2 <- plot_df$Value[plot_df$Combo == pair[2]]
    if (length(data1) > 0 && length(data2) > 0) {
      tryCatch(
        wilcox.test(data1, data2)$p.value,
        error = function(e) NA
      )
    } else {
      NA
    }
  })
  significance_labels <- sapply(pvals, p_to_symbol)
  
  # Now plot
  p <- ggplot(plot_df, aes(x = x_pos, y = Value, fill = IRAE, group = Combo)) + # x_pos Combo
    geom_jitter(width = 0.15, size = 0.5, alpha = 0.7) +
    geom_boxplot(outlier.shape = NA, alpha = 0.6) +
    scale_x_continuous(
      breaks = c(1.5, 3.5, 5.5),              # middle of each group (two bars)
      labels = c("Pre", "On", "FC"), # c("Pre-treatment", "On-treatment", "Fold change")
      expand = expansion(add = 0.4)
    ) +
    scale_fill_manual(values = c("0" = "#4E79A7", "1" = "#F28E2B"), name = phenotype, labels = c("No", "Yes")) +
    labs(title = cell_display_name, x = "", y = "", fill = "IRAE") + # y = "Cell abundance or fold change"
    coord_cartesian(ylim = c(0, ymax)) +
    scale_y_sqrt(breaks = seq(0, ymax, by = ytickstep)) +
    theme_classic(base_size = 12) +
    theme(plot.title = element_text(size = 9, hjust = 0.5),
          axis.text = element_text(color = "black"))
  
  # X-axis positions of the Combo levels
  comparison_pairs <- list(c(1, 2), c(3, 4), c(5, 6))
  
  # Add fixed-height significance bars with uniform tick length
  y_sig <- ylabel_height
  tick_len <- 0.04
  
  for (i in seq_along(comparison_pairs)) {
    pair <- comparison_pairs[[i]]
    p <- p +
      annotate("segment", x = pair[1], xend = pair[2], y = y_sig, yend = y_sig, size = 0.5) +  # horizontal bar
      annotate("segment", x = pair[1], xend = pair[1], y = y_sig, yend = y_sig - tick_len, size = 0.4) +  # left tick
      annotate("segment", x = pair[2], xend = pair[2], y = y_sig, yend = y_sig - tick_len, size = 0.4) +  # right tick
      annotate("text", x = mean(pair), y = y_sig + labeltextadd, label = significance_labels[i]) # , size = 4
  }
  return(p)
}

```

# load data 
```{r}
### load HCI002 data
HCI002_all_info <- read_csv(file = paste0(data_dir, "HCI002_all_info_20241008.csv"), name_repair = "minimal") 
HCI002_map_df = read.csv(file = paste0(data_dir, "OlinkID_ProteinName_map.csv"))
proteins_mapping <- setNames(HCI002_map_df$Assay, HCI002_map_df$OlinkID)  # Named vector: OlinkID -> gene symbol

```

# Prepare cor. data (CyTOF, OLINK)
```{r}

HCI002_filtered_info = HCI002_all_info[60:25931]
feature_names <- colnames(HCI002_filtered_info)
######### baseline
CyTOF_baseline_ind <- (grepl("FC_baseline", feature_names) & !grepl("_ABS$", feature_names) & !grepl("IGG4", feature_names))
OLINK_baseline_ind <- (grepl("OID", feature_names) & grepl("baseline", feature_names))

CyTOF_baseline_data <- as.matrix(HCI002_filtered_info[CyTOF_baseline_ind])
colnames(CyTOF_baseline_data) <- gsub("_FC_baseline.*", "", colnames(CyTOF_baseline_data))

OLINK_baseline_data <- as.matrix(HCI002_filtered_info[OLINK_baseline_ind])
colnames(OLINK_baseline_data) <- gsub("_OLINK_baseline", "", colnames(OLINK_baseline_data))
mapped_names <- proteins_mapping[colnames(OLINK_baseline_data)]
# Handle unmapped IDs (optional, but recommended for robustness)
mapped_names[is.na(mapped_names)] <- colnames(OLINK_baseline_data)[is.na(mapped_names)]
colnames(OLINK_baseline_data) <- mapped_names
OLINK_baseline_data <- OLINK_baseline_data[, !duplicated(colnames(OLINK_baseline_data))]


######### ontreat
CyTOF_ontreat_ind <- (grepl("FC_ontreat", feature_names) & !grepl("_ABS$", feature_names) & !grepl("IGG4", feature_names))
OLINK_ontreat_ind <- (grepl("OID", feature_names) & grepl("ontreat", feature_names))

CyTOF_ontreat_data <- as.matrix(HCI002_filtered_info[CyTOF_ontreat_ind])
colnames(CyTOF_ontreat_data) <- gsub("_FC_ontreat.*", "", colnames(CyTOF_ontreat_data))

OLINK_ontreat_data <- as.matrix(HCI002_filtered_info[OLINK_ontreat_ind])
colnames(OLINK_ontreat_data) <- gsub("_OLINK_ontreat", "", colnames(OLINK_ontreat_data))
mapped_names <- proteins_mapping[colnames(OLINK_ontreat_data)]
# Handle unmapped IDs (optional, but recommended for robustness)
mapped_names[is.na(mapped_names)] <- colnames(OLINK_ontreat_data)[is.na(mapped_names)]
colnames(OLINK_ontreat_data) <- mapped_names
OLINK_ontreat_data <- OLINK_ontreat_data[, !duplicated(colnames(OLINK_ontreat_data))]


######### fold change
CyTOF_FC_ind <- (grepl("FC_FC", feature_names) & !grepl("_ABS$", feature_names) & !grepl("IGG4", feature_names))
OLINK_FC_ind <- (grepl("OID", feature_names) & grepl("_FC_on_pre", feature_names))

CyTOF_FC_data <- as.matrix(HCI002_filtered_info[CyTOF_FC_ind])
colnames(CyTOF_FC_data) <- gsub("_FC_FC.*", "", colnames(CyTOF_FC_data))

OLINK_FC_data <- as.matrix(HCI002_filtered_info[OLINK_FC_ind])
colnames(OLINK_FC_data) <- gsub("_OLINK_FC_on_pre", "", colnames(OLINK_FC_data))
mapped_names <- proteins_mapping[colnames(OLINK_FC_data)]
# Handle unmapped IDs (optional, but recommended for robustness)
mapped_names[is.na(mapped_names)] <- colnames(OLINK_FC_data)[is.na(mapped_names)]
colnames(OLINK_FC_data) <- mapped_names
OLINK_FC_data <- OLINK_FC_data[, !duplicated(colnames(OLINK_FC_data))]

```

# Figure 1F-G/SFigure 1C-D. Make cor. heatmap within CyTOF or OLINK 
```{r}

modality <- "OLINK" # CyTOF OLINK
timePt <- "On" # Pre On FC

if (modality == "CyTOF"){
  if (timePt == "Pre"){
    data_df <- CyTOF_baseline_data
  }else if (timePt == "On"){
    data_df <- CyTOF_ontreat_data
  }else if (timePt == "FC"){
    data_df <- CyTOF_FC_data
  }
}else if (modality == "OLINK"){
  if (timePt == "Pre"){
    data_df <- OLINK_baseline_data
  }else if (timePt == "On"){
    data_df <- OLINK_ontreat_data
  }else if (timePt == "FC"){
    data_df <- OLINK_FC_data
  }
}

# Count NAs in each column
na_counts <- colSums(is.na(data_df))
# Calculate threshold (more than 50% NAs)
na_threshold <- nrow(data_df) * 0.5
# Filter columns with <= 50% NAs
data_df <- data_df[, na_counts <= na_threshold]

# Compute the correlation matrix
correlation_matrix <- cor(data_df, use = "complete.obs", method = "spearman")
# Remove rows and cols where all values are NA
correlation_matrix <- correlation_matrix[rowSums(is.na(correlation_matrix)) != ncol(correlation_matrix)-1, ]
correlation_matrix <- correlation_matrix[, colSums(is.na(correlation_matrix)) != nrow(correlation_matrix)]

correlation_matrix0 <- correlation_matrix

### Define genes to show
if (modality == "CyTOF"){
  genes_show <- c("T_CELL", "GDT", "B_CELL", "B_NAIVE", "B_MEM", "PB", "NK", "MONO", "DC", "NKT", "DPT", "DNT", "CD8_T", "CD8_TNAIVE", "CD8_TCM", "CD8_TEM", "CD8_TEMRA", "TREG", "CD4_TNAIVE", "CD4_TCM", "CD4_TEM") 
  labels_show <- c("T", "gdT", "B", "Bnaive", "Bmemory", "Plasma", "NK", "Monocytes", "DC", "NKT", "DPT", "DNT", "CD8+ T", "CD8+ Tnaive", "CD8+ Tcm", "CD8+ Tem", "CD8+ Temra", "Tregs", "CD4+ Tnaive", "CD4+ Tcm", "CD4+ Tem") 
}else if (modality == "OLINK"){
  genes_show <- c("CXCL9", "CXCL10", "CXCL11", "IFNG", "CXCL13", "IL17A", "IL10", "IL12A", "IL21", "B4GALT1", "CD276", "CD70", "FLT3LG", "FSTL3", "IL2RB", "KRT19", "PGF")
  genes_show <- genes_show[(genes_show %in% rownames(correlation_matrix))]
  labels_show <- genes_show

  set.seed(123)  # for reproducibility
  # Get all available gene names
  all_genes <- rownames(correlation_matrix)
  # Exclude the genes_show from sampling pool to avoid duplication
  remaining_genes <- setdiff(all_genes, genes_show)
  # Calculate how many more genes to sample
  n_to_sample <- 1000 - length(genes_show)
  # Sample the remaining genes
  sampled_genes <- sample(remaining_genes, n_to_sample)
  # Combine with genes_show and reorder if desired
  final_genes <- c(genes_show, sampled_genes)
  # Subset the correlation matrix
  correlation_matrix <- correlation_matrix[final_genes, final_genes]
}

# Create annotation for selected genes
row_anno <- rowAnnotation(
  genes = anno_mark(
    at = match(genes_show, rownames(correlation_matrix)),
    labels = labels_show,
    labels_gp = gpar(fontsize = 8),
    link_width = unit(5, "mm")
  )
)

# Create the heatmap
pdf(file = paste0(result_figure_dir, "Heatmap_corplot_", modality, "_", timePt, ".pdf"), height = 6, width = 8) 
p <- Heatmap(
  correlation_matrix,
  name = "Correlation",
  col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",
  show_row_names = FALSE,
  show_column_names = FALSE,
  right_annotation = row_anno,
  border = FALSE
)
print(p)
dev.off()

```

# Extract genes from the big cluster that contains most reported proteins that cor with irAEs
```{r}

# Draw the heatmap and capture clustering
ht_drawn <- draw(p)

# Get row dendrogram and cut into k clusters
row_dend <- row_dend(ht_drawn)
row_clusters <- cutree(as.hclust(row_dend), k = 10)  # adjust k to isolate the desired cluster
# Extract gene names from cluster 1 (e.g. the large upper-left cluster)
genes_in_cluster1 <- names(row_clusters[row_clusters == 1])
table(row_clusters)

known_proteins <- c("CXCL9", "CXCL10", "CXCL11", "IFNG", "CXCL13", "IL17A", "IL10", "B4GALT1", "CD276", "CD70", "FLT3LG", "FSTL3", "IL2RB", "KRT19", "PGF") # "IL12A", "IL21", # 15
sum(known_proteins %in% genes_in_cluster1)
known_proteins[!(known_proteins %in% genes_in_cluster1)]


genes_in_cluster1_pre <- genes_in_cluster1
genes_in_cluster1_on <- genes_in_cluster1

common_gene <- intersect(genes_in_cluster1_pre, genes_in_cluster1_on)

######################## plot change of cluster 1 size and known gene coverage with K
library(ComplexHeatmap)
library(ggplot2)
library(dplyr)

# Your known proteins of interest
known_proteins <- c("CXCL9", "CXCL10", "CXCL11", "IFNG", "CXCL13", "IL17A", "IL10",
                    "B4GALT1", "CD276", "CD70", "FLT3LG", "FSTL3", "IL2RB", "KRT19", "PGF")

# Get row dendrogram
row_dend <- row_dend(ht_drawn)

# Prepare results storage
results_df <- data.frame(k = integer(), total_genes = integer(), known_genes = integer())

# Loop over k values
for (k in 8:20) {
  row_clusters <- cutree(as.hclust(row_dend), k = k)
  cluster1_genes <- names(row_clusters[row_clusters == 1])
  
  total_genes <- length(cluster1_genes)
  known_in_cluster <- sum(known_proteins %in% cluster1_genes)
  
  results_df <- rbind(results_df, data.frame(
    k = k,
    total_genes = total_genes,
    known_genes = known_in_cluster
  ))
}

# Convert to long format for ggplot
plot_df <- results_df %>%
  pivot_longer(cols = c("total_genes", "known_genes"),
               names_to = "Metric", values_to = "Count")

# Plot
ggplot(plot_df, aes(x = k, y = Count, color = Metric)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c("total_genes" = "gray40", "known_genes" = "red"),
                     labels = c("Total genes in Cluster 1", "Known proteins in Cluster 1")) +
  labs(
    x = "Number of clusters (k)",
    y = "Gene count in Cluster 1",
    color = NULL,
    title = "Cluster 1 content vs. clustering resolution"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  )

```

# SFigure 5H (upper). Venn plot of overlaped genes between pre and on-treatment clustering
```{r}
# Install and load necessary package
if (!requireNamespace("VennDiagram", quietly = TRUE)) {
  install.packages("VennDiagram")
}
library(VennDiagram)

# Create a Venn diagram
venn.plot <- draw.pairwise.venn(
  area1 = length(genes_in_cluster1_pre),
  area2 = length(genes_in_cluster1_on),
  cross.area = length(intersect(genes_in_cluster1_pre, genes_in_cluster1_on)),
  category = c("Pre-treatment", "On-treatment"),
  fill = c("skyblue", "pink1"),
  lty = "blank",
  cex = 2,
  cat.cex = 2,
  cat.pos = c(-20, 20)
)

fig_width = 5
fig_height = 3
pdf_file <- paste0(result_figure_dir, "venn_cluster1_pre_on_HCI002.pdf")
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
grid.draw(venn.plot)
dev.off()

```

# SFigure 5H (lower). Test predictive power of irAE of the common 77 proteins and plot the AUC values with 95% CI. order by values
```{r}

phenotype_vec <- HCI002_all_info$IRAEactionable01
feature_names <- colnames(HCI002_all_info)

potential_predictors <- as.character(auc_results$Protein[auc_results$AUC > 0.6])

OLINK_FC_ind <- (grepl("OID", feature_names) & grepl("FC_on_pre", feature_names))
OLINK_FC_data <- as.matrix(HCI002_all_info[OLINK_FC_ind])
colnames(OLINK_FC_data) <- gsub("_OLINK_FC_on_pre", "", colnames(OLINK_FC_data))
mapped_names <- proteins_mapping[colnames(OLINK_FC_data)]
# Handle unmapped IDs (optional, but recommended for robustness)
mapped_names[is.na(mapped_names)] <- colnames(OLINK_FC_data)[is.na(mapped_names)]
colnames(OLINK_FC_data) <- mapped_names
OLINK_FC_data <- OLINK_FC_data[, !duplicated(colnames(OLINK_FC_data))]
OLINK_FC_data_filtered <- OLINK_FC_data[, common_gene]
OLINK_FC_data_filtered <- as.data.frame(scale(OLINK_FC_data_filtered))

######################## plot the AUC values with 95% CI. order by values
# Step 1: Compute AUCs and 95% CI
auc_results <- data.frame(
  Protein = character(),
  AUC = numeric(),
  CI_low = numeric(),
  CI_high = numeric(),
  stringsAsFactors = FALSE
)

for (protein in colnames(OLINK_FC_data_filtered)) {
  predictor <- OLINK_FC_data_filtered[[protein]]
  roc_obj <- roc(phenotype_vec, predictor, direction = "<", quiet = TRUE)
  ci_vals <- ci.auc(roc_obj, conf.level = 0.95)

  auc_results <- rbind(auc_results, data.frame(
    Protein = protein,
    AUC = as.numeric(auc(roc_obj)),
    CI_low = ci_vals[1],
    CI_high = ci_vals[3]
  ))
}

# Step 2: Sort by AUC ascending and factor Protein accordingly
auc_results <- auc_results %>%
  arrange(AUC) %>%
  mutate(Protein = factor(Protein, levels = Protein))  # keep sorted order

# Step 3: Set fill color based on AUC threshold
auc_results$FillColor <- ifelse(auc_results$AUC > 0.6, "#4CAF50", "white")  # green vs white

# Step 4: Plot
fig_width = 10
fig_height = 3.2
pdf_file <- paste0(result_figure_dir, "barplot_compare_AUCs_unsupervisedProteins_FC_HCI002.pdf")
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
ggplot(auc_results, aes(x = Protein, y = AUC, fill = FillColor)) +
  geom_bar(stat = "identity", color = "black") +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_fill_identity() +
  ylab("AUC (with 95% CI)") +
  xlab("") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_text()
  )
dev.off()

```

# Figure 3F. cor. heatmap between CyTOF and OLINK (major cell types + activated CD8TCM, existing biomarker proteins), use baseline CyTOF vs Fold-change OLINK proteins
```{r}

major_celltypes <- c("T_CELL", "GDT", "B_CELL", "B_NAIVE", "B_MEM", "PB", "NK", "MONO", "DC", "NKT", "DPT", "DNT", "CD8_T", "CD8_TNAIVE", "CD8_TCM", "CD8_TEM", "CD8_TEMRA", "TREG", "CD4_TNAIVE", "CD4_TCM", "CD4_TEM" , "CD8_TCM > HLADR", "CD8_TCM > CD38", "CD8_TCM > KI67", "CD4_TEM > HLADR", "CD4_TEM > CD38", "CD4_TEM > KI67") 
major_celltypes_show <- c("T", "gdT", "B", "Bnaive", "Bmemory", "Plasma", "NK", "Monocytes", "DC", "NKT", "DPT", "DNT", "CD8+ T", "CD8+ Tnaive", "CD8+ Tcm", "CD8+ Tem", "CD8+ Temra", "Tregs", "CD4+ Tnaive", "CD4+ Tcm", "CD4+ Tem", "HLADR+ CD8+ Tcm", "CD38+ CD8+ Tcm", "Ki67+ CD8+ Tcm", "HLADR+ CD4+ Tem", "CD38+ CD4+ Tem", "Ki67+ CD4+ Tem")
CyTOF_baseline_data_subset <- CyTOF_baseline_data[, major_celltypes]

OLINK_proteins_show <- c("CXCL9", "CXCL10", "CXCL11", "CXCL13", "IFNG", "IL2RB", "IL10", "IL12A", "IL17A", "IL21", "B4GALT1", "CD70", "CD276", "FLT3LG", "FSTL3",  "KRT19", "PGF")
OLINK_proteins_show <- intersect(OLINK_proteins_show, colnames(OLINK_FC_data))
OLINK_FC_data_subset <- OLINK_FC_data[, OLINK_proteins_show]

# 1. Compute correlation and p-values
cor_matrix <- matrix(NA, nrow = ncol(OLINK_FC_data_subset), ncol = ncol(CyTOF_baseline_data_subset))
p_matrix <- matrix(NA, nrow = ncol(OLINK_FC_data_subset), ncol = ncol(CyTOF_baseline_data_subset))
rownames(cor_matrix) <- colnames(OLINK_FC_data_subset)
colnames(cor_matrix) <- major_celltypes_show
rownames(p_matrix) <- rownames(cor_matrix)
colnames(p_matrix) <- colnames(cor_matrix)

# Fill correlation and p-value matrices
for (i in seq_len(ncol(OLINK_FC_data_subset))) {
  for (j in seq_len(ncol(CyTOF_baseline_data_subset))) {
    test <- suppressWarnings(cor.test(
      OLINK_FC_data_subset[, i],
      CyTOF_baseline_data_subset[, j],
      method = "spearman",
      use = "pairwise.complete.obs"
    ))
    cor_matrix[i, j] <- test$estimate
    p_matrix[i, j] <- test$p.value
  }
}

# 2. Order columns by mean correlation
col_means <- colMeans(cor_matrix, na.rm = TRUE)
ordered_colnames <- names(sort(col_means))  # ascending
cor_matrix <- cor_matrix[, ordered_colnames]
p_matrix <- p_matrix[, ordered_colnames]

# 3. Create heatmap with significance stars
p <- Heatmap(
  cor_matrix,
  name = "Correlation",
  col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_side = "left",
  column_names_side = "bottom",
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p_matrix[i, j]) && p_matrix[i, j] < 0.05) {
      grid.text("*", x = x, y = y, gp = gpar(fontsize = 10))
    }
  }
)
pdf(file = paste0(result_figure_dir, "Heatmap_corplot_CyTOF_OLINK_predictors.pdf"), height = 6*0.7, width = 8*0.8) 
print(p)
dev.off()

```

