---
title: "Correlate CyTOF and Olink features in HCI002"
author: "Tiangen Chang <tiangen.chang@sjtu.edu.cn>"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Correlate CyTOF and Olink features in HCI002}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

####### This script makes the below displays for the paper
## Figure 3A,B,D-H
## SFigure 4C,D
## SFigure 14A (left)
## SFigure 6A-E
## SFigure 7
## SFigure 8
## SFigure 9B,C
## SFigure 10H
## SFigure 14A
## SFigure 16A

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "markup", message = FALSE, warning = FALSE)
source("utils.R")
```



# Load data
```{r}

HCI002_all_info <- read_csv(
  file.path(data_dir, "HCI002_all_info.csv"),
  name_repair = "minimal",
  show_col_types = FALSE
)
HCI002_map_df <- read.csv(file.path(data_dir, "HCI002_OlinkID_ProteinName_map.csv"))
proteins_mapping <- setNames(HCI002_map_df$Assay, HCI002_map_df$OlinkID)
HCI002_data <- read.csv(file.path(data_dir, "/merged_cohorts/HCI002_info_save.csv"), stringsAsFactors = FALSE)
HCI002_all_info <- HCI002_all_info %>%
  merge(
    HCI002_data[, c("patient_therapy", setdiff(colnames(HCI002_data), colnames(HCI002_all_info)))],
    by = "patient_therapy",
    all.x = TRUE
  )


HCI001_all_info <- read_csv(
  file.path(data_dir, "HCI001_all_info.csv"),
  name_repair = "minimal",
  show_col_types = FALSE
)

HCI001_data <- read.csv(file.path(data_dir, "/merged_cohorts/HCI001_info_save.csv"), stringsAsFactors = FALSE)
HCI001_all_info <- HCI001_all_info %>%
  merge(
    HCI001_data[, c("patient_therapy", setdiff(colnames(HCI001_data), colnames(HCI001_all_info)))],
    by = "patient_therapy",
    all.x = TRUE
  )

```

# Figure 3A,B. SFigure 14A (left).
```{r}
main_feature <- "CD8T_CM_tripletORp_ratio_pre"

for (phenotype_type in phenotype_types) {
  HCI002_all_info <- set_irAE_outcome(HCI002_all_info, phenotype_type)
  HCI001_all_info <- set_irAE_outcome(HCI001_all_info, phenotype_type)

  hci_datasets <- list(HCI001 = HCI001_all_info, HCI002 = HCI002_all_info)

  for (dataset_name in names(hci_datasets)) {
    plot_data <- prepare_binary_plot_data(hci_datasets[[dataset_name]], main_feature)

    save_irAE_barplot2(
      plot_data = plot_data,
      paste0("barplot_", dataset_name, "_", phenotype_type, ".pdf"),
      width = 1.5 * 0.86,
      height = 2.7 * 0.9,
      phenotype_type = "" # phenotype_type
    )

    save_feature_boxplot(
      plot_data,
      paste0("boxplot_", dataset_name, "_", phenotype_type, ".pdf"),
      width = 1.5*0.95, height = 3.3*0.9
    )

    save_roc_plot(
      plot_data,
      paste0("ROC_", dataset_name, "_", phenotype_type, ".pdf")
    )
  }
}

```

# SFigure 8.
```{r}

data_used <- "NR" # All R NR
irAE_type <- "conventional"  # c("actionable", "conventional")
HCI002_all_info <- set_irAE_outcome(HCI002_all_info, "conventional")

if (data_used == "All"){
  data_selected <- HCI002_all_info
}else if (data_used == "R"){
  data_selected <- HCI002_all_info[HCI002_all_info$Response01 == 1, ]
}else if (data_used == "NR"){
  data_selected <- HCI002_all_info[HCI002_all_info$Response01 == 0, ]
}

if (irAE_type == "actionable"){
  plot_data <- data_selected[, c("CD8T_CM_tripletORp_ratio_pre", "IRAEactionable01")]
}else if (irAE_type == "conventional"){
  plot_data <- data_selected[, c("CD8T_CM_tripletORp_ratio_pre", "severe_irAE")]
}

colnames(plot_data) <- c("feature", "phenotype")
save_roc_plot(
  plot_data,
  paste0("ROC_HCI002_",irAE_type,"_",data_used,".pdf")
)

table(HCI002_all_info$Response01, HCI002_all_info$IRAEactionable01)

```


# SFigure 9B.
```{r}
main_feature <- "CD8T_CM_tripletORp_ratio_pre" 

age <- "young" # old young

for (phenotype_type in phenotype_types[1]) {
  temp <- set_irAE_outcome(HCI002_all_info, phenotype_type)
  if (age == "old"){
    temp <- temp[temp$Age >= 65, ]
  }else if (age == "young"){
    temp <- temp[temp$Age < 65, ]
  }
  hci_datasets <- list(HCI002 = temp)

  for (dataset_name in names(hci_datasets)) {
    plot_data <- prepare_binary_plot_data(hci_datasets[[dataset_name]], main_feature)
    plot_data <- na.omit(plot_data)
    print(paste("N = ", nrow(plot_data), ", events = ", sum(plot_data$phenotype)))
    
    save_irAE_barplot2(
      plot_data = plot_data,
      paste0("barplot_", dataset_name, "_", phenotype_type, "_", age, ".pdf"),
      width = 1.5, height = 1.85,
      phenotype_type = "" # phenotype_type
    )

    save_roc_plot(
      plot_data,
      paste0("ROC_", dataset_name, "_", phenotype_type, "_", age, ".pdf")
    )
  }
}

```


# SFigure 9C.
```{r}

main_feature <- "CD8T_CM_tripletORp_ratio_pre" # "CD8T_CM_tripletORp_ratio_pre" "CXCL9_FC"
age <- "old" # old young

clinical_feats <- c(
  "CD8T_CM_tripletORp_ratio_pre",
  "CXCL9_pre",
  "CXCL9_FC",
  "irAE_grade",
  "prior_BRAFi_MEKi",
  "prior_ICI",
  "Drug",
  "Sex",
  "Age",
  "Stage",
  "OS_time",
  "OS_event",
  "PFS_time",
  "PFS_event"
)

pooled_datasets <- list(
  HCI002 = read_cohort("HCI002_info_save.csv"),
  HCI001 = read_cohort("HCI001_info_save.csv"),
  Nicolas_FACS = read_cohort("Nicolas_all_info_FACS_save.csv", melanoma_only = TRUE),
  Nicolas_Cytek = read_cohort("Nicolas_all_info_Cytek_save.csv", melanoma_only = TRUE),
  Nicolas_CyTOF = read_cohort("Nicolas_all_info_CyTOF_save.csv", melanoma_only = TRUE),
  Lozano = read_cohort("Lozano_save.csv"),
  Gerber = read_cohort("Gerber_save.csv"),
  Switzerland1 = read_cohort("Switzerland1_save.csv"),
  Switzerland2 = read_cohort("Switzerland2_save.csv"),
  Switzerland3 = read_cohort("Switzerland3_save.csv")
)


pooled_datasets <- lapply(pooled_datasets, ensure_feats, feats = clinical_feats)
pooled_datasets <- lapply(pooled_datasets, function(df) df[, clinical_feats])
pooled_datasets <- lapply(pooled_datasets, make_severe_irAE, out_col = pheno_feat)
standardize_clinical_types <- function(df) {
  df %>%
    mutate(
      prior_BRAFi_MEKi = as.character(prior_BRAFi_MEKi),
      prior_ICI = as.character(prior_ICI),
      Drug = as.character(Drug),
      Sex = as.character(Sex),
      Stage = as.character(Stage),
      irAE_grade = suppressWarnings(as.numeric(irAE_grade)),
      severe_irAE = suppressWarnings(as.numeric(severe_irAE)),
      OS_time = suppressWarnings(as.numeric(OS_time)),
      OS_event = suppressWarnings(as.numeric(OS_event)),
      PFS_time = suppressWarnings(as.numeric(PFS_time)),
      PFS_event = suppressWarnings(as.numeric(PFS_event))
    )
}

pooled_datasets <- lapply(pooled_datasets, standardize_clinical_types)

all_merged_age <- bind_rows(
  transform(pooled_datasets$HCI002, cohort = "HCI002"),
  transform(pooled_datasets$HCI001, cohort = "HCI001"),
  transform(pooled_datasets$Nicolas_FACS, cohort = "Nicolas_FACS"),
  transform(pooled_datasets$Nicolas_Cytek, cohort = "Nicolas_Cytek"),
  transform(pooled_datasets$Switzerland1, cohort = "Switzerland1"),
  transform(pooled_datasets$Switzerland2, cohort = "Switzerland2"),
  transform(pooled_datasets$Switzerland3, cohort = "Switzerland3")
)


all_merged_age <- all_merged_age %>%
  mutate(
    Sex = recode(Sex, M = "Male", F = "Female"),
    Stage = ifelse(!is.na(Stage) & Stage != "4", "1-3", Stage),
    prior_ICI = recode(as.character(prior_ICI), `0` = "no", `1` = "yes")
  )

dataset_name <- "nonHCI002"
all_merged_age <- all_merged_age[all_merged_age$cohort != "HCI002", ]

if (age == "old"){
  temp <- all_merged_age[all_merged_age$Age >= 65, ]
}else if (age == "young"){
  temp <- all_merged_age[all_merged_age$Age < 65, ]
}

temp <- make_binary_score(temp, score_col = main_feature, high_rule = ">")

plot_data <- temp %>%
  select(group = all_of(main_feature), phenotype = severe_irAE) %>%
  drop_na() %>%
  mutate(group = factor(group, levels = c("Low", "High")))

save_irAE_barplot2(plot_data, paste0("barplot_pooled_", dataset_name, "_", age,".pdf"), width = 1.5, height = 1.85, phenotype_type = "")

```



# SFigure 16A.
```{r}

main_feature <- "CD8T_CM_tripletORp_ratio_pre"

# set phenotype
HCI002_actionable <- set_irAE_outcome(
  HCI002_all_info,
  "actionable"
)

# prepare continuous ROC data
HCI002_actionable_plot <- prepare_binary_plot_data(
  HCI002_actionable,
  feature_col = main_feature
)

# calculate Youden cutoff
HCI002_youden <- get_youden_cutoff(
  HCI002_actionable_plot
)

cutoff_youden <- HCI002_youden$threshold

print(HCI002_youden)

# ROC plot with cutoff point
save_roc_plot_with_youden(
  HCI002_actionable_plot,
  "ROC_HCI002_actionable_Youden_cutoff.pdf"
)


datasets <- list(
  HCI001 = HCI001_all_info,
  HCI002 = HCI002_all_info
)

for (phenotype_type in phenotype_types) {
  for (dataset_name in names(datasets)) {
    df <- set_irAE_outcome(
      datasets[[dataset_name]],
      phenotype_type
    )
    plot_data <- prepare_binary_plot_data(
      df,
      feature_col = main_feature,
      cutoff = cutoff_youden
    )
    cat(
      dataset_name,
      phenotype_type,
      "cutoff =",
      attr(plot_data, "cutoff"),
      "\n"
    )
    save_irAE_barplot2(
      plot_data,
      paste0(
        "barplot_",
        dataset_name,
        "_",
        phenotype_type,
        "_Youden_cutoff.pdf"
      ),
      width = 1.5 * 1.2,
      height = 2.7 * 0.95,
      phenotype_type = phenotype_type
    )
  }
}

```




# Prepare CyTOF and Olink data
```{r}
HCI002_filtered_info <- HCI002_all_info[20:25891]
feature_names <- colnames(HCI002_filtered_info)

CyTOF_baseline_data <- extract_cytof(HCI002_filtered_info, feature_names, "FC_baseline")
CyTOF_ontreat_data <- extract_cytof(HCI002_filtered_info, feature_names, "FC_ontreat")
CyTOF_FC_data <- extract_cytof(HCI002_filtered_info, feature_names, "FC_FC")

OLINK_baseline_data <- extract_olink(
  HCI002_filtered_info, feature_names,
  pattern = "baseline",
  suffix = "_OLINK_baseline",
  protein_map = proteins_mapping
)

OLINK_ontreat_data <- extract_olink(
  HCI002_filtered_info, feature_names,
  pattern = "ontreat",
  suffix = "_OLINK_ontreat",
  protein_map = proteins_mapping
)

OLINK_FC_data <- extract_olink(
  HCI002_filtered_info, feature_names,
  pattern = "_FC_on_pre",
  suffix = "_OLINK_FC_on_pre",
  protein_map = proteins_mapping
)

```

# SFigure 4C,D. 
```{r}
cytof_markers <- c(
  "T_CELL", "GDT", "B_CELL", "B_NAIVE", "B_MEM", "PB", "NK", "MONO", "DC",
  "NKT", "DPT", "DNT", "CD8_T", "CD8_TNAIVE", "CD8_TCM", "CD8_TEM",
  "CD8_TEMRA", "TREG", "CD4_TNAIVE", "CD4_TCM", "CD4_TEM"
)

cytof_labels <- c(
  "T", "gdT", "B", "Bnaive", "Bmemory", "Plasma", "NK", "Monocytes", "DC",
  "NKT", "DPT", "DNT", "CD8+ T", "CD8+ Tnaive", "CD8+ Tcm", "CD8+ Tem",
  "CD8+ Temra", "Tregs", "CD4+ Tnaive", "CD4+ Tcm", "CD4+ Tem"
)

olink_markers <- c(
  "CXCL9", "CXCL10", "CXCL11", "IFNG", "CXCL13", "IL17A", "IL10",
  "IL12A", "IL21", "B4GALT1", "CD276", "CD70", "FLT3LG", "FSTL3",
  "IL2RB", "KRT19", "PGF"
)

input_list <- list(
  CyTOF = list(Pre = CyTOF_baseline_data, On = CyTOF_ontreat_data),
  OLINK = list(Pre = OLINK_baseline_data, On = OLINK_ontreat_data)
)

genes_in_cluster1 <- list()

for (modality in names(input_list)) {
  for (time_pt in names(input_list[[modality]])) {
    data_mat <- filter_sparse_features(input_list[[modality]][[time_pt]])

    cor_mat <- cor(data_mat, use = "complete.obs", method = "spearman")
    cor_mat <- cor_mat[rowSums(is.na(cor_mat)) != ncol(cor_mat) - 1, , drop = FALSE]
    cor_mat <- cor_mat[, colSums(is.na(cor_mat)) != nrow(cor_mat), drop = FALSE]

    if (modality == "CyTOF") {
      genes_show <- cytof_markers
      labels_show <- cytof_labels
    } else {
      genes_show <- intersect(olink_markers, rownames(cor_mat))
      labels_show <- genes_show

      set.seed(123)
      sampled_genes <- sample(setdiff(rownames(cor_mat), genes_show), 1000 - length(genes_show))
      final_genes <- c(genes_show, sampled_genes)
      cor_mat <- cor_mat[final_genes, final_genes]
    }

    row_anno <- rowAnnotation(
      genes = anno_mark(
        at = match(genes_show, rownames(cor_mat)),
        labels = labels_show,
        labels_gp = gpar(fontsize = 8),
        link_width = unit(5, "mm")
      )
    )

    heatmap_obj <- Heatmap(
      cor_mat,
      name = "Correlation",
      col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
      clustering_distance_rows = "euclidean",
      clustering_distance_columns = "euclidean",
      clustering_method_rows = "complete",
      clustering_method_columns = "complete",
      show_row_names = FALSE,
      show_column_names = FALSE,
      right_annotation = row_anno,
      border = FALSE
    )

    save_pdf(
      print(heatmap_obj),
      file.path(result_figure_dir, paste0("Heatmap_HCI002_corplot_", modality, "_", time_pt, ".pdf")),
      width = 8,
      height = 6
    )

    if (modality == "OLINK") {
      ht_drawn <- draw(heatmap_obj)
      row_clusters <- cutree(as.hclust(row_dend(ht_drawn)), k = 10)
      genes_in_cluster1[[time_pt]] <- names(row_clusters[row_clusters == 1])
    }
  }
}

common_gene <- intersect(genes_in_cluster1$Pre, genes_in_cluster1$On)

```

# SFigure 10H (upper). 
```{r}
venn_plot <- draw.pairwise.venn(
  area1 = length(genes_in_cluster1$Pre),
  area2 = length(genes_in_cluster1$On),
  cross.area = length(common_gene),
  category = c("Pre-treatment", "On-treatment"),
  fill = c("skyblue", "pink1"),
  lty = "blank",
  cex = 2,
  cat.cex = 2,
  cat.pos = c(-20, 20)
)

save_pdf(
  grid.draw(venn_plot),
  file.path(result_figure_dir, "venn_HCI002_cluster1_pre_on.pdf"),
  width = 5,
  height = 3
)

```

# SFigure 10H (lower).
```{r}
phenotype_vec <- HCI002_all_info$IRAEactionable01

OLINK_FC_data_filtered <- OLINK_FC_data[, common_gene, drop = FALSE] |>
  scale() |>
  as.data.frame()

auc_results <- lapply(colnames(OLINK_FC_data_filtered), function(protein) {
  roc_obj <- roc(phenotype_vec, OLINK_FC_data_filtered[[protein]], direction = "<", quiet = TRUE)
  ci_vals <- ci.auc(roc_obj, conf.level = 0.95)

  data.frame(
    Protein = protein,
    AUC = as.numeric(auc(roc_obj)),
    CI_low = ci_vals[1],
    CI_high = ci_vals[3]
  )
}) |>
  bind_rows() |>
  arrange(AUC) |>
  mutate(
    Protein = factor(Protein, levels = Protein),
    FillColor = ifelse(AUC > 0.6, "#4CAF50", "white")
  )

auc_plot <- ggplot(auc_results, aes(x = Protein, y = AUC, fill = FillColor)) +
  geom_col(color = "black") +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_fill_identity() +
  labs(x = NULL, y = "AUC (with 95% CI)") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black"),
    axis.text.y = element_text(color = "black")
  )

save_pdf(
  print(auc_plot),
  file.path(result_figure_dir, "barplot_HCI002_compare_AUCs_unsupervisedProteins_FC.pdf"),
  width = 10,
  height = 3.2
)

```

# Figure 3D.
```{r}
major_celltypes <- c(
  "T_CELL", "GDT", "B_CELL", "B_NAIVE", "B_MEM", "PB", "NK", "MONO", "DC",
  "NKT", "DPT", "DNT", "CD8_T", "CD8_TNAIVE", "CD8_TCM", "CD8_TEM",
  "CD8_TEMRA", "TREG", "CD4_TNAIVE", "CD4_TCM", "CD4_TEM",
  "CD8_TCM > HLADR", "CD8_TCM > CD38", "CD8_TCM > KI67",
  "CD4_TEM > HLADR", "CD4_TEM > CD38", "CD4_TEM > KI67"
)

major_celltypes_show <- c(
  "T", "gdT", "B", "Bnaive", "Bmemory", "Plasma", "NK", "Monocytes", "DC",
  "NKT", "DPT", "DNT", "CD8+ T", "CD8+ Tnaive", "CD8+ Tcm", "CD8+ Tem",
  "CD8+ Temra", "Tregs", "CD4+ Tnaive", "CD4+ Tcm", "CD4+ Tem",
  "HLADR+ CD8+ Tcm", "CD38+ CD8+ Tcm", "Ki67+ CD8+ Tcm",
  "HLADR+ CD4+ Tem", "CD38+ CD4+ Tem", "Ki67+ CD4+ Tem"
)

olink_proteins_show <- c(
  "CXCL9", "CXCL10", "CXCL11", "CXCL13", "IFNG", "IL2RB", "IL10",
  "IL12A", "IL17A", "IL21", "B4GALT1", "CD70", "CD276", "FLT3LG",
  "FSTL3", "KRT19", "PGF"
)

CyTOF_baseline_subset <- CyTOF_baseline_data[, major_celltypes, drop = FALSE]
OLINK_FC_subset <- OLINK_FC_data[, intersect(olink_proteins_show, colnames(OLINK_FC_data)), drop = FALSE]

cor_results <- get_cor_and_p(
  x = OLINK_FC_subset,
  y = CyTOF_baseline_subset,
  row_names = colnames(OLINK_FC_subset),
  col_names = major_celltypes_show
)

cor_matrix <- cor_results$cor
p_matrix <- cor_results$p

ordered_cols <- names(sort(colMeans(cor_matrix, na.rm = TRUE)))
cor_matrix <- cor_matrix[, ordered_cols, drop = FALSE]
p_matrix <- p_matrix[, ordered_cols, drop = FALSE]

cytof_olink_heatmap <- Heatmap(
  cor_matrix,
  name = "Correlation",
  col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_side = "left",
  column_names_side = "bottom",
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p_matrix[i, j]) && p_matrix[i, j] < 0.05) {
      grid.text("*", x = x, y = y, gp = gpar(fontsize = 10))
    }
  }
)

save_pdf(
  print(cytof_olink_heatmap),
  file.path(result_figure_dir, "Heatmap_HCI002_corplot_CyTOF_OLINK_predictors.pdf"),
  width = 8 * 0.8,
  height = 6 * 0.7
)

```

# Figure 3E.
```{r}
cytometry_predictors <- c(
  "CD8T_CM_tripletORp_ratio_pre",
  "CD4T_EM_ratio_pre",
  "CD4T_EM_tripletORp_ratio_pre",
  "CD4TM_tripletORp_ratio_pre",
  "CD8T_EMRA_CD38p_ratio_FC",
  "CD8T_KI67p_ratio_FC",
  "CD4T_KI67p_ratio_FC",
  "Treg_KI67p_ratio_FC",
  "CD4T_Naive_ratio_pre"
)

irae_subtypes <- c(
  "IRAE_GI01",
  "IRAE_HB01",
  "IRAE_skin01",
  "IRAE_MSK01",
  "IRAEactionable01"
)

irae_labels <- c(
  IRAE_GI01 = "Gastrointestinal",
  IRAE_HB01 = "Hepatobiliary",
  IRAE_skin01 = "Skin",
  IRAE_MSK01 = "Musculoskeletal",
  IRAEactionable01 = "Overall"
)

predictor_labels <- c(
  CD8T_CM_tripletORp_ratio_pre = "Activated CD8Tcm",
  CD4T_EM_ratio_pre = "CD4Tem",
  CD4T_EM_tripletORp_ratio_pre = "Activated CD4Tem",
  CD4TM_tripletORp_ratio_pre = "Activated CD4Tm",
  CD8T_EMRA_CD38p_ratio_FC = "CD38+ CD8Temra",
  CD8T_KI67p_ratio_FC = "Ki67+ CD8T",
  CD4T_KI67p_ratio_FC = "Ki67+ CD4T",
  Treg_KI67p_ratio_FC = "Ki67+ Treg",
  CD4T_Naive_ratio_pre = "Naive CD4T"
)

compute_auc_pvalue <- function(df, outcome_col, predictor_col) {
  roc_df <- df %>%
    select(outcome = all_of(outcome_col), predictor = all_of(predictor_col)) %>%
    drop_na()

  if (nrow(roc_df) < 5 || length(unique(roc_df$outcome)) < 2) {
    return(c(AUC = NA_real_, p_value = NA_real_))
  }

  roc_obj <- pROC::roc(
    response = roc_df$outcome,
    predictor = roc_df$predictor,
    direction = "<",
    levels = c(0, 1),
    quiet = TRUE
  )

  null_roc <- pROC::roc(
    response = roc_df$outcome,
    predictor = rep(0.5, nrow(roc_df)),
    levels = c(0, 1),
    quiet = TRUE
  )

  roc_p <- suppressMessages(
    pROC::roc.test(roc_obj, null_roc, method = "delong")$p.value
  )
  c(AUC = as.numeric(roc_obj$auc), p_value = roc_p)
}

heatmap_data <- HCI002_all_info[, c(cytometry_predictors, irae_subtypes)] 

auc_matrix <- matrix(
  NA_real_,
  nrow = length(cytometry_predictors),
  ncol = length(irae_subtypes),
  dimnames = list(cytometry_predictors, irae_subtypes)
)

p_matrix <- auc_matrix

for (outcome in irae_subtypes) {
  for (predictor in cytometry_predictors) {
    if (!predictor %in% colnames(heatmap_data)) next

    result <- compute_auc_pvalue(
      df = heatmap_data,
      outcome_col = outcome,
      predictor_col = predictor
    )

    auc_matrix[predictor, outcome] <- result["AUC"]
    p_matrix[predictor, outcome] <- result["p_value"]
  }
}

colnames(auc_matrix) <- irae_labels[colnames(auc_matrix)]
rownames(auc_matrix) <- predictor_labels[rownames(auc_matrix)]

colnames(p_matrix) <- colnames(auc_matrix)
rownames(p_matrix) <- rownames(auc_matrix)

auc_heatmap <- Heatmap(
  auc_matrix,
  name = "AUC",
  col = colorRamp2(
    c(min(auc_matrix, na.rm = TRUE), 0.5, max(auc_matrix, na.rm = TRUE)),
    c("blue", "white", "red")
  ),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_side = "left",
  column_names_side = "bottom",
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p_matrix[i, j]) && p_matrix[i, j] < 0.05) {
      grid.text("*", x = x, y = y, gp = gpar(fontsize = 12))
    }
  }
)

pdf(
  file.path(result_figure_dir, "Heatmap_HCI002_AUC_pvalue_CyTOF_vs_irAE_subtypes.pdf"),
  width = 3.5,
  height = 4
)
draw(auc_heatmap)
dev.off()

```

# Figure 3F.
```{r}
group_levels <- c("nsTOX,R", "nsTOX,NR", "sTOX,R", "sTOX,NR")

predictor_df <- HCI002_all_info[, c("patient_therapy", "IRAEactionable01", "Response01", "CD8T_CM_tripletORp_ratio_pre")] %>%
  drop_na(IRAEactionable01, Response01, CD8T_CM_tripletORp_ratio_pre) %>%
  mutate(
    Group = case_when(
      IRAEactionable01 == 0 & Response01 == 1 ~ "nsTOX,R",
      IRAEactionable01 == 0 & Response01 == 0 ~ "nsTOX,NR",
      IRAEactionable01 == 1 & Response01 == 1 ~ "sTOX,R",
      IRAEactionable01 == 1 & Response01 == 0 ~ "sTOX,NR",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = group_levels),
    activated_CD8Tcm_pct = CD8T_CM_tripletORp_ratio_pre * 100
  ) %>%
  drop_na(Group)

p <- ggplot(predictor_df, aes(x = Group, y = activated_CD8Tcm_pct)) +
  geom_jitter(width = 0.15, size = 1.8, alpha = 0.6, color = "black") +
  geom_boxplot(
    fill = "#2C7BB6",
    color = "black",
    outlier.shape = NA,
    width = 0.6,
    alpha = 0.9
  ) +
  labs(
    x = NULL,
    y = "% activated CD8Tcm / CD8Tcm"
  ) +
  theme_classic() +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

pdf(
  file.path(result_figure_dir, "Boxplot_actCD8TCM_4groups_HCI002.pdf"),
  onefile = FALSE,
  width = 2,
  height = 3.5
)
print(p)
dev.off()

wilcox_comparisons <- list(
  c("nsTOX,R", "nsTOX,NR"),
  c("nsTOX,R", "sTOX,R"),
  c("nsTOX,R", "sTOX,NR"),
  c("nsTOX,NR", "sTOX,NR"),
  c("sTOX,R", "sTOX,NR")
)

wilcox_results <- lapply(wilcox_comparisons, function(groups) {
  x <- predictor_df$CD8T_CM_tripletORp_ratio_pre[predictor_df$Group == groups[1]]
  y <- predictor_df$CD8T_CM_tripletORp_ratio_pre[predictor_df$Group == groups[2]]

  data.frame(
    group_1 = groups[1],
    group_2 = groups[2],
    p_value = wilcox.test(x, y)$p.value
  )
}) %>%
  bind_rows()

wilcox_results

```

# Figure 3G/SFigure 6D,E.
```{r}
feature_test_vec <- c(
  "CD8T_CM_tripletORp_ratio_pre")

therapy_type_vec <- c("Active", "All", "Adjuvant")
phenotype_str_vec <- c("OS", "PFS")

survival_labels <- c(
  OS = "Overall \nsurvival probability",
  PFS = "Progression-free \nsurvival probability"
)

make_survival_group <- function(df, feature_col) {
  cutoff <- median(df[[feature_col]], na.rm = TRUE)

  df %>%
    mutate(
      Score = ifelse(.data[[feature_col]] > cutoff, "High", "Low"),
      Score = factor(Score, levels = c("Low", "High"))
    )
}

plot_km_curve <- function(df, phenotype, feature_col, therapy_type) {
  survival_df <- df %>%
    select(
      Score,
      time = all_of(paste0(phenotype, "_time")),
      event = all_of(paste0(phenotype, "_event"))
    ) %>%
    drop_na()

  if (nrow(survival_df) < 5 || length(unique(survival_df$Score)) < 2) {
    return(NULL)
  }

  surv_fit <- survfit(Surv(time, event) ~ Score, data = survival_df)
  cox_fit <- coxph(Surv(time, event) ~ Score, data = survival_df)
  cox_summary <- summary(cox_fit)

  hr <- cox_summary$coefficients[2]
  p_value <- cox_summary$coefficients[5]
  hr_ci <- cox_summary$conf.int[3:4]

  km_plot <- ggsurvplot(
    surv_fit,
    data = survival_df,
    size = 1,
    palette = c("#00305d", "#9c1915"),
    conf.int = FALSE,
    pval = FALSE,
    xlim = c(-3, 70),
    ylim = c(0, 1),
    xlab = "Time (months)",
    ylab = survival_labels[[phenotype]],
    break.time.by = 10,
    risk.table = TRUE,
    risk.table.height = 0.25,
    risk.table.pos = "out",
    risk.table.col = "black",
    risk.table.y.text = FALSE,
    tables.y.text = FALSE,
    tables.theme = theme_cleantable(),
    legend.labs = c("Low", "High"),
    legend.title = "",
    legend = c(0.75, 0.975),
    font.main = 12,
    font.caption = 12,
    font.legend = 12,
    font.tickslab = 12,
    font.x = 12,
    font.y = 12,
    ggtheme = theme(
      legend.background = element_rect(fill = NA, color = NA),
      legend.key = element_rect(fill = NA, color = NA),
      plot.margin = unit(c(0.2, 0.2, 0, 0.2), "cm"),
      panel.background = element_rect(fill = "white"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black")
    )
  )

  km_plot$plot <- km_plot$plot +
    annotate(
      "text",
      x = 0,
      y = 0.15,
      hjust = 0,
      size = 5,
      label = paste0(
        "HR = ", round(hr, 2),
        " (", round(hr_ci[1], 2), "-", round(hr_ci[2], 2), ")",
        "\np = ", sprintf("%.3f", p_value)
      )
    )

  pdf(
    file.path(
      result_figure_dir,
      paste0("KM_curve_", therapy_type, "_", phenotype, "_", feature_col, "_HCI002.pdf")
    ),
    width = 3.3,
    height = 3
  )
  print(km_plot, newpage = FALSE)
  dev.off()

  invisible(
    data.frame(
      therapy_type = therapy_type,
      phenotype = phenotype,
      feature = feature_col,
      HR = hr,
      HR_low = hr_ci[1],
      HR_high = hr_ci[2],
      p_value = p_value
    )
  )
}

survival_results <- list()

for (feature_col in feature_test_vec) {
  for (therapy_type in therapy_type_vec) {
    survival_data <- HCI002_all_info
    if (therapy_type != "All") {
      survival_data <- survival_data %>%
        filter(TherapyType == therapy_type)
    }
    survival_data <- survival_data %>%
      make_survival_group(feature_col)

    for (phenotype in phenotype_str_vec) {
      result <- plot_km_curve(
        df = survival_data,
        phenotype = phenotype,
        feature_col = feature_col,
        therapy_type = therapy_type
      )

      if (!is.null(result)) {
        survival_results[[length(survival_results) + 1]] <- result
      }
    }
  }
}

survival_results <- bind_rows(survival_results)
survival_results
```


# Figure 3H. 
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

control_vars <- c(
  "Age",
  "Sex",
  "Stage",
  "Drug",
  "previousICB"
  )

variable_labels <- c(
  "Activated CD8Tcm",
  "Age",
  "Sex (Female vs male)",
  "Stage (IV vs I-III)",
  "Drug (anti-PD(L)1 vs\n anti-CTLA4-containing)",
  "Previous ICB therapy"
  )

hci002_survival_df <- prepare_hci002_survival_data(feature_col)

cox_os <- fit_multivariable_cox(
  df = hci002_survival_df,
  feature_col = feature_col,
  control_vars = control_vars,
  outcome_prefix = "OS"
)

cox_pfs <- fit_multivariable_cox(
  df = hci002_survival_df,
  feature_col = feature_col,
  control_vars = control_vars,
  outcome_prefix = "PFS"
)

result_df <- tibble(
  Variable = variable_labels,
  HR_OS = cox_os$HR,
  HR_OS_low = cox_os$HR_low,
  HR_OS_up = cox_os$HR_high,
  p_val_OS = cox_os$p_value,
  HR_PFS = cox_pfs$HR,
  HR_PFS_low = cox_pfs$HR_low,
  HR_PFS_up = cox_pfs$HR_high,
  p_val_PFS = cox_pfs$p_value
)

save_forest_plot(
  plot_data = tibble(
    mean = result_df$HR_OS,
    lower = result_df$HR_OS_low,
    upper = result_df$HR_OS_up,
    Variable = result_df$Variable,
    pval = result_df$p_val_OS
  ),
  file_name = "Forest_HCI002_actCD8Tcm_clinicalVARs_MHR_OS.pdf",
  x_label = "HR of OS"
)

save_forest_plot(
  plot_data = tibble(
    mean = result_df$HR_PFS,
    lower = result_df$HR_PFS_low,
    upper = result_df$HR_PFS_up,
    Variable = result_df$Variable,
    pval = result_df$p_val_PFS
  ),
  file_name = "Forest_HCI002_actCD8Tcm_clinicalVARs_MHR_PFS.pdf",
  x_label = "HR of PFS"
)

```


# AUC summary across irAE definitions in HCI001 and HCI002
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

irae_types <- c(
  "IRAEgrade01",
  "IRAEactionable01",
  "IRAE_GI01",
  "IRAE_HB01",
  "IRAE_skin01",
  "IRAE_MSK01"
)

compute_irae_auc_table <- function(df, feature_col, irae_types) {
  bind_rows(lapply(irae_types, function(irae_type) {
    compute_auc_summary(df, outcome_col = irae_type, predictor_col = feature_col) %>%
      mutate(irAE_type = irae_type, .before = 1)
  })) %>%
    rename(
      irAE_AUC = AUC,
      irAE_AUC_lower = AUC_low,
      irAE_AUC_upper = AUC_high,
      irAE_pval = p_value
    )
}

AUC_irAEs_df_HCI002 <- compute_irae_auc_table(HCI002_all_info, feature_col, irae_types)
AUC_irAEs_df_HCI001 <- compute_irae_auc_table(HCI001_all_info, feature_col, irae_types)

```


# SFigure 6A. 
```{r}
irae_label_map <- c(
  IRAE_GI01 = "Gastrointestinal",
  IRAE_HB01 = "Hepatobiliary",
  IRAE_skin01 = "Skin",
  IRAE_MSK01 = "Musculoskeletal"
)

auc_dataset_list <- list(
  HCI001 = AUC_irAEs_df_HCI001,
  HCI002 = AUC_irAEs_df_HCI002
)

for (dataset_name in names(auc_dataset_list)) {
  plot_df <- auc_dataset_list[[dataset_name]] %>%
    filter(irAE_type %in% names(irae_label_map)) %>%
    mutate(label = irae_label_map[irAE_type]) %>%
    rename(
      AUC = irAE_AUC,
      AUC_low = irAE_AUC_lower,
      AUC_high = irAE_AUC_upper,
      p_value = irAE_pval
    )

  save_auc_barplot(
    plot_df = plot_df,
    file_name = paste0("barplot_AUC_compare_irAE_types_", dataset_name, ".pdf")
  )
}

```

# SFigure 6C.
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"
feature_label <- "% activated CD8Tcm"
dataset_name <- "HCI002"

plot_df <- HCI002_all_info %>%
  transmute(
    Predictor = .data[[feature_col]],
    Age = `Age at C1`,
    TMB = `TMB (mutation /mb)`,
    PDL1_TPS = `PD-L1 TPS`,
    PDL1_CPS = `PD-L1 CPS`,
    Gender = Gender,
    BRAF = case_when(
      is.na(`BRAF Mutation`) ~ "Unknown",
      grepl("V600E|V600K|Wildtype", `BRAF Mutation`) ~ `BRAF Mutation`,
      TRUE ~ "Other mut."
    )
  )

continuous_vars <- c(
  Age = "Age",
  TMB = "TMB (mut. / Mb)",
  PDL1_TPS = "PDL1 TPS (%)",
  PDL1_CPS = "PDL1 CPS (%)"
)

for (var_name in names(continuous_vars)) {
  p <- ggplot(plot_df, aes(x = .data[[var_name]], y = Predictor)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
    labs(x = continuous_vars[[var_name]], y = feature_label) +
    theme_classic() +
    theme(axis.text = element_text(color = "black"))

  ggsave(
    file.path(result_figure_dir, paste0("scatter_", dataset_name, "_", feature_col, "_vs_", var_name, ".pdf")),
    plot = p,
    width = 2.1,
    height = 2.4
  )
}

categorical_vars <- c("Gender", "BRAF")

for (var_name in categorical_vars) {
  p <- ggplot(plot_df, aes(x = .data[[var_name]], y = Predictor)) +
    geom_boxplot(outlier.shape = NA, fill = "lightgray", color = "black") +
    geom_jitter(width = 0.2, size = 1, aes(color = .data[[var_name]])) +
    labs(x = var_name, y = feature_label) +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )

  ggsave(
    file.path(result_figure_dir, paste0("barplot_", dataset_name, "_", feature_col, "_vs_", var_name, ".pdf")),
    plot = p,
    width = 1.5,
    height = 3
  )
}

wilcox_results <- bind_rows(lapply(categorical_vars, function(var_name) {
  test_df <- plot_df %>%
    mutate(Group = .data[[var_name]]) %>%
    mutate(Group = ifelse(var_name == "BRAF" & !is.na(Group), ifelse(Group == "Wildtype", "Wildtype", "Mutant"), Group)) %>%
    select(Predictor, Group) %>%
    drop_na()

  p_value <- if (n_distinct(test_df$Group) == 2) {
    wilcox.test(Predictor ~ Group, data = test_df)$p.value
  } else {
    NA_real_
  }

  tibble(Variable = var_name, P_value = p_value)
}))

wilcox_results

```

# SFigure 6B.
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

subgroup_df <- HCI002_all_info %>%
  transmute(
    Feature = .data[[feature_col]],
    Phenotype = IRAEactionable01,
    Drug = Drug,
    ICB_exposure = ifelse(`ICI Naïve?` == "Yes", "ICB-naive", "ICB-experienced"),
    Stage = ifelse(grepl("IV", `Baseline Stage / TNM`), "Stage IV", "Stage I-III"),
    TherapyType = case_when(
      TherapyType == "Active" ~ "Active treatment",
      TherapyType == "Adjuvant" ~ "Adjuvant",
      TherapyType == "Neoadjuvant" ~ "Neoadjuvant",
      TRUE ~ TherapyType
    )
  )

subgroup_specs <- list(
  Drug = c("PD1/PDL1", "CTLA4-containing"),
  ICB_exposure = c("ICB-naive", "ICB-experienced"),
  Stage = c("Stage I-III", "Stage IV"),
  TherapyType = c("Active treatment", "Adjuvant")
)

auc_subgroup_results <- bind_rows(lapply(names(subgroup_specs), function(group_col) {
  bind_rows(lapply(subgroup_specs[[group_col]], function(group_level) {
    subgroup_df %>%
      filter(.data[[group_col]] == group_level) %>%
      compute_auc_summary(outcome_col = "Phenotype", predictor_col = "Feature") %>%
      mutate(
        subgroup = group_col,
        label = group_level,
        .before = 1
      )
  }))
}))

for (group_col in names(subgroup_specs)) {
  plot_df <- auc_subgroup_results %>%
    filter(subgroup == group_col) %>%
    mutate(label = paste0(label, "\n(n=", n, ")"))

  save_auc_barplot(
    plot_df = plot_df,
    file_name = paste0("barplot_AUC_compare_subgroups_", group_col, "_HCI002.pdf")
  )
}

```

# SFigure 7A.
```{r}
library(ComplexHeatmap)
library(circlize)

marker_patient_level_expr_df <- read.csv(paste0(data_dir, "HCI002_marker_median_expr_patient_level_CD8Tcm.csv"), row.names = 1)

marker_cell_level_expr_df <- read.csv(
    paste0(data_dir, "HCI002_marker_median_expr_cell_level.csv"),
    row.names = 1
)
marker_cell_level_expr_df$LAG3 <- NULL
marker_cell_level_expr_df$CD57 <- NULL
marker_cell_level_expr_df$CTLA4 <- NULL
marker_cell_level_expr_df$CD25 <- NULL
marker_cell_level_expr_df$TIM3 <- NULL

fc <- as.numeric(marker_cell_level_expr_df["activated_CD8T_CM", ] /
      marker_cell_level_expr_df["inactivated_CD8T_CM", ])

marker_cell_level_expr_df <- marker_cell_level_expr_df[
    ,
    order(fc, decreasing = TRUE),
    drop = FALSE
]

fc[order(fc, decreasing = TRUE)]

z_expr <- scale(marker_cell_level_expr_df)

z_expr <- z_expr[
    ,
    apply(z_expr, 2, function(x) all(is.finite(x))),
    drop = FALSE
]

z_expr[z_expr > 2]  <- 2
z_expr[z_expr < -2] <- -2

rownames(z_expr) <- c(
    "Activated CD8 TCM",
    "Inactivated CD8 TCM",
    "Overall CD8 T",
    "Overall T",
    "All cells"
)

col_fun <- colorRamp2(
    c(-2, 0, 2),
    c("#2166AC", "white", "#B2182B")
)

ht <- Heatmap(
    z_expr,
    name = "Z-score",
    col = col_fun,

    cluster_rows = FALSE,
    cluster_columns = FALSE,

    show_row_dend = FALSE,
    show_column_dend = FALSE,

    row_names_side = "left",
    row_names_gp = grid::gpar(fontsize = 10),

    column_names_rot = 45,
    column_names_gp = grid::gpar(fontsize = 9),

    rect_gp = grid::gpar(
        col = "white",
        lwd = 1
    ),

    heatmap_legend_param = list(
        title = "Z-score",
        at = c(-2, -1, 0, 1, 2),
        direction = "vertical"
    ),

    border = TRUE
)

pdf(
    paste0(
        result_figure_dir,
        "activated_CD8T_CM_state_heatmap.pdf"
    ),
    width = 9*0.5,
    height = 4.2*0.4,
    useDingbats = FALSE
)

draw(
    ht,
    heatmap_legend_side = "right"
)

dev.off()


##### patient level paired 
markers_to_plot <- c("CD39", "ICOS", "PD1", "TIGIT")

font_size <- 10

for (marker in markers_to_plot) {

    act_col   <- paste0(marker, "_activated")
    inact_col <- paste0(marker, "_inactivated")

    df <- marker_patient_level_expr_df[
        complete.cases(
            marker_patient_level_expr_df[, c(act_col, inact_col)]
        ),
        c("Patient_ID", act_col, inact_col)
    ]

    plot_df <- rbind(
        data.frame(
            Patient_ID = df$Patient_ID,
            Group = "Inactivated",
            Expression = df[[inact_col]]
        ),
        data.frame(
            Patient_ID = df$Patient_ID,
            Group = "Activated",
            Expression = df[[act_col]]
        )
    )

    plot_df$Group <- factor(
        plot_df$Group,
        levels = c("Inactivated", "Activated")
    )

    pval <- wilcox.test(
        df[[inact_col]],
        df[[act_col]],
        paired = TRUE,
        exact = FALSE
    )$p.value

    p_label <- paste0(
        "P = ",
        format(
            pval,
            scientific = TRUE,
            digits = 2
        )
    )

    p_label <- gsub("e-0", "e-", p_label)
    p_label <- gsub("e\\+0", "e+", p_label)

    y_max <- max(plot_df$Expression, na.rm = TRUE)
    y_min <- min(plot_df$Expression, na.rm = TRUE)
    y_range <- y_max - y_min

    p_y <- y_max + 0.08 * y_range

    p <- ggplot(
        plot_df,
        aes(
            x = Group,
            y = Expression,
            group = Patient_ID
        )
    ) +
        geom_line(
            color = "grey75",
            linewidth = 0.4,
            alpha = 0.65
        ) +
        geom_boxplot(
            aes(group = Group),
            width = 0.42,
            outlier.shape = NA,
            fill = "white",
            color = "black",
            linewidth = 0.7
        ) +
        geom_point(
            size = 1.7,
            color = "#377EB8",
            alpha = 0.45,
            position = position_jitter(width = 0.06, height = 0)
        ) +
        annotate(
            "text",
            x = 1.5,
            y = p_y,
            label = p_label,
            size = font_size / ggplot2::.pt
        ) +
        labs(
            x = NULL,
            y = paste0(marker, " expression"),
            title = marker
        ) +
        scale_y_continuous(
            expand = expansion(
                mult = c(0.03, 0.15)
            )
        ) +
        theme_classic(base_size = font_size) +
        theme(
            # Same font size everywhere
            plot.title = element_text(
                size = font_size,
                face = "plain",
                hjust = 0.5
            ),
            axis.title = element_text(
                size = font_size,
                face = "plain"
            ),
            axis.text = element_text(
                size = font_size,
                color = "black"
            ),
            axis.text.x = element_text(
                angle = 45,
                hjust = 1
            ),
            axis.line = element_line(
                linewidth = 0.6
            ),
            axis.ticks = element_line(
                linewidth = 0.5
            )
        )

    ggsave(
        filename = paste0(
            result_figure_dir,
            marker,
            "_activated_vs_inactivated_CD8TCM.pdf"
        ),
        plot = p,
        width = 3.2*0.5,
        height = 4.0*0.8,
        units = "in",
        useDingbats = FALSE
    )
}

```

# SFigure 7B.
```{r}

HCI002_all_info <- read_csv(
  file.path(data_dir, "HCI002_all_info.csv"),
  name_repair = "minimal",
  show_col_types = FALSE
)
HCI002_map_df <- read.csv(file.path(data_dir, "HCI002_OlinkID_ProteinName_map.csv"))
proteins_mapping <- setNames(HCI002_map_df$Assay, HCI002_map_df$OlinkID)
HCI002_data <- read.csv(file.path(data_dir, "/merged_cohorts/HCI002_info_save.csv"), stringsAsFactors = FALSE)
HCI002_all_info <- HCI002_all_info %>%
  merge(
    HCI002_data[, c("patient_therapy", setdiff(colnames(HCI002_data), colnames(HCI002_all_info)))],
    by = "patient_therapy",
    all.x = TRUE
  )

old_names <- colnames(HCI002_all_info)
oid <- sub("^(OID[0-9]+).*", "\\1", old_names)
matched <- oid %in% names(proteins_mapping)
new_names <- old_names
new_names[matched] <- paste0(
    proteins_mapping[oid[matched]],
    sub("^OID[0-9]+_OLINK", "", old_names[matched])
)
colnames(HCI002_all_info) <- new_names

x <- HCI002_all_info$CD8T_CM_tripletORp_ratio_pre
OLINK_proteins <- as.character(proteins_mapping)

cor_results <- do.call(
    rbind,
    lapply(OLINK_proteins, function(protein) {

        var_test <- paste0(protein, "_FC_on_pre")

        if (!var_test %in% colnames(HCI002_all_info)) {
            return(
                data.frame(
                    protein = protein,
                    SCC = NA_real_,
                    pval = NA_real_,
                    n = 0
                )
            )
        }

        y <- HCI002_all_info[[var_test]]
        keep <- complete.cases(x, y)

        if (sum(keep) < 3) {
            return(
                data.frame(
                    protein = protein,
                    SCC = NA_real_,
                    pval = NA_real_,
                    n = sum(keep)
                )
            )
        }

        test <- suppressWarnings(
            cor.test(
                x[keep],
                y[keep],
                method = "spearman",
                exact = FALSE
            )
        )

        data.frame(
            protein = protein,
            SCC = unname(test$estimate),
            pval = test$p.value,
            n = sum(keep)
        )
    })
)

rownames(cor_results) <- NULL

cor_results <- cor_results[
    order(-cor_results$SCC),
]

cor_results <- na.omit(cor_results)

head(cor_results, 50)

top_hits <- c(
  "CXCL9", "CXCL10", "CXCL11", "CXCL13",
  "IL17A", "XCL1", "CCL3", "IL18",
  "CD70", "WARS", "TIGIT", "GBP1"
)

plot_df <- cor_results %>%
  dplyr::mutate(
    neglog10P = -log10(pmax(pval, .Machine$double.xmin)),
    significance = ifelse(pval < 0.05, "P < 0.05", "P ≥ 0.05")
  )

# Inspect selected proteins
plot_df %>%
  dplyr::filter(.data$protein %in% top_hits) %>%
  dplyr::arrange(.data$pval) %>%
  dplyr::select(protein, SCC, pval, n)

p <- ggplot(
  plot_df,
  aes(x = SCC, y = neglog10P)
) +

  # All proteins
  geom_point(
    aes(color = significance),
    size = 1.8,
    alpha = 0.65
  ) +

  # P = 0.05 threshold
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    linewidth = 0.6
  ) +

  # SCC = 0 reference
  geom_vline(
    xintercept = 0,
    linetype = "dotted",
    linewidth = 0.4
  ) +

  # Highlight selected biologically meaningful proteins
  geom_point(
    data = plot_df %>%
      dplyr::filter(.data$protein %in% top_hits),
    size = 2.7,
    shape = 21,
    stroke = 0.8
  ) +

  # Labels
  geom_text_repel(
    data = plot_df %>%
      dplyr::filter(.data$protein %in% top_hits),
    aes(label = protein),
    size = 4,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0,
    segment.size = 0.35,
    max.overlaps = Inf,
    seed = 123
  ) +

  scale_color_manual(
    values = c(
      "P ≥ 0.05" = "grey75",
      "P < 0.05" = "#C43C39"
    )
  ) +

  labs(
    x = "Spearman correlation coefficient",
    y = expression(-log[10](P)),
    color = NULL
  ) +

  theme_classic(base_size = 14) +

  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 14),
    legend.text = element_text(size = 14),
    legend.position = "top",
    legend.justification = "left",
    axis.line = element_line(linewidth = 0.6),
    axis.ticks = element_line(linewidth = 0.6),
    plot.margin = margin(8, 15, 8, 8)
  )

ggsave(
  paste0(result_figure_dir, "aCD8Tcm_olink_SCC_volcano.pdf"),
  plot = p,
  width = 6.5*0.8,
  height = 5.5*0.8,
  device = "pdf"
)




## ranked-proteome GSEA
gsea_df <- cor_results %>%
  dplyr::filter(
    !is.na(SCC),
    is.finite(SCC),
    !is.na(protein),
    protein != ""
  ) %>%
  dplyr::arrange(desc(abs(SCC))) %>%
  dplyr::distinct(protein, .keep_all = TRUE)

# Positive SCC -> positive NES
# Negative SCC -> negative NES
ranks <- gsea_df$SCC
names(ranks) <- gsea_df$protein
ranks <- sort(ranks, decreasing = TRUE)

stopifnot(
  all(is.finite(ranks)),
  !anyDuplicated(names(ranks)),
  all(diff(ranks) <= 0)
)


hallmark_df <- msigdbr(
  species = "Homo sapiens",
  collection = "H"
)

hallmark_pathways <- split(
  hallmark_df$gene_symbol,
  hallmark_df$gs_name
)

set.seed(123)

fgsea_hallmark <- fgseaMultilevel(
  pathways = hallmark_pathways,
  stats = ranks,
  minSize = 10,
  maxSize = 500,
  scoreType = "std",
  eps = 0
)

fgsea_hallmark <- as.data.frame(fgsea_hallmark) %>%
  dplyr::arrange(pval)

clean_hallmark_name <- function(x) {

  x %>%
    stringr::str_remove("^HALLMARK_") %>%
    stringr::str_replace_all("_", " ") %>%
    stringr::str_to_lower() %>%
    stringr::str_to_title() %>%

    # Improve common biological abbreviations
    stringr::str_replace_all("Ifn ", "IFN-") %>%
    stringr::str_replace_all("Il2 Stat5", "IL2/STAT5") %>%
    stringr::str_replace_all("Il6 Jak Stat3", "IL6/JAK/STAT3") %>%
    stringr::str_replace_all("Tnfa Signaling Via Nfkb",
                             "TNFA signaling via NF-kB") %>%
    stringr::str_replace_all("Pi3k Akt Mtor",
                             "PI3K/AKT/mTOR") %>%
    stringr::str_replace_all("Mtorc1", "mTORC1") %>%
    stringr::str_replace_all("G2m", "G2M") %>%
    stringr::str_replace_all("E2f", "E2F") %>%
    stringr::str_replace_all("Myc", "MYC") %>%
    stringr::str_replace_all("P53", "p53") %>%
    stringr::str_replace_all("Kras", "KRAS") %>%
    stringr::str_replace_all("Tgf Beta", "TGF-beta") %>%
    stringr::str_replace_all("Wnt Beta Catenin",
                             "WNT-beta-catenin")
}

fgsea_hallmark <- fgsea_hallmark %>%
  dplyr::mutate(
    Description = clean_hallmark_name(pathway)
  )


plot_df <- fgsea_hallmark %>%
  dplyr::filter(padj < 0.05) %>%
  dplyr::mutate(
    neglog10_padj = -log10(padj)
  ) %>%
  dplyr::arrange(neglog10_padj) %>%
  dplyr::mutate(
    Description = factor(
      Description,
      levels = Description
    )
  )

p_gsea <- ggplot(
  plot_df,
  aes(
    x = neglog10_padj,
    y = Description,
    fill = NES
  )
) +

  geom_col(
    width = 0.72
  ) +

  # FDR = 0.05 reference
  geom_vline(
    xintercept = -log10(0.05),
    linetype = "dashed",
    linewidth = 0.5,
    color = "grey40"
  ) +

  # Diverging NES color scale
  scale_fill_gradient2(
    low = "#3C6FA8",
    mid = "white",
    high = "#C94F4F",
    midpoint = 0,
    name = "NES"
  ) +

  labs(
    x = expression(-log[10]("p.adjust")),
    y = NULL
  ) +

  theme_classic(base_size = 10) +

  theme(
    panel.grid = element_blank(),

    axis.line.y = element_blank(),
    axis.line.x = element_line(
      color = "black",
      linewidth = 0.5
    ),

    axis.ticks.y = element_blank(),

    axis.text = element_text(
      color = "black",
      size = 9
    ),

    axis.text.y = element_text(
      size = 9,
      lineheight = 0.9
    ),

    axis.title.x = element_text(
      size = 10,
      margin = margin(t = 6)
    ),

    legend.position = "right",

    legend.title = element_text(
      size = 9
    ),

    legend.text = element_text(
      size = 8
    ),

    plot.margin = margin(
      t = 5,
      r = 5,
      b = 5,
      l = 5
    )
  )

ggsave(
  filename = file.path(
    result_figure_dir,
    "aCD8Tcm_olink_Hallmark_GSEA.pdf"
  ),
  plot = p_gsea,
  width = 5.2 * 0.9,
  height = max(3.0, 0.25 * nrow(plot_df) + 1.2) * 0.9,
  device = "pdf"
)

```


