---
title: "Correlate CyTOF and Olink features in HCI002"
author: "Tiangen Chang"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Correlate CyTOF and Olink features in HCI002}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

####### This script makes the below displays for the paper
## Figure 3a,b,d-h
## SFigure 1c,d
## SFigure 5a,b
## SFigure 4a-g
## SFigure 6h

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "markup", message = FALSE, warning = FALSE)
```

# Package Loading
```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(pROC)
library(VennDiagram)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(pROC)

```

# Parameters and shared functions
```{r}
result_figure_dir <- "./Figures/"
data_dir <- "./SourceData/"

dir.create(result_figure_dir, showWarnings = FALSE, recursive = TRUE)

map_olink_names <- function(x, protein_map) {
  mapped <- protein_map[colnames(x)]
  mapped[is.na(mapped)] <- colnames(x)[is.na(mapped)]
  colnames(x) <- mapped
  x[, !duplicated(colnames(x)), drop = FALSE]
}

extract_cytof <- function(df, feature_names, pattern) {
  keep <- grepl(pattern, feature_names) &
    !grepl("_ABS$", feature_names) &
    !grepl("IGG4", feature_names)

  x <- as.matrix(df[keep])
  colnames(x) <- gsub(paste0("_", pattern, ".*"), "", colnames(x))
  x
}

extract_olink <- function(df, feature_names, pattern, suffix, protein_map) {
  keep <- grepl("OID", feature_names) & grepl(pattern, feature_names)
  x <- as.matrix(df[keep])
  colnames(x) <- gsub(suffix, "", colnames(x))
  map_olink_names(x, protein_map)
}

filter_sparse_features <- function(x, max_na_fraction = 0.5) {
  x[, colMeans(is.na(x)) <= max_na_fraction, drop = FALSE]
}

save_pdf <- function(plot_expr, file, width, height) {
  pdf(file, width = width, height = height, onefile = FALSE)
  on.exit(dev.off(), add = TRUE)
  force(plot_expr)
}

get_cor_and_p <- function(x, y, row_names, col_names) {
  cor_matrix <- matrix(NA_real_, nrow = ncol(x), ncol = ncol(y),
                       dimnames = list(row_names, col_names))
  p_matrix <- cor_matrix

  for (i in seq_len(ncol(x))) {
    for (j in seq_len(ncol(y))) {
      test <- suppressWarnings(cor.test(
        x[, i], y[, j],
        method = "spearman",
        use = "pairwise.complete.obs"
      ))
      cor_matrix[i, j] <- unname(test$estimate)
      p_matrix[i, j] <- test$p.value
    }
  }

  list(cor = cor_matrix, p = p_matrix)
}

make_severe_irAE <- function(df, grade_col = "irAE_grade", out_col = "severe_irAE") {
  if (!grade_col %in% colnames(df)) stop(sprintf("Column '%s' not found.", grade_col))
  grade <- suppressWarnings(as.numeric(df[[grade_col]]))
  df[[out_col]] <- ifelse(is.na(grade), NA, ifelse(grade >= 3, 1, 0))
  df
}

set_irAE_outcome <- function(df, phenotype_type) {
  if (phenotype_type == "actionable") {
    df$severe_irAE <- df$IRAEactionable01
  } else if (phenotype_type == "conventional") {
    df <- make_severe_irAE(df, out_col = "severe_irAE")
  } else {
    stop("phenotype_type must be 'actionable' or 'conventional'.")
  }
  df
}


prepare_binary_plot_data <- function(df, feature_col, outcome_col = "severe_irAE") {
  df %>%
    select(feature = all_of(feature_col), phenotype = all_of(outcome_col)) %>%
    drop_na() %>%
    mutate(
      group = factor(ifelse(feature <= median(feature, na.rm = TRUE), "Low", "High"), levels = c("Low", "High")),
      phenotype_label = factor(phenotype, levels = c(0, 1), labels = c("No severe irAE", "Severe irAE"))
    )
}

calc_group_rate_or <- function(plot_data) {
  rates <- aggregate(
    phenotype ~ group,
    data = plot_data,
    FUN = function(x) mean(as.numeric(x) == 1) * 100
  )

  odds_high <- rates$phenotype[2] / (100 - rates$phenotype[2])
  odds_low <- rates$phenotype[1] / (100 - rates$phenotype[1])
  list(rates = rates, or = round(odds_high / odds_low, 1))
}

save_irAE_barplot <- function(plot_data, file_name, width = 1.5 * 0.95, height = 2.7 * 0.95) {
  stat <- calc_group_rate_or(plot_data)

  p <- ggplot(stat$rates, aes(x = group, y = phenotype, fill = group)) +
    geom_bar(stat = "identity", color = "black", width = 0.6) +
    labs(x = NULL, y = "Severe irAE (%)") +
    ylim(0, 100) +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      legend.position = "none"
    ) +
    geom_segment(aes(x = 1, xend = 2, y = 90, yend = 90), linewidth = 0.4) +
    geom_segment(aes(x = 1, xend = 1, y = 90, yend = 86), linewidth = 0.4) +
    geom_segment(aes(x = 2, xend = 2, y = 90, yend = 86), linewidth = 0.4) +
    annotate("text", x = 1.5, y = 97, label = paste0("OR = ", stat$or), size = 3)

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = width, height = height)
  print(p)
  dev.off()

  invisible(stat$or)
}

save_feature_boxplot <- function(plot_data, file_name) {
  p_val <- wilcox.test(feature ~ phenotype_label, data = plot_data)$p.value

  p <- ggplot(plot_data, aes(x = phenotype_label, y = feature * 100, fill = phenotype_label)) +
    geom_jitter(aes(color = phenotype_label), width = 0.2, size = 2) +
    geom_boxplot(outlier.shape = NA, alpha = 0.5, color = "black") +
    scale_color_manual(values = c("blue", "brown")) +
    scale_fill_manual(values = c("blue", "brown")) +
    labs(x = NULL, y = "% activated CD8Tcm / CD8Tcm") +
    ylim(0, 90) +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    ) +
    geom_signif(
      comparisons = list(c("No severe irAE", "Severe irAE")),
      annotations = paste0("p = ", signif(p_val, 2)),
      textsize = 4,
      y_position = 80,
      vjust = -0.5,
      tip_length = 0.05
    )

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = 1.5, height = 3.3)
  print(p)
  dev.off()
}

save_roc_plot <- function(plot_data, file_name) {
  roc_obj <- pROC::roc(
    response = plot_data$phenotype,
    predictor = plot_data$feature,
    levels = c(0, 1),
    direction = "<",
    quiet = TRUE
  )

  auc_ci <- pROC::ci.auc(roc_obj)
  label <- sprintf("AUC = %.2f (%.2f,%.2f)", auc_ci[2], auc_ci[1], auc_ci[3])

  roc_df <- data.frame(
    fpr = 1 - roc_obj$specificities,
    sensitivity = roc_obj$sensitivities
  ) %>%
    arrange(fpr, sensitivity) %>%
    distinct()

  p <- ggplot(roc_df, aes(x = fpr, y = sensitivity)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray70") +
    geom_step(direction = "vh", color = "black", linewidth = 1) +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    annotate("text", x = 0.17, y = 0.1, hjust = 0, label = label, size = 4) +
    theme_classic() +
    theme(axis.text = element_text(color = "black"))

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = 2.7, height = 2.2)
  print(p)
  dev.off()
}

```


# Load data
```{r}
HCI002_all_info <- read_csv(
  file.path(data_dir, "HCI002_all_info.csv"),
  name_repair = "minimal",
  show_col_types = FALSE
)
HCI002_map_df <- read.csv(file.path(data_dir, "HCI002_OlinkID_ProteinName_map.csv"))
proteins_mapping <- setNames(HCI002_map_df$Assay, HCI002_map_df$OlinkID)
HCI002_data <- read.csv(file.path(data_dir, "/merged_cohorts/HCI002_info_save.csv"), stringsAsFactors = FALSE)
HCI002_all_info <- HCI002_all_info %>%
  merge(
    HCI002_data[, c("patient_therapy", setdiff(colnames(HCI002_data), colnames(HCI002_all_info)))],
    by = "patient_therapy",
    all.x = TRUE
  )


HCI001_all_info <- read_csv(
  file.path(data_dir, "HCI001_all_info.csv"),
  name_repair = "minimal",
  show_col_types = FALSE
)
HCI001_all_info$`RedCap ID0` <- sub("_.*", "", HCI001_all_info$`RedCap ID`)
HCI001_all_info$`Current Therapy` <- HCI001_all_info$`Current Therapy` %>%
  str_replace_all("\\u00A0", "") %>%   # remove all non-breaking spaces
  str_trim(side = "both")              # trim any remaining whitespace
HCI001_all_info$patient_therapy = paste(HCI001_all_info$`RedCap ID0`, HCI001_all_info$`Current Therapy`, HCI001_all_info$`Type of therapy`, sep = "_")
HCI001_all_info <- HCI001_all_info[!duplicated(HCI001_all_info$patient_therapy), ]
patient_therapy_final <- read_excel(paste0(data_dir, "TableS2.xlsx"))
patient_therapy_final <- patient_therapy_final$patient_therapy
HCI001_all_info <- HCI001_all_info[HCI001_all_info$patient_therapy %in% patient_therapy_final, ]
HCI001_data <- read.csv(file.path(data_dir, "/merged_cohorts/HCI001_info_save.csv"), stringsAsFactors = FALSE)
HCI001_all_info <- HCI001_all_info %>%
  merge(
    HCI001_data[, c("patient_therapy", setdiff(colnames(HCI001_data), colnames(HCI001_all_info)))],
    by = "patient_therapy",
    all.x = TRUE
  )

```


# Figure 3a-b (phenotype_type = "actionable") and SFigure 5a-b (phenotype_type = "conventional"). ROC plots, box plots, bar plots.
```{r}
phenotype_types <- c("actionable", "conventional")
main_feature <- "CD8T_CM_tripletORp_ratio_pre"

for (phenotype_type in phenotype_types) {
  HCI002_all_info <- set_irAE_outcome(HCI002_all_info, phenotype_type)
  HCI001_all_info <- set_irAE_outcome(HCI001_all_info, phenotype_type)

  hci_datasets <- list(HCI001 = HCI001_all_info, HCI002 = HCI002_all_info)

  for (dataset_name in names(hci_datasets)) {
    plot_data <- prepare_binary_plot_data(hci_datasets[[dataset_name]], main_feature)

    save_irAE_barplot(
      plot_data,
      paste0("barplot_", dataset_name, "_", phenotype_type, ".pdf")
    )

    save_feature_boxplot(
      plot_data,
      paste0("boxplot_", dataset_name, "_", phenotype_type, ".pdf")
    )

    save_roc_plot(
      plot_data,
      paste0("ROC_", dataset_name, "_", phenotype_type, ".pdf")
    )
  }
}
```


# Prepare CyTOF and Olink data
```{r}
HCI002_filtered_info <- HCI002_all_info[61:25932]
feature_names <- colnames(HCI002_filtered_info)

CyTOF_baseline_data <- extract_cytof(HCI002_filtered_info, feature_names, "FC_baseline")
CyTOF_ontreat_data <- extract_cytof(HCI002_filtered_info, feature_names, "FC_ontreat")
CyTOF_FC_data <- extract_cytof(HCI002_filtered_info, feature_names, "FC_FC")

OLINK_baseline_data <- extract_olink(
  HCI002_filtered_info, feature_names,
  pattern = "baseline",
  suffix = "_OLINK_baseline",
  protein_map = proteins_mapping
)

OLINK_ontreat_data <- extract_olink(
  HCI002_filtered_info, feature_names,
  pattern = "ontreat",
  suffix = "_OLINK_ontreat",
  protein_map = proteins_mapping
)

OLINK_FC_data <- extract_olink(
  HCI002_filtered_info, feature_names,
  pattern = "_FC_on_pre",
  suffix = "_OLINK_FC_on_pre",
  protein_map = proteins_mapping
)

```

# SFigure 1c,d. Within-modality correlation heatmaps.
```{r}
cytof_markers <- c(
  "T_CELL", "GDT", "B_CELL", "B_NAIVE", "B_MEM", "PB", "NK", "MONO", "DC",
  "NKT", "DPT", "DNT", "CD8_T", "CD8_TNAIVE", "CD8_TCM", "CD8_TEM",
  "CD8_TEMRA", "TREG", "CD4_TNAIVE", "CD4_TCM", "CD4_TEM"
)

cytof_labels <- c(
  "T", "gdT", "B", "Bnaive", "Bmemory", "Plasma", "NK", "Monocytes", "DC",
  "NKT", "DPT", "DNT", "CD8+ T", "CD8+ Tnaive", "CD8+ Tcm", "CD8+ Tem",
  "CD8+ Temra", "Tregs", "CD4+ Tnaive", "CD4+ Tcm", "CD4+ Tem"
)

olink_markers <- c(
  "CXCL9", "CXCL10", "CXCL11", "IFNG", "CXCL13", "IL17A", "IL10",
  "IL12A", "IL21", "B4GALT1", "CD276", "CD70", "FLT3LG", "FSTL3",
  "IL2RB", "KRT19", "PGF"
)

input_list <- list(
  CyTOF = list(Pre = CyTOF_baseline_data, On = CyTOF_ontreat_data),
  OLINK = list(Pre = OLINK_baseline_data, On = OLINK_ontreat_data)
)

genes_in_cluster1 <- list()

for (modality in names(input_list)) {
  for (time_pt in names(input_list[[modality]])) {
    data_mat <- filter_sparse_features(input_list[[modality]][[time_pt]])

    cor_mat <- cor(data_mat, use = "complete.obs", method = "spearman")
    cor_mat <- cor_mat[rowSums(is.na(cor_mat)) != ncol(cor_mat) - 1, , drop = FALSE]
    cor_mat <- cor_mat[, colSums(is.na(cor_mat)) != nrow(cor_mat), drop = FALSE]

    if (modality == "CyTOF") {
      genes_show <- cytof_markers
      labels_show <- cytof_labels
    } else {
      genes_show <- intersect(olink_markers, rownames(cor_mat))
      labels_show <- genes_show

      set.seed(123)
      sampled_genes <- sample(setdiff(rownames(cor_mat), genes_show), 1000 - length(genes_show))
      final_genes <- c(genes_show, sampled_genes)
      cor_mat <- cor_mat[final_genes, final_genes]
    }

    row_anno <- rowAnnotation(
      genes = anno_mark(
        at = match(genes_show, rownames(cor_mat)),
        labels = labels_show,
        labels_gp = gpar(fontsize = 8),
        link_width = unit(5, "mm")
      )
    )

    heatmap_obj <- Heatmap(
      cor_mat,
      name = "Correlation",
      col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
      clustering_distance_rows = "euclidean",
      clustering_distance_columns = "euclidean",
      clustering_method_rows = "complete",
      clustering_method_columns = "complete",
      show_row_names = FALSE,
      show_column_names = FALSE,
      right_annotation = row_anno,
      border = FALSE
    )

    save_pdf(
      print(heatmap_obj),
      file.path(result_figure_dir, paste0("Heatmap_HCI002_corplot_", modality, "_", time_pt, ".pdf")),
      width = 8,
      height = 6
    )

    if (modality == "OLINK") {
      ht_drawn <- draw(heatmap_obj)
      row_clusters <- cutree(as.hclust(row_dend(ht_drawn)), k = 10)
      genes_in_cluster1[[time_pt]] <- names(row_clusters[row_clusters == 1])
    }
  }
}

common_gene <- intersect(genes_in_cluster1$Pre, genes_in_cluster1$On)

```

# SFigure 6h (upper). Venn plot.
```{r}
venn_plot <- draw.pairwise.venn(
  area1 = length(genes_in_cluster1$Pre),
  area2 = length(genes_in_cluster1$On),
  cross.area = length(common_gene),
  category = c("Pre-treatment", "On-treatment"),
  fill = c("skyblue", "pink1"),
  lty = "blank",
  cex = 2,
  cat.cex = 2,
  cat.pos = c(-20, 20)
)

save_pdf(
  grid.draw(venn_plot),
  file.path(result_figure_dir, "venn_HCI002_cluster1_pre_on.pdf"),
  width = 5,
  height = 3
)

```

# SFigure 6h (lower). AUCs for common proteins.
```{r}
phenotype_vec <- HCI002_all_info$IRAEactionable01

OLINK_FC_data_filtered <- OLINK_FC_data[, common_gene, drop = FALSE] |>
  scale() |>
  as.data.frame()

auc_results <- lapply(colnames(OLINK_FC_data_filtered), function(protein) {
  roc_obj <- roc(phenotype_vec, OLINK_FC_data_filtered[[protein]], direction = "<", quiet = TRUE)
  ci_vals <- ci.auc(roc_obj, conf.level = 0.95)

  data.frame(
    Protein = protein,
    AUC = as.numeric(auc(roc_obj)),
    CI_low = ci_vals[1],
    CI_high = ci_vals[3]
  )
}) |>
  bind_rows() |>
  arrange(AUC) |>
  mutate(
    Protein = factor(Protein, levels = Protein),
    FillColor = ifelse(AUC > 0.6, "#4CAF50", "white")
  )

auc_plot <- ggplot(auc_results, aes(x = Protein, y = AUC, fill = FillColor)) +
  geom_col(color = "black") +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_fill_identity() +
  labs(x = NULL, y = "AUC (with 95% CI)") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black"),
    axis.text.y = element_text(color = "black")
  )

save_pdf(
  print(auc_plot),
  file.path(result_figure_dir, "barplot_HCI002_compare_AUCs_unsupervisedProteins_FC.pdf"),
  width = 10,
  height = 3.2
)

```

# Figure 3d. CyTOF–Olink correlation heatmap.
```{r}
major_celltypes <- c(
  "T_CELL", "GDT", "B_CELL", "B_NAIVE", "B_MEM", "PB", "NK", "MONO", "DC",
  "NKT", "DPT", "DNT", "CD8_T", "CD8_TNAIVE", "CD8_TCM", "CD8_TEM",
  "CD8_TEMRA", "TREG", "CD4_TNAIVE", "CD4_TCM", "CD4_TEM",
  "CD8_TCM > HLADR", "CD8_TCM > CD38", "CD8_TCM > KI67",
  "CD4_TEM > HLADR", "CD4_TEM > CD38", "CD4_TEM > KI67"
)

major_celltypes_show <- c(
  "T", "gdT", "B", "Bnaive", "Bmemory", "Plasma", "NK", "Monocytes", "DC",
  "NKT", "DPT", "DNT", "CD8+ T", "CD8+ Tnaive", "CD8+ Tcm", "CD8+ Tem",
  "CD8+ Temra", "Tregs", "CD4+ Tnaive", "CD4+ Tcm", "CD4+ Tem",
  "HLADR+ CD8+ Tcm", "CD38+ CD8+ Tcm", "Ki67+ CD8+ Tcm",
  "HLADR+ CD4+ Tem", "CD38+ CD4+ Tem", "Ki67+ CD4+ Tem"
)

olink_proteins_show <- c(
  "CXCL9", "CXCL10", "CXCL11", "CXCL13", "IFNG", "IL2RB", "IL10",
  "IL12A", "IL17A", "IL21", "B4GALT1", "CD70", "CD276", "FLT3LG",
  "FSTL3", "KRT19", "PGF"
)

CyTOF_baseline_subset <- CyTOF_baseline_data[, major_celltypes, drop = FALSE]
OLINK_FC_subset <- OLINK_FC_data[, intersect(olink_proteins_show, colnames(OLINK_FC_data)), drop = FALSE]

cor_results <- get_cor_and_p(
  x = OLINK_FC_subset,
  y = CyTOF_baseline_subset,
  row_names = colnames(OLINK_FC_subset),
  col_names = major_celltypes_show
)

cor_matrix <- cor_results$cor
p_matrix <- cor_results$p

ordered_cols <- names(sort(colMeans(cor_matrix, na.rm = TRUE)))
cor_matrix <- cor_matrix[, ordered_cols, drop = FALSE]
p_matrix <- p_matrix[, ordered_cols, drop = FALSE]

cytof_olink_heatmap <- Heatmap(
  cor_matrix,
  name = "Correlation",
  col = colorRamp2(c(-1, 0, 1), c("blue", "white", "red")),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_side = "left",
  column_names_side = "bottom",
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p_matrix[i, j]) && p_matrix[i, j] < 0.05) {
      grid.text("*", x = x, y = y, gp = gpar(fontsize = 10))
    }
  }
)

save_pdf(
  print(cytof_olink_heatmap),
  file.path(result_figure_dir, "Heatmap_HCI002_corplot_CyTOF_OLINK_predictors.pdf"),
  width = 8 * 0.8,
  height = 6 * 0.7
)

```



# Figure 3e. AUC heatmap of baseline cytometric predictors across irAE subtypes
```{r}
cytometry_predictors <- c(
  "CD8T_CM_tripletORp_ratio_pre",
  "CD4T_EM_ratio_pre",
  "CD4T_EM_tripletORp_ratio_pre",
  "CD4TM_tripletORp_ratio_pre",
  "CD8T_EMRA_CD38p_ratio_FC",
  "CD8T_KI67p_ratio_FC",
  "CD4T_KI67p_ratio_FC",
  "Treg_KI67p_ratio_FC",
  "CD4T_Naive_ratio_pre"
)

irae_subtypes <- c(
  "IRAE_GI01",
  "IRAE_HB01",
  "IRAE_skin01",
  "IRAE_MSK01",
  "IRAEactionable01"
)

irae_labels <- c(
  IRAE_GI01 = "Gastrointestinal",
  IRAE_HB01 = "Hepatobiliary",
  IRAE_skin01 = "Skin",
  IRAE_MSK01 = "Musculoskeletal",
  IRAEactionable01 = "Overall"
)

predictor_labels <- c(
  CD8T_CM_tripletORp_ratio_pre = "Activated CD8Tcm",
  CD4T_EM_ratio_pre = "CD4Tem",
  CD4T_EM_tripletORp_ratio_pre = "Activated CD4Tem",
  CD4TM_tripletORp_ratio_pre = "Activated CD4Tm",
  CD8T_EMRA_CD38p_ratio_FC = "CD38+ CD8Temra",
  CD8T_KI67p_ratio_FC = "Ki67+ CD8T",
  CD4T_KI67p_ratio_FC = "Ki67+ CD4T",
  Treg_KI67p_ratio_FC = "Ki67+ Treg",
  CD4T_Naive_ratio_pre = "Naive CD4T"
)

compute_auc_pvalue <- function(df, outcome_col, predictor_col) {
  roc_df <- df %>%
    select(outcome = all_of(outcome_col), predictor = all_of(predictor_col)) %>%
    drop_na()

  if (nrow(roc_df) < 5 || length(unique(roc_df$outcome)) < 2) {
    return(c(AUC = NA_real_, p_value = NA_real_))
  }

  roc_obj <- pROC::roc(
    response = roc_df$outcome,
    predictor = roc_df$predictor,
    direction = "<",
    levels = c(0, 1),
    quiet = TRUE
  )

  null_roc <- pROC::roc(
    response = roc_df$outcome,
    predictor = rep(0.5, nrow(roc_df)),
    levels = c(0, 1),
    quiet = TRUE
  )

  roc_p <- suppressMessages(
    pROC::roc.test(roc_obj, null_roc, method = "delong")$p.value
  )

  c(AUC = as.numeric(roc_obj$auc), p_value = roc_p)
}

heatmap_data <- HCI002_all_info[, c(cytometry_predictors, irae_subtypes)] 

auc_matrix <- matrix(
  NA_real_,
  nrow = length(cytometry_predictors),
  ncol = length(irae_subtypes),
  dimnames = list(cytometry_predictors, irae_subtypes)
)

p_matrix <- auc_matrix

for (outcome in irae_subtypes) {
  for (predictor in cytometry_predictors) {
    if (!predictor %in% colnames(heatmap_data)) next

    result <- compute_auc_pvalue(
      df = heatmap_data,
      outcome_col = outcome,
      predictor_col = predictor
    )

    auc_matrix[predictor, outcome] <- result["AUC"]
    p_matrix[predictor, outcome] <- result["p_value"]
  }
}

colnames(auc_matrix) <- irae_labels[colnames(auc_matrix)]
rownames(auc_matrix) <- predictor_labels[rownames(auc_matrix)]

colnames(p_matrix) <- colnames(auc_matrix)
rownames(p_matrix) <- rownames(auc_matrix)

auc_heatmap <- Heatmap(
  auc_matrix,
  name = "AUC",
  col = colorRamp2(
    c(min(auc_matrix, na.rm = TRUE), 0.5, max(auc_matrix, na.rm = TRUE)),
    c("blue", "white", "red")
  ),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_side = "left",
  column_names_side = "bottom",
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p_matrix[i, j]) && p_matrix[i, j] < 0.05) {
      grid.text("*", x = x, y = y, gp = gpar(fontsize = 12))
    }
  }
)

pdf(
  file.path(result_figure_dir, "Heatmap_HCI002_AUC_pvalue_CyTOF_vs_irAE_subtypes.pdf"),
  width = 3.5,
  height = 4
)
draw(auc_heatmap)
dev.off()

```

# Figure 3f. Activated CD8Tcm abundance across irAE and ICB response groups in HCI002
```{r}
group_levels <- c("nsTOX,R", "nsTOX,NR", "sTOX,R", "sTOX,NR")

predictor_df <- HCI002_all_info[, c("patient_therapy", "IRAEactionable01", "Response01", "CD8T_CM_tripletORp_ratio_pre")] %>%
  drop_na(IRAEactionable01, Response01, CD8T_CM_tripletORp_ratio_pre) %>%
  mutate(
    Group = case_when(
      IRAEactionable01 == 0 & Response01 == 1 ~ "nsTOX,R",
      IRAEactionable01 == 0 & Response01 == 0 ~ "nsTOX,NR",
      IRAEactionable01 == 1 & Response01 == 1 ~ "sTOX,R",
      IRAEactionable01 == 1 & Response01 == 0 ~ "sTOX,NR",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = group_levels),
    activated_CD8Tcm_pct = CD8T_CM_tripletORp_ratio_pre * 100
  ) %>%
  drop_na(Group)

p <- ggplot(predictor_df, aes(x = Group, y = activated_CD8Tcm_pct)) +
  geom_jitter(width = 0.15, size = 1.8, alpha = 0.6, color = "black") +
  geom_boxplot(
    fill = "#2C7BB6",
    color = "black",
    outlier.shape = NA,
    width = 0.6,
    alpha = 0.9
  ) +
  labs(
    x = NULL,
    y = "% activated CD8Tcm / CD8Tcm"
  ) +
  theme_classic() +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

pdf(
  file.path(result_figure_dir, "Boxplot_actCD8TCM_4groups_HCI002.pdf"),
  onefile = FALSE,
  width = 2,
  height = 3.5
)
print(p)
dev.off()

wilcox_comparisons <- list(
  c("nsTOX,R", "nsTOX,NR"),
  c("nsTOX,R", "sTOX,R"),
  c("nsTOX,R", "sTOX,NR"),
  c("nsTOX,NR", "sTOX,NR"),
  c("sTOX,R", "sTOX,NR")
)

wilcox_results <- lapply(wilcox_comparisons, function(groups) {
  x <- predictor_df$CD8T_CM_tripletORp_ratio_pre[predictor_df$Group == groups[1]]
  y <- predictor_df$CD8T_CM_tripletORp_ratio_pre[predictor_df$Group == groups[2]]

  data.frame(
    group_1 = groups[1],
    group_2 = groups[2],
    p_value = wilcox.test(x, y)$p.value
  )
}) %>%
  bind_rows()

wilcox_results

```

# Figure 3g/SFigure 4d,e,g. HCI002 survival analysis by cytometric predictors
```{r}
feature_test_vec <- c(
  "CD8T_CM_tripletORp_ratio_pre",
  "CD4_TEM_FC_baselineL2"
)

therapy_type_vec <- c("Active", "All", "Adjuvant")
phenotype_str_vec <- c("OS", "PFS")

survival_labels <- c(
  OS = "Overall \nsurvival probability",
  PFS = "Progression-free \nsurvival probability"
)

survival_info_df <- readxl::read_excel(file.path(data_dir, "HCI002_clinical_info.xlsx")) %>%
  select(
    `Vial ID`,
    `Record ID`,
    `Systemic Therapy Name`,
    `Systemic Therapy Type`,
    OS_time,
    OS_event,
    PFS_time,
    PFS_event
  ) %>%
  mutate(
    uniqueID = paste0(
      `Record ID`, "_",
      `Systemic Therapy Name`, "_",
      `Systemic Therapy Type`, "_",
      `Vial ID`
    ),
    OS_time = OS_time / 30,
    PFS_time = PFS_time / 30
  )

prepare_survival_data <- function(feature_col, therapy_type) {
  predictor_df <- HCI002_all_info

  if (therapy_type != "All") {
    predictor_df <- predictor_df %>%
      filter(TherapyType == therapy_type)
  }

  predictor_df %>%
    select(`Vial ID`, patient_therapy, all_of(feature_col)) %>%
    drop_na(all_of(feature_col)) %>%
    mutate(uniqueID = paste0(patient_therapy, "_", `Vial ID`)) %>%
    distinct(uniqueID, .keep_all = TRUE) %>%
    merge(survival_info_df, by = "uniqueID", all.x = TRUE)
}

make_survival_group <- function(df, feature_col) {
  cutoff <- median(df[[feature_col]], na.rm = TRUE)

  df %>%
    mutate(
      Score = ifelse(.data[[feature_col]] > cutoff, "High", "Low"),
      Score = factor(Score, levels = c("Low", "High"))
    )
}

plot_km_curve <- function(df, phenotype, feature_col, therapy_type) {
  survival_df <- df %>%
    select(
      Score,
      time = all_of(paste0(phenotype, "_time")),
      event = all_of(paste0(phenotype, "_event"))
    ) %>%
    drop_na()

  if (nrow(survival_df) < 5 || length(unique(survival_df$Score)) < 2) {
    return(NULL)
  }

  surv_fit <- survfit(Surv(time, event) ~ Score, data = survival_df)
  cox_fit <- coxph(Surv(time, event) ~ Score, data = survival_df)
  cox_summary <- summary(cox_fit)

  hr <- cox_summary$coefficients[2]
  p_value <- cox_summary$coefficients[5]
  hr_ci <- cox_summary$conf.int[3:4]

  km_plot <- ggsurvplot(
    surv_fit,
    data = survival_df,
    size = 1,
    palette = c("#00305d", "#9c1915"),
    conf.int = FALSE,
    pval = FALSE,
    xlim = c(-3, 70),
    ylim = c(0, 1),
    xlab = "Time (months)",
    ylab = survival_labels[[phenotype]],
    break.time.by = 10,
    risk.table = TRUE,
    risk.table.height = 0.25,
    risk.table.pos = "out",
    risk.table.col = "black",
    risk.table.y.text = FALSE,
    tables.y.text = FALSE,
    tables.theme = theme_cleantable(),
    legend.labs = c("Low", "High"),
    legend.title = "",
    legend = c(0.75, 0.975),
    font.main = 12,
    font.caption = 12,
    font.legend = 12,
    font.tickslab = 12,
    font.x = 12,
    font.y = 12,
    ggtheme = theme(
      legend.background = element_rect(fill = NA, color = NA),
      legend.key = element_rect(fill = NA, color = NA),
      plot.margin = unit(c(0.2, 0.2, 0, 0.2), "cm"),
      panel.background = element_rect(fill = "white"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black")
    )
  )

  km_plot$plot <- km_plot$plot +
    annotate(
      "text",
      x = 0,
      y = 0.15,
      hjust = 0,
      size = 5,
      label = paste0(
        "HR = ", round(hr, 2),
        " (", round(hr_ci[1], 2), "-", round(hr_ci[2], 2), ")",
        "\np = ", sprintf("%.3f", p_value)
      )
    )

  pdf(
    file.path(
      result_figure_dir,
      paste0("KM_curve_", therapy_type, "_", phenotype, "_", feature_col, "_HCI002.pdf")
    ),
    width = 3.3,
    height = 3
  )
  print(km_plot, newpage = FALSE)
  dev.off()

  invisible(
    data.frame(
      therapy_type = therapy_type,
      phenotype = phenotype,
      feature = feature_col,
      HR = hr,
      HR_low = hr_ci[1],
      HR_high = hr_ci[2],
      p_value = p_value
    )
  )
}

survival_results <- list()

for (feature_col in feature_test_vec) {
  for (therapy_type in therapy_type_vec) {
    survival_data <- prepare_survival_data(feature_col, therapy_type) %>%
      make_survival_group(feature_col)

    for (phenotype in phenotype_str_vec) {
      result <- plot_km_curve(
        df = survival_data,
        phenotype = phenotype,
        feature_col = feature_col,
        therapy_type = therapy_type
      )

      if (!is.null(result)) {
        survival_results[[length(survival_results) + 1]] <- result
      }
    }
  }
}

survival_results <- bind_rows(survival_results)
survival_results
```



# Helper functions
```{r}
format_p_value <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p >= 0.05 ~ as.character(round(p, 2)),
    p >= 0.0095 ~ as.character(round(p, 3)),
    TRUE ~ format(p, scientific = TRUE, digits = 2)
  )
}

compute_auc_summary <- function(df, outcome_col, predictor_col) {
  auc_df <- df %>%
    select(outcome = all_of(outcome_col), predictor = all_of(predictor_col)) %>%
    drop_na()

  if (nrow(auc_df) < 5 || length(unique(auc_df$outcome)) < 2) {
    return(tibble(AUC = NA_real_, AUC_low = NA_real_, AUC_high = NA_real_, p_value = NA_real_, n = nrow(auc_df)))
  }

  roc_obj <- pROC::roc(
    response = auc_df$outcome,
    predictor = auc_df$predictor,
    direction = "<",
    levels = c(0, 1),
    quiet = TRUE
  )

  auc_ci <- pROC::ci.auc(roc_obj)

  tibble(
    AUC = as.numeric(auc_ci[2]),
    AUC_low = as.numeric(auc_ci[1]),
    AUC_high = as.numeric(auc_ci[3]),
    p_value = verification::roc.area(auc_df$outcome, auc_df$predictor)$p.value,
    n = nrow(auc_df)
  )
}

save_auc_barplot <- function(plot_df, file_name, width = 2, height = 3.5) {
  p <- ggplot(plot_df, aes(x = label, y = AUC, fill = label)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = AUC_low, ymax = AUC_high), width = 0.2, color = "black", linewidth = 0.5) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey", linewidth = 0.5) +
    labs(x = NULL, y = "AUC") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      axis.text.x = element_text(angle = 47, hjust = 1),
      legend.position = "none"
    )

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = width, height = height)
  print(p)
  dev.off()
}

prepare_hci002_survival_data <- function(feature_col) {
  predictor_df <- HCI002_all_info %>%
    select(
      `Vial ID`,
      patient_therapy,
      all_of(feature_col),
      Gender,
      `Age at C1`,
      `ICI Naïve?`,
      `Systemic Therapy Type`,
      `Baseline Stage / TNM`,
      `Duration of treatment`,
      IRAEactionable01,
      `Systemic Therapy Name`
    ) %>%
    mutate(
      Sex = factor(ifelse(Gender == "Male", 1, 0), levels = c(0, 1)),
      Age = `Age at C1`,
      previousICB = factor(ifelse(`ICI Naïve?` == "Yes", 0, 1), levels = c(0, 1)),
      activeICB = factor(ifelse(`Systemic Therapy Type` == "active treatment", 1, 0), levels = c(0, 1)),
      Stage = factor(ifelse(grepl("Stage IV", `Baseline Stage / TNM`), 1, 0), levels = c(0, 1)),
      ICBduration = `Duration of treatment`,
      Drug = ifelse(
        `Systemic Therapy Name` %in% c(
          "Nivolumab/ Ipilimumab",
          "PD1-CTLA4 bispecific",
          "Ipi 10 / temozolomide",
          "Nivolumab/ Ipilimumab flipped dose",
          "Ipi 3 / temozolomide",
          "Pembrolizumab / CTLA-4-LAG3 bispecific",
          "CTLA-4-LAG3 bispecific",
          "Pembrolizumab/ anti-CTLA4",
          "anti-CTLA4"
        ),
        "CTLA4-containing",
        "PD1/PDL1"
      ),
      uniqueID = paste0(patient_therapy, "_", `Vial ID`)
    ) %>%
    select(
      uniqueID,
      all_of(feature_col),
      Sex,
      Age,
      previousICB,
      activeICB,
      Stage,
      ICBduration,
      IRAEactionable01,
      Drug
    ) %>%
    drop_na() %>%
    distinct(uniqueID, .keep_all = TRUE)

  survival_info_df <- readxl::read_excel(file.path(data_dir, "HCI002_clinical_info.xlsx")) %>%
    select(
      `Vial ID`,
      `Record ID`,
      `Systemic Therapy Name`,
      `Systemic Therapy Type`,
      OS_time,
      OS_event,
      PFS_time,
      PFS_event
    ) %>%
    mutate(
      uniqueID = paste0(`Record ID`, "_", `Systemic Therapy Name`, "_", `Systemic Therapy Type`, "_", `Vial ID`),
      OS_time = as.numeric(OS_time) / 30,
      PFS_time = as.numeric(PFS_time) / 30,
      OS_event = as.numeric(OS_event),
      PFS_event = as.numeric(PFS_event)
    )

  merge(predictor_df, survival_info_df, by = "uniqueID", all.x = TRUE) %>%
    filter(!is.na(.data[[feature_col]]))
}

fit_multivariable_cox <- function(df, feature_col, control_vars, outcome_prefix, cutoff = 0.04) {
  model_df <- df %>%
    mutate(
      Score = factor(ifelse(.data[[feature_col]] > cutoff, "High", "Low"), levels = c("Low", "High"))
    ) %>%
    select(
      Score,
      all_of(control_vars),
      time = all_of(paste0(outcome_prefix, "_time")),
      event = all_of(paste0(outcome_prefix, "_event"))
    ) %>%
    drop_na()

  model <- coxph(
    as.formula(paste("Surv(time, event) ~ Score +", paste(control_vars, collapse = " + "))),
    data = model_df
  )

  cox_sum <- summary(model)

  tibble(
    HR = cox_sum$coefficients[, 2],
    HR_low = cox_sum$conf.int[, 3],
    HR_high = cox_sum$conf.int[, 4],
    p_value = cox_sum$coefficients[, 5]
  )
}

save_forest_plot <- function(plot_data, file_name, x_label) {
  plot_data <- plot_data %>%
    mutate(
      effSize = sprintf("%.3f", mean),
      pval = format_p_value(pval)
    )

  fp <- forestplot(
    plot_data,
    labeltext = c(effSize, Variable, pval),
    mean = mean,
    lower = lower,
    upper = upper,
    graph.pos = 3,
    boxsize = 0.25,
    vertices = TRUE,
    clip = c(0, 4),
    xlog = FALSE,
    zero = 1,
    txt_gp = fpTxtGp(
      ticks = gpar(cex = 1.1),
      xlab = gpar(cex = 1.1),
      label = gpar(cex = 1.1),
      title = gpar(cex = 1.1)
    ),
    xlab = x_label,
    xticks = c(0, 1, 2, 3),
    graphwidth = unit(2.4, "cm"),
    lineheight = unit(0.85, "cm")
  ) %>%
    fp_set_style(
      box = "black",
      line = "black",
      align = "llr",
      summary = "black"
    ) %>%
    fp_add_header(
      effSize = "HR",
      Variable = "Variable",
      pval = "P-value"
    )

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = 5.5, height = 3.4)
  print(fp)
  dev.off()
}

save_km_plot <- function(df, feature_col, phenotype, cohort_name, xlim_max = 50) {
  y_labels <- c(
    OS = "Overall \nsurvival probability",
    PFS = "Progression-free \nsurvival probability"
  )

  cutoff <- median(df[[feature_col]], na.rm = TRUE)

  survival_df <- df %>%
    mutate(
      Score = factor(ifelse(.data[[feature_col]] > cutoff, "High", "Low"), levels = c("Low", "High"))
    ) %>%
    select(
      Score,
      time = all_of(paste0(phenotype, "_time")),
      event = all_of(paste0(phenotype, "_event"))
    ) %>%
    drop_na()

  surv_fit <- survfit(Surv(time, event) ~ Score, data = survival_df)
  cox_fit <- coxph(Surv(time, event) ~ Score, data = survival_df)
  cox_sum <- summary(cox_fit)

  hr <- cox_sum$coefficients[2]
  p_value <- cox_sum$coefficients[5]
  hr_ci <- cox_sum$conf.int[3:4]

  km_plot <- ggsurvplot(
    surv_fit,
    data = survival_df,
    size = 1,
    palette = c("#00305d", "#9c1915"),
    conf.int = FALSE,
    pval = FALSE,
    xlim = c(-3, xlim_max),
    ylim = c(0, 1),
    xlab = "Time (months)",
    ylab = y_labels[[phenotype]],
    break.time.by = 10,
    risk.table = TRUE,
    risk.table.height = 0.25,
    risk.table.pos = "out",
    risk.table.col = "black",
    risk.table.y.text = FALSE,
    tables.y.text = FALSE,
    tables.theme = theme_cleantable(),
    legend.labs = c("Low", "High"),
    legend.title = "",
    legend = c(0.75, 0.975),
    font.main = 12,
    font.caption = 12,
    font.legend = 12,
    font.tickslab = 12,
    font.x = 12,
    font.y = 12,
    ggtheme = theme(
      legend.background = element_rect(fill = NA, color = NA),
      legend.key = element_rect(fill = NA, color = NA),
      plot.margin = unit(c(0.2, 0.2, 0, 0.2), "cm"),
      panel.background = element_rect(fill = "white"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text = element_text(color = "black")
    )
  )

  km_plot$plot <- km_plot$plot +
    annotate(
      "text",
      x = 0,
      y = 0.15,
      hjust = 0,
      size = 5,
      label = paste0(
        "HR = ", round(hr, 2),
        " (", round(hr_ci[1], 2), "-", round(hr_ci[2], 2), ")",
        "\np = ", sprintf("%.3f", p_value)
      )
    )

  pdf(file.path(result_figure_dir, paste0("KM_curve_", phenotype, "_", feature_col, "_", cohort_name, ".pdf")), width = 3.3, height = 3)
  print(km_plot, newpage = FALSE)
  dev.off()
}
```


# Figure 3h. Multivariable Cox regression forest plots in HCI002
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

control_vars <- c(
  "Age",
  "Sex",
  "Stage",
  "Drug",
  "previousICB",
  "IRAEactionable01"
)

variable_labels <- c(
  "Activated CD8Tcm",
  "Age",
  "Sex (Female vs male)",
  "Stage (IV vs I-III)",
  "Drug (anti-PD(L)1 vs\n anti-CTLA4-containing)",
  "Previous ICB therapy",
  "Severe irAE"
)

hci002_survival_df <- prepare_hci002_survival_data(feature_col)

cox_os <- fit_multivariable_cox(
  df = hci002_survival_df,
  feature_col = feature_col,
  control_vars = control_vars,
  outcome_prefix = "OS"
)

cox_pfs <- fit_multivariable_cox(
  df = hci002_survival_df,
  feature_col = feature_col,
  control_vars = control_vars,
  outcome_prefix = "PFS"
)

result_df <- tibble(
  Variable = variable_labels,
  HR_OS = cox_os$HR,
  HR_OS_low = cox_os$HR_low,
  HR_OS_up = cox_os$HR_high,
  p_val_OS = cox_os$p_value,
  HR_PFS = cox_pfs$HR,
  HR_PFS_low = cox_pfs$HR_low,
  HR_PFS_up = cox_pfs$HR_high,
  p_val_PFS = cox_pfs$p_value
)

save_forest_plot(
  plot_data = tibble(
    mean = result_df$HR_OS,
    lower = result_df$HR_OS_low,
    upper = result_df$HR_OS_up,
    Variable = result_df$Variable,
    pval = result_df$p_val_OS
  ),
  file_name = "Forest_HCI002_actCD8Tcm_clinicalVARs_MHR_OS.pdf",
  x_label = "HR of OS"
)

save_forest_plot(
  plot_data = tibble(
    mean = result_df$HR_PFS,
    lower = result_df$HR_PFS_low,
    upper = result_df$HR_PFS_up,
    Variable = result_df$Variable,
    pval = result_df$p_val_PFS
  ),
  file_name = "Forest_HCI002_actCD8Tcm_clinicalVARs_MHR_PFS.pdf",
  x_label = "HR of PFS"
)

```


# SFigure 4f. HCI001 survival analysis by activated CD8Tcm abundance
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

hci001_predictor_df <- HCI001_all_info %>%
  select(`RedCap ID`, all_of(feature_col)) %>%
  drop_na() %>%
  distinct(`RedCap ID`, .keep_all = TRUE)

hci001_survival_info <- readxl::read_excel(file.path(data_dir, "HCI001_clinical_info.xlsx")) %>%
  select(`RedCap ID`, OS_time, OS_event, PFS_time, PFS_event) %>%
  distinct(`RedCap ID`, .keep_all = TRUE) %>%
  mutate(
    OS_time = as.numeric(OS_time) / 30,
    PFS_time = as.numeric(PFS_time) / 30,
    OS_event = as.numeric(OS_event),
    PFS_event = as.numeric(PFS_event)
  )

hci001_survival_df <- merge(
  hci001_predictor_df,
  hci001_survival_info,
  by = "RedCap ID",
  all.x = TRUE
) %>%
  filter(!is.na(.data[[feature_col]]))

for (phenotype in c("PFS", "OS")) {
  save_km_plot(
    df = hci001_survival_df,
    feature_col = feature_col,
    phenotype = phenotype,
    cohort_name = "HCI001",
    xlim_max = 50
  )
}

```


# AUC summary across irAE definitions in HCI001 and HCI002
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

irae_types <- c(
  "IRAEgrade01",
  "minIRAEgrade01",
  "IRAEactionable01",
  "IRAE_GI01",
  "IRAE_HB01",
  "IRAE_skin01",
  "IRAE_MSK01"
)

compute_irae_auc_table <- function(df, feature_col, irae_types) {
  bind_rows(lapply(irae_types, function(irae_type) {
    compute_auc_summary(df, outcome_col = irae_type, predictor_col = feature_col) %>%
      mutate(irAE_type = irae_type, .before = 1)
  })) %>%
    rename(
      irAE_AUC = AUC,
      irAE_AUC_lower = AUC_low,
      irAE_AUC_upper = AUC_high,
      irAE_pval = p_value
    )
}

AUC_irAEs_df_HCI002 <- compute_irae_auc_table(HCI002_all_info, feature_col, irae_types)
AUC_irAEs_df_HCI001 <- compute_irae_auc_table(HCI001_all_info, feature_col, irae_types)

```


# SFigure 4a. AUC comparison across irAE subtypes
```{r}
irae_label_map <- c(
  IRAE_GI01 = "Gastrointestinal",
  IRAE_HB01 = "Hepatobiliary",
  IRAE_skin01 = "Skin",
  IRAE_MSK01 = "Musculoskeletal"
)

auc_dataset_list <- list(
  HCI001 = AUC_irAEs_df_HCI001,
  HCI002 = AUC_irAEs_df_HCI002
)

for (dataset_name in names(auc_dataset_list)) {
  plot_df <- auc_dataset_list[[dataset_name]] %>%
    filter(irAE_type %in% names(irae_label_map)) %>%
    mutate(label = irae_label_map[irAE_type]) %>%
    rename(
      AUC = irAE_AUC,
      AUC_low = irAE_AUC_lower,
      AUC_high = irAE_AUC_upper,
      p_value = irAE_pval
    )

  save_auc_barplot(
    plot_df = plot_df,
    file_name = paste0("barplot_AUC_compare_irAE_types_", dataset_name, ".pdf")
  )
}

```

# SFigure 4c. Association between activated CD8Tcm abundance and baseline clinical variables in HCI002
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"
feature_label <- "% activated CD8Tcm"
dataset_name <- "HCI002"

plot_df <- HCI002_all_info %>%
  transmute(
    Predictor = .data[[feature_col]],
    Age = `Age at C1`,
    TMB = `TMB (mutation /mb)`,
    PDL1_TPS = `PD-L1 TPS`,
    PDL1_CPS = `PD-L1 CPS`,
    Gender = Gender,
    BRAF = case_when(
      is.na(`BRAF Mutation`) ~ "Unknown",
      grepl("V600E|V600K|Wildtype", `BRAF Mutation`) ~ `BRAF Mutation`,
      TRUE ~ "Other mut."
    )
  )

continuous_vars <- c(
  Age = "Age",
  TMB = "TMB (mut. / Mb)",
  PDL1_TPS = "PDL1 TPS (%)",
  PDL1_CPS = "PDL1 CPS (%)"
)

for (var_name in names(continuous_vars)) {
  p <- ggplot(plot_df, aes(x = .data[[var_name]], y = Predictor)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
    labs(x = continuous_vars[[var_name]], y = feature_label) +
    theme_classic() +
    theme(axis.text = element_text(color = "black"))

  ggsave(
    file.path(result_figure_dir, paste0("scatter_", dataset_name, "_", feature_col, "_vs_", var_name, ".pdf")),
    plot = p,
    width = 2.1,
    height = 2.4
  )
}

categorical_vars <- c("Gender", "BRAF")

for (var_name in categorical_vars) {
  p <- ggplot(plot_df, aes(x = .data[[var_name]], y = Predictor)) +
    geom_boxplot(outlier.shape = NA, fill = "lightgray", color = "black") +
    geom_jitter(width = 0.2, size = 1, aes(color = .data[[var_name]])) +
    labs(x = var_name, y = feature_label) +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )

  ggsave(
    file.path(result_figure_dir, paste0("barplot_", dataset_name, "_", feature_col, "_vs_", var_name, ".pdf")),
    plot = p,
    width = 1.5,
    height = 3
  )
}

wilcox_results <- bind_rows(lapply(categorical_vars, function(var_name) {
  test_df <- plot_df %>%
    mutate(Group = .data[[var_name]]) %>%
    mutate(Group = ifelse(var_name == "BRAF" & !is.na(Group), ifelse(Group == "Wildtype", "Wildtype", "Mutant"), Group)) %>%
    select(Predictor, Group) %>%
    drop_na()

  p_value <- if (n_distinct(test_df$Group) == 2) {
    wilcox.test(Predictor ~ Group, data = test_df)$p.value
  } else {
    NA_real_
  }

  tibble(Variable = var_name, P_value = p_value)
}))

wilcox_results

```

# SFigure 4b. AUC comparison across clinical subgroups in HCI002
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

subgroup_df <- HCI002_all_info %>%
  transmute(
    Feature = .data[[feature_col]],
    Phenotype = IRAEactionable01,
    Drug = Drug,
    ICB_exposure = ifelse(`ICI Naïve?` == "Yes", "ICB-naive", "ICB-experienced"),
    Stage = ifelse(grepl("IV", `Baseline Stage / TNM`), "Stage IV", "Stage I-III"),
    TherapyType = case_when(
      TherapyType == "Active" ~ "Active treatment",
      TherapyType == "Adjuvant" ~ "Adjuvant",
      TherapyType == "Neoadjuvant" ~ "Neoadjuvant",
      TRUE ~ TherapyType
    )
  )

subgroup_specs <- list(
  Drug = c("PD1/PDL1", "CTLA4-containing"),
  ICB_exposure = c("ICB-naive", "ICB-experienced"),
  Stage = c("Stage I-III", "Stage IV"),
  TherapyType = c("Active treatment", "Adjuvant")
)

auc_subgroup_results <- bind_rows(lapply(names(subgroup_specs), function(group_col) {
  bind_rows(lapply(subgroup_specs[[group_col]], function(group_level) {
    subgroup_df %>%
      filter(.data[[group_col]] == group_level) %>%
      compute_auc_summary(outcome_col = "Phenotype", predictor_col = "Feature") %>%
      mutate(
        subgroup = group_col,
        label = group_level,
        .before = 1
      )
  }))
}))

for (group_col in names(subgroup_specs)) {
  plot_df <- auc_subgroup_results %>%
    filter(subgroup == group_col) %>%
    mutate(label = paste0(label, "\n(n=", n, ")"))

  save_auc_barplot(
    plot_df = plot_df,
    file_name = paste0("barplot_AUC_compare_subgroups_", group_col, "_HCI002.pdf")
  )
}

```



