---
title: "Predicting irAE with Olink serum proteins"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

####### This script makes the below displays for the paper
## Figure 4D-F

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE, results = 'markup')
```

# load required package
```{r}

library(data.table)
library(ggplot2)
library(readxl)
library(viridis)
library(tidyr)
library(pROC)
library(verification)
library(reshape2)
library(ggpubr)

library(dplyr)
library(survival)
library(survminer)


library(glmnet)
library(caret)
library(pROC)
library(ranger)
library(readr)

library(AnnotationDbi)
library(org.Hs.eg.db) 
library(ReactomePA)
library(enrichplot)
library(gridExtra)
library(clusterProfiler)
library(msigdbr)

library(stringr)

```


# parameters and constants and common functions
```{r}

my.cols <- c("#1c4f27", "#81144e", "#79831e", "#00305d", "#9c1915", "black", "grey", "#f58231", "#e6194b", "#3cb44b", "#42d4f4") 

# function mapping gene IDs and Symbols
convert_gene_ids_to_symbols <- function(gene_ids, mapping) {
  lapply(gene_ids, function(x) {
    entrez_ids <- unlist(strsplit(x, "/"))
    gene_symbols <- names(mapping)[match(entrez_ids, mapping)]
    paste(gene_symbols, collapse = ", ")
  })
}

# Create a function to make olink gene names unique
make_unique <- function(names) {
  # Initialize an empty vector to store unique names
  unique_names <- character(length(names))
  # Create a named vector to count occurrences
  name_count <- table(names)
  # Track the current count of each name
  name_tracker <- setNames(rep(0, length(name_count)), names(name_count))
  # Iterate over the names and create unique ones
  for (i in seq_along(names)) {
    name <- names[i]
    # If this name is duplicated, append the appropriate number
    if (name_tracker[name] > 0) {
      new_name <- paste0(name, "_", name_tracker[name] + 1)
    } else {
      new_name <- name
    }
    # Assign the unique name
    unique_names[i] <- new_name
    # Increment the tracker for this name
    name_tracker[name] <- name_tracker[name] + 1
  }
  return(unique_names)
}

```

# Figure 4D. AUC comparison of CXCL9, CA6 in HCI002 at different time points
```{r}

proteins_mapping <- setNames(HCI002_map_df$Assay, HCI002_map_df$OlinkID)  # Named vector: OlinkID -> gene symbol

proteins_test <- c("CXCL9", "CA6", "CXCL10")
proteins_test_OID <- HCI002_map_df$OlinkID[HCI002_map_df$Assay %in% proteins_test]

joined_string <- paste(proteins_test_OID, collapse = "|")
HCI002_feature_df <- HCI002_all_info0[grepl(joined_string, colnames(HCI002_all_info0))]
time_pts <- c("baseline", "ontreat", "FC_on_pre")
joined_string <- paste(time_pts, collapse = "|")
HCI002_feature_df <- HCI002_feature_df[grepl(joined_string, colnames(HCI002_feature_df))]
colnames(HCI002_feature_df) <- gsub("_OLINK", "", colnames(HCI002_feature_df))
# Replace OlinkID substrings with gene symbols
for (oid in names(proteins_mapping)) {
  gene <- proteins_mapping[[oid]]
  colnames(HCI002_feature_df) <- gsub(oid, gene, colnames(HCI002_feature_df))
}

HCI002_feature_df$CXCL9_CA6_baseline <- HCI002_feature_df$CXCL9_baseline - HCI002_feature_df$CA6_baseline
HCI002_feature_df$CXCL9_CA6_ontreat <- HCI002_feature_df$CXCL9_ontreat - HCI002_feature_df$CA6_ontreat
HCI002_feature_df$CXCL9_CA6_FC_on_pre <- HCI002_feature_df$CXCL9_FC_on_pre - HCI002_feature_df$CA6_FC_on_pre

phenotype_name = "IRAEactionable01" # IRAEactionable01: column 55;  Response01: column 51
phenotype_df = HCI002_all_info0[,phenotype_name, drop = F]
HCI002_filtered_info = cbind(HCI002_feature_df, phenotype_df)


####### calculate AUCs of CXCL9/CA6 in HCI002 at different time points
# Separate predictors and phenotype
predictor_df <- HCI002_filtered_info[ , -ncol(HCI002_filtered_info)]
phenotype <- HCI002_filtered_info[[ncol(HCI002_filtered_info)]]

# Prepare result storage
result_df <- data.frame(
  Feature = character(),
  AUC = numeric(),
  AUC_low = numeric(),
  AUC_up = numeric(),
  P_value = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each predictor column
for (feature_name in colnames(predictor_df)) {
  values <- predictor_df[[feature_name]]
  
  # Keep only non-NA pairs
  valid_idx <- !is.na(values) & !is.na(phenotype)
  if (sum(valid_idx) < 3 || length(unique(phenotype[valid_idx])) < 2) {
    next  # skip if not enough data or classes
  }
  
  values_valid <- values[valid_idx]
  pheno_valid <- phenotype[valid_idx]
  
  # Compute AUC
  roc_result <- roc(pheno_valid, values_valid, direction = "<", levels = c(0, 1), quiet = TRUE)
  AUC_ci <- ci.auc(roc_result)
  auc_val <- AUC_ci[2]
  auc_low_val <- AUC_ci[1]
  auc_up_val <- AUC_ci[3]
  
  # Compute p-value vs random (0.5)
  suppressMessages({
    roc_test <- roc.test(
      roc_result,
      roc(pheno_valid, rep(0.5, length(pheno_valid)), quiet = TRUE),
      method = "delong"
    )
  })
  auc_pval <- roc_test$p.value
  
  # Store result
  result_df <- rbind(result_df, data.frame(
    Feature = feature_name,
    AUC = auc_val,
    AUC_low = auc_low_val,
    AUC_up = auc_up_val,
    P_value = auc_pval
  ))
}
print(result_df)

result_df <- result_df[!grepl("CXCL9_CA6", result_df$Feature), ]

####### barplot AUCs of CXCL9/CA6 in HCI002 at different time points
# Extract protein and timepoint from Feature
plot_df <- result_df %>%
  mutate(
    Protein = case_when(
      grepl("CXCL9_CA6", Feature) ~ "CXCL9 - CA6",
      grepl("CXCL9", Feature) ~ "CXCL9",
      grepl("CA6", Feature) ~ "CA6"
    ),
    Timepoint = case_when(
      grepl("baseline", Feature) ~ "Pre-treatment",
      grepl("ontreat", Feature) ~ "On-treatment",
      grepl("FC_on_pre", Feature) ~ "Fold-change",
      TRUE ~ "unknown"
    ),
    irAE_pval_str = sapply(P_value, function(p) {
      if (p == 0) return("p==0")
      exp10  <- floor(log10(p))
      coef   <- signif(p / 10^exp10, 2)
      sprintf("p==%s%%*%%10^%s", coef, exp10)
    })
  )

# Identify rows where AUC < 0.5
idx <- plot_df$AUC < 0.5

# Apply 1 - x to all three columns for those rows
plot_df$AUC[idx] <- 1 - plot_df$AUC[idx]
plot_df$AUC_low[idx] <- 1 - plot_df$AUC_low[idx]
plot_df$AUC_up[idx] <- 1 - plot_df$AUC_up[idx]

# Set factor levels to control bar order
plot_df$Timepoint <- factor(plot_df$Timepoint, levels = c("Pre-treatment", "On-treatment", "Fold-change"))
plot_df$Protein <- factor(plot_df$Protein, levels = c("CXCL9", "CA6", "CXCL9 - CA6"))

# Define custom colors
custom_colors <- c(
  "CXCL9"      = "#1B9E77",  # Teal-green
  "CA6"        = "#D95F02",  # Warm orange
  "CXCL9 - CA6" = "#7570B3"  # Elegant purplish blue
)

plot_df <- na.omit(plot_df)

# Plot
fig_width = 3*0.8
fig_height = 4*0.8
pdf_file <- paste0(result_figure_dir, "barplot_AUC_compare_CXCL9_CA6_timePts_HCI002.pdf")
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)

ggplot(plot_df, aes(x = Timepoint, y = AUC, fill = Protein)) +
  geom_bar(stat = "identity", width = 0.7, position = position_dodge(width = 0.8)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey", linewidth = 0.5) +
  geom_errorbar(
    aes(ymin = AUC_low, ymax = AUC_up),
    width = 0.2,
    color = "black",
    linewidth = 0.5,
    position = position_dodge(width = 0.8)
  ) +
  scale_fill_manual(values = custom_colors) +
  labs(x = "", y = "AUC") +
  ylim(0, 0.8) + 
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text.y = element_text(color = "black",size = 10),
    axis.text.x = element_text(color = "black", angle = 15, hjust = 1,size = 10),
    legend.position = "top",
    legend.title = element_blank(),
    plot.margin = margin(t = 5, r = 20, b = 5, l = 15)  
  )

dev.off()


```

# Figure 4E. K-M survival curves to test whether the CXCL9 signature could predict time-to-severe irAE (HCI002) 
```{r}

data_test_df <- cbind(HCI002_feature_df, HCI002_all_info0[c("IRAEactionable01", "RECIST01", "Response01", "IRAE_GI01", "IRAE_HB01", "IRAE_skin01", "IRAE_MSK01", "IRAEgrade01")])

feature_test <- "CXCL9_FC_on_pre" 
quantile_cutoff <- 0.5
phenotype_str <- "Nonsevere_irAE" 
y_label_str <- "Severe-irAE-free \nsurvival probability"

# get the predictor from the HCI002_all_info0
predictor_df <- cbind(HCI002_all_info0[c("Vial ID", "patient_therapy")], data_test_df[feature_test]) 
predictor_df$uniqueID <- paste0(predictor_df$patient_therapy, "_", predictor_df$`Vial ID`)
predictor_df <- predictor_df[!duplicated(predictor_df$uniqueID), ]


## load survival information
survival_info_df <- read_excel("HCI002 Teiko clinical info 7.12.23_clean_newest.xlsx")
survival_info_df <- survival_info_df[c("Vial ID", "Record ID", "Systemic Therapy Name", "Systemic Therapy Type", "OS_time", "OS_event", "PFS_time", "PFS_event", "Nonsevere_irAE_time", "Nonsevere_irAE_event")]
survival_info_df$uniqueID <- paste0(survival_info_df$`Record ID`, "_", survival_info_df$`Systemic Therapy Name`, "_", survival_info_df$`Systemic Therapy Type`, "_", survival_info_df$`Vial ID`)

survival_all_info_HCI002 <- merge(predictor_df, survival_info_df, by = "uniqueID", x.all = T)

survival_all_info_HCI002 <- survival_all_info_HCI002[!is.na(survival_all_info_HCI002[[feature_test]]), ]

## convert days to months
survival_all_info_HCI002$OS_time <- survival_all_info_HCI002$OS_time / 30
survival_all_info_HCI002$PFS_time <- survival_all_info_HCI002$PFS_time / 30
survival_all_info_HCI002$Nonsevere_irAE_time <- survival_all_info_HCI002$Nonsevere_irAE_time / 30

############# Survival test for OS / PFS / time-irAEmax
Score_Pred = survival_all_info_HCI002[[feature_test]]
Score=Score_Pred
exp_cutoff = quantile(Score_Pred, quantile_cutoff)
#exp_cutoff = 0.04
Score[Score_Pred>exp_cutoff]="High"
Score[Score_Pred<=exp_cutoff]="Low"
Score <- factor(Score)
Score <- relevel(Score, ref = "Low")
cancerData=data.frame(Score,survival_all_info_HCI002[[paste0(phenotype_str, "_time")]],survival_all_info_HCI002[[paste0(phenotype_str, "_event")]])
colnames(cancerData) = c("Score","OS_time","OS_event")

sfit <- survfit(Surv(OS_time, OS_event) ~ Score, data=cancerData)
scox <- coxph(Surv(OS_time, OS_event)~Score, data=cancerData)
scox_coef = summary(scox)$coefficients
HR_value = scox_coef[2] # hazard ratio
Z_value=scox_coef[4]
P_value=scox_coef[5]
HR_CI = summary(scox)$conf.int[3:4]

##### plot
fontSize = 12
survp=ggsurvplot(
  sfit,
  data = cancerData,
  size = 1,                 # change line size
  palette =
    c("#00305d", "#9c1915"),# custom color palettes
  conf.int = FALSE,          # Add confidence interval
  pval = FALSE,              # Add p-value
  xlim = c(-3,70),
  ylim=c(0,1),
  xlab = "Time (months)", ylab= y_label_str,#
  break.time.by = 10,
  risk.table=TRUE,
  risk.table.height = 0.25, # Useful to change when you have multiple groups
  risk.table.pos="out",
  risk.table.col="black",
  risk.table.y.text = FALSE,
  tables.y.text = FALSE,
  tables.theme = theme_cleantable(),
  legend.labs =c("Low", "High"),    # Change legend labels
  legend.title="",
  legend = c(0.75, 0.975), # legend relative position
  font.main = c(fontSize),
  font.caption = c(fontSize),
  font.legend = c(fontSize),
  font.tickslab = c(fontSize),
  font.x = c(fontSize),
  font.y = c(fontSize),
  ggtheme = theme(legend.background = element_rect(fill = NA, color=NA),legend.key = element_rect(fill = NA, color=NA),
                 plot.margin = unit(c(0.2, 0.2, 0, 0.2),"cm"),
                 panel.background = element_rect(fill = "white"),
                 panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                 panel.border = element_blank(),axis.line = element_line(colour = "black"),
                 axis.text.x = element_text(colour="black"),axis.text.y = element_text(colour="black")),  # top, right, bot, left
) 
survp$plot = survp$plot+
            ggplot2::annotate("text", x=0, y=0.15, label=paste0('HR',' = ',round(HR_value,2),' (', round(HR_CI[1],2),'-',round(HR_CI[2],2),')','\n','p = ',sprintf('%.3f', P_value)),size = 5, hjust = 0)
pdf(paste0(result_figure_dir, "KM_curve_",phenotype_str,"_FoldChange_",feature_test,".pdf"),width=3.3, height=3)
print(survp, newpage = FALSE)
dev.off()

```


# Prepare input of different datasets for AUC
```{r}

### HCI001
proteins_mapping <- setNames(HCI001_map_df$Assay, HCI001_map_df$OlinkID)  # Named vector: OlinkID -> gene symbol
proteins_test <- c("CXCL9", "CXCL10", "CXCL11", "IFNG", "CXCL13", "IL17A", "IL10", "IL12A", "IL21", "B4GALT1", "CD276", "CD70", "FLT3LG", "FSTL3", "IL2RB", "KRT19", "PGF", "CA6")
proteins_test_OID <- HCI001_map_df$OlinkID[HCI001_map_df$Assay %in% proteins_test]

joined_string <- paste(proteins_test_OID, collapse = "|")
HCI001_feature_df <- HCI001_all_info0[grepl(joined_string, colnames(HCI001_all_info0))]
time_pts <- c("FC_irAEmax_pre")
joined_string <- paste(time_pts, collapse = "|")
HCI001_feature_df <- HCI001_feature_df[grepl(joined_string, colnames(HCI001_feature_df))]
colnames(HCI001_feature_df) <- gsub("_OLINK", "", colnames(HCI001_feature_df))
# Replace OlinkID substrings with gene symbols
for (oid in names(proteins_mapping)) {
  gene <- proteins_mapping[[oid]]
  colnames(HCI001_feature_df) <- gsub(oid, gene, colnames(HCI001_feature_df))
}
HCI001_feature_df$CXCL9_CA6_FC <- HCI001_feature_df$CXCL9_FC_irAEmax_pre - HCI001_feature_df$CA6_FC_irAEmax_pre
joined_string <- paste(Integrated_proteins_vec, collapse = "|")
HCI001_feature_df$Integrated_proteins_FC <- rowMeans(HCI001_feature_df[grepl(joined_string, colnames(HCI001_feature_df))])
colnames(HCI001_feature_df) <- gsub("_irAEmax_pre", "", colnames(HCI001_feature_df))
HCI001_feature_df$irAE_severe <- HCI001_all_info0$IRAEactionable01 # dim: 69 4

HCI001_feature_df %>%
  summarise(across(everything(), ~ median(.x, na.rm = TRUE)))
print(colnames((HCI001_feature_df)))
HCI001_feature_df$Patient_ID <- HCI001_all_info0$`RedCap ID`

### HCI002
proteins_mapping <- setNames(HCI002_map_df$Assay, HCI002_map_df$OlinkID)  # Named vector: OlinkID -> gene symbol
proteins_test_OID <- HCI002_map_df$OlinkID[HCI002_map_df$Assay %in% proteins_test]

joined_string <- paste(proteins_test_OID, collapse = "|")
HCI002_feature_df <- HCI002_all_info0[grepl(joined_string, colnames(HCI002_all_info0))]
time_pts <- c("FC_on_pre")
joined_string <- paste(time_pts, collapse = "|")
HCI002_feature_df <- HCI002_feature_df[grepl(joined_string, colnames(HCI002_feature_df))]
colnames(HCI002_feature_df) <- gsub("_OLINK", "", colnames(HCI002_feature_df))
# Replace OlinkID substrings with gene symbols
for (oid in names(proteins_mapping)) {
  gene <- proteins_mapping[[oid]]
  colnames(HCI002_feature_df) <- gsub(oid, gene, colnames(HCI002_feature_df))
}
HCI002_feature_df$CXCL9_CA6_FC <- HCI002_feature_df$CXCL9_FC_on_pre - HCI002_feature_df$CA6_FC_on_pre
joined_string <- paste(Integrated_proteins_vec, collapse = "|")
HCI002_feature_df$Integrated_proteins_FC <- rowMeans(HCI002_feature_df[grepl(joined_string, colnames(HCI002_feature_df))])
colnames(HCI002_feature_df) <- gsub("_on_pre", "", colnames(HCI002_feature_df))
HCI002_feature_df$irAE_severe <- HCI002_all_info0$IRAEactionable01 # dim: 167   4

HCI002_feature_df %>%
  summarise(across(everything(), ~ median(.x, na.rm = TRUE)))
print(colnames((HCI002_feature_df)))
HCI002_feature_df$Patient_ID <- HCI002_all_info0$patient_therapy


## Nunez cohort1
Nicolas_melanoma_info0 <- Nicolas_all_info0[Nicolas_all_info0$Cancer == "Melanoma", ]
string_match <- paste0(paste(proteins_test, collapse = "_FC_T2_T1|"), "_FC_T2_T1")
colnames(Nicolas_melanoma_info0) <- gsub("gamma", "G", colnames(Nicolas_melanoma_info0))
Nicolas1_feature_df <- Nicolas_melanoma_info0[ , grepl(string_match, colnames(Nicolas_melanoma_info0), ignore.case = TRUE)]
Nicolas1_feature_df$CXCL9_CA6_FC <- Nicolas_melanoma_info0$CXCL9_FC_T2_T1 # only CXCL9 measured, CA6 not measured
joined_string <- paste(Integrated_proteins_vec, collapse = "|")
Nicolas1_feature_df$Integrated_proteins_FC <- rowMeans(Nicolas1_feature_df[grepl(joined_string, colnames(Nicolas1_feature_df))])
colnames(Nicolas1_feature_df) <- gsub("_T2_T1", "", colnames(Nicolas1_feature_df))
Nicolas1_feature_df$`Patient ID` <- Nicolas_melanoma_info0$Patient.ID

irAE_info_df = read_excel("Nunez1_Patient_character.xlsx", sheet = "Patient charact (Discovery)")
irAE_info_df = irAE_info_df[c("Patient ID", "Autoimmune toxicities (grade)")]
irAE_info_df = irAE_info_df[!is.na(irAE_info_df$`Patient ID`), ]
irAE_info_df$irAE_severe = NA
irAE_info_df$irAE_severe[grepl("1|2|^-$", irAE_info_df$`Autoimmune toxicities (grade)`)] = 0
irAE_info_df$irAE_severe[grepl("3|4", irAE_info_df$`Autoimmune toxicities (grade)`)] = 1
irAE_info_df$`Autoimmune toxicities (grade)` = NULL
Nicolas1_feature_df = merge(Nicolas1_feature_df, irAE_info_df, by = "Patient ID") # dim: 25 4
print(colnames((Nicolas1_feature_df)))
Nicolas1_feature_df$Patient_ID <- Nicolas_melanoma_info0$`Patient.ID_new`


## Nunez cohort2
### Olink data
Olinkdata_filename = paste0(data_dir, "Proteomics_non_log_normalized.xlsx") 
Olinkdata_df <- read_excel(Olinkdata_filename)
Olinkdata_df <- Olinkdata_df[Olinkdata_df$Cancer == "Melanoma", ]
Olinkdata_df = Olinkdata_df[c("Patient_ID", "CXCL11 (T2/T1)", "CXCL9 (T2/T1)", "CXCL10 (T2/T1)", "IFNg (T2/T1)")]
colnames(Olinkdata_df) <- c("Patient_ID", "CXCL11_FC", "CXCL9_FC", "CXCL10_FC", "IFNG_FC")
Olinkdata_df$Patient_ID = toupper(Olinkdata_df$Patient_ID)

# get CyTEK severe irAE annotation
irAE_info_df = read_excel("Nunez2_Patient_character.xlsx", sheet = "Patient charact (Validation)")
irAE_info_df$Patient_ID <- irAE_info_df$`Patient ID`
irAE_info_df = irAE_info_df[c("Patient_ID", "Autoimmune toxicities (grade)")]
irAE_info_df$irAE_severe = NA
irAE_info_df$irAE_severe[grepl("1|2|^-$", irAE_info_df$`Autoimmune toxicities (grade)`)] = 0
irAE_info_df$irAE_severe[grepl("3|4", irAE_info_df$`Autoimmune toxicities (grade)`)] = 1
irAE_info_df$`Autoimmune toxicities (grade)` = NULL
Nicolas2_feature_df = merge(Olinkdata_df, irAE_info_df, by = "Patient_ID")
Nicolas2_feature_df$CXCL9_CA6_FC <- Nicolas2_feature_df$`CXCL9_FC` # only CXCL9 measured, CA6 not measured. dim: 123 7
joined_string <- paste(Integrated_proteins_vec, collapse = "|")
Nicolas2_feature_df$Integrated_proteins_FC <- rowMeans(Nicolas2_feature_df[grepl(joined_string, colnames(Nicolas2_feature_df))])

```

# Calculate AUCs of all predictors across all datasets
```{r}

# data frame of AUCs for different biomarkers
AUC_result_df <- data.frame(
  dataset = c("HCI002", "HCI001", "Nicolas1", "Nicolas2"), # Initialize dataset column with values
  CXCL9_CA6 = NA,
  CXCL9 = NA,
  CXCL10 = NA,
  CXCL11 = NA,
  IFNG = NA,
  CXCL13 = NA,
  IL17A = NA,
  IL10 = NA,
  IL12A = NA,
  IL21 = NA,
  B4GALT1 = NA,
  CD276 = NA,
  CD70 = NA,
  FLT3LG = NA,
  FSTL3 = NA,
  IL2RB = NA,
  KRT19 = NA,
  PGF = NA,
  Integrated_proteins = NA
)

rownames(AUC_result_df) = AUC_result_df$dataset
AUC_result_df$dataset = NULL


#### compare AUCs of different biomarkers in HCI001
all_single_predictors <- c(paste(colnames(AUC_result_df)[1:(ncol(AUC_result_df)-1)], "FC", sep = "_"))
Integrated_proteins_predictor <- c(paste(Integrated_proteins_vec, "FC", sep = "_"))
data_df <- as.data.frame(HCI001_feature_df)
response <- data_df$irAE_severe
AUC_vec = c()
for (sp in all_single_predictors){
  if (sp %in% colnames(data_df)){
    predictor <- data_df[[sp]]
    roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
    AUC = as.numeric(auc(roc_obj))
    AUC_vec = c(AUC_vec, AUC)
  }else{
    AUC_vec = c(AUC_vec, NA)
  }
}
predictor = rowSums(log1p(data_df[colnames(data_df) %in% Integrated_proteins_predictor]))
roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
AUC = as.numeric(auc(roc_obj))
AUC_vec = c(AUC_vec, AUC)
AUC_result_df["HCI001", ] = AUC_vec


#### compare AUCs of different biomarkers in HCI002
data_df <- as.data.frame(HCI002_feature_df)
response <- data_df$irAE_severe
AUC_vec = c()
for (sp in all_single_predictors){
  if (sp %in% colnames(data_df)){
    predictor <- data_df[[sp]]
    roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
    AUC = as.numeric(auc(roc_obj))
    AUC_vec = c(AUC_vec, AUC)
  }else{
    AUC_vec = c(AUC_vec, NA)
  }
}
predictor = rowSums(log1p(data_df[colnames(data_df) %in% Integrated_proteins_predictor]))
roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
AUC = as.numeric(auc(roc_obj))
AUC_vec = c(AUC_vec, AUC)
AUC_result_df["HCI002", ] = AUC_vec


#### compare AUCs of different biomarkers in Nicolas1
data_df <- as.data.frame(Nicolas1_feature_df)
response <- data_df$irAE_severe
AUC_vec = c()
for (sp in all_single_predictors){
  if (sp %in% colnames(data_df)){
    predictor <- data_df[[sp]]
    roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
    AUC = as.numeric(auc(roc_obj))
    AUC_vec = c(AUC_vec, AUC)
  }else{
    AUC_vec = c(AUC_vec, NA)
  }
}
predictor = rowSums(log1p(data_df[colnames(data_df) %in% Integrated_proteins_predictor]))
roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
AUC = as.numeric(auc(roc_obj))
AUC_vec = c(AUC_vec, AUC)
AUC_result_df["Nicolas1", ] = AUC_vec


#### compare AUCs of different biomarkers in Nicolas2
data_df <- as.data.frame(Nicolas2_feature_df)
response <- data_df$irAE_severe
AUC_vec = c()
for (sp in all_single_predictors){
  if (sp %in% colnames(data_df)){
    predictor <- data_df[[sp]]
    roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
    AUC = as.numeric(auc(roc_obj))
    AUC_vec = c(AUC_vec, AUC)
  }else{
    AUC_vec = c(AUC_vec, NA)
  }
}
predictor = rowSums(log1p(data_df[colnames(data_df) %in% Integrated_proteins_predictor]))
roc_obj <- roc(response, predictor, direction="<",levels=c(0, 1))
AUC = as.numeric(auc(roc_obj))
AUC_vec = c(AUC_vec, AUC)
AUC_result_df["Nicolas2", ] = AUC_vec


```


# Figure 4F. Plot AUC comparison of different predictors (across 4 datasets)
```{r}

AUC_result_df$IL12A = NULL
AUC_result_df$IL21 = NULL
AUC_result_df$CXCL9_CA6 = NULL
AUC_result_df$Integrated_proteins = NULL

### Calculate mean values for each column, omitting NAs
mean_values <- colMeans(AUC_result_df, na.rm = TRUE)
AUC_result_df["Mean", ] <- mean_values

ordered_cols <- order(-as.numeric(AUC_result_df["Mean", ]))
AUC_result_df <- AUC_result_df[, ordered_cols]

plot_vars = colnames(AUC_result_df)
AUC_result_df2 = AUC_result_df[1:4,]
AUC_result_df2 = AUC_result_df2[colnames(AUC_result_df2) %in% plot_vars]

# Reshape the data to a long format
AUC_result_df2$Dataset <- rownames(AUC_result_df2)
AUC_result_df_long <- melt(AUC_result_df2, id.vars = "Dataset", variable.name = "GeneSignature", value.name = "AUC")

# clean the labels
AUC_result_df_long$Dataset[AUC_result_df_long$Dataset == "Nicolas1"] <- "Nunez 1"
AUC_result_df_long$Dataset[AUC_result_df_long$Dataset == "Nicolas2"] <- "Nunez 2"
AUC_result_df_long$Dataset = factor(AUC_result_df_long$Dataset, levels = c("HCI001","HCI002","Nunez 1","Nunez 2"))

AUC_result_df_long$GeneSignature <- gsub("_", " ",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("CXCL9 CA6", "CXCL9 - CA6",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("Ki67", "Ki67+",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("CD38", "CD38+",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("CM", "cm",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("EM", "em",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("Eff", "emra",AUC_result_df_long$GeneSignature)
AUC_result_df_long$GeneSignature <- gsub("IFNG", "IFN-gamma",AUC_result_df_long$GeneSignature)

# Calculate the mean AUC for each gene signature
gene_signature_means <- aggregate(AUC ~ GeneSignature, data = AUC_result_df_long, FUN = mean)

# Order the gene signatures by mean AUC from high to low
ordered_gene_signatures <- as.character(gene_signature_means$GeneSignature[order(gene_signature_means$AUC, decreasing = TRUE)])

# Create a boxplot with different colors for each dataset
pdf_file <- paste0(result_figure_dir,paste0("barplot_AUC_compare_AllDatasets_Olink_FoldChange.pdf")) 
fig_width = 4.5*1.2
fig_height = 3.3*1.2
dot_size <- 2
pdf(pdf_file, onefile=FALSE, width = fig_width,height=fig_height)
ggplot(AUC_result_df_long, aes(x = GeneSignature, y = AUC)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge", fill = "white", color = "black") +  
  geom_jitter(aes(color = Dataset), position = position_jitter(0.2), size = dot_size) +  
  scale_x_discrete(limits = ordered_gene_signatures) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey", linewidth = 0.5) +
  labs(title = "", x = "", y = "AUC") +
  ylim(0, 1) +
  scale_color_manual(values = my.cols, name = "") +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color = "black"),
        axis.line.x = element_line(color = "black"),
        axis.ticks.y = element_line(color = "black"),
        axis.ticks.x = element_line(color = "black"),
        axis.text.y = element_text(color = "black"),
        axis.text.x = element_text(color = "black", angle = 50, hjust = 1),
        legend.spacing.y = unit(0.1, "cm"),  # reduce vertical space
        legend.box.spacing = unit(0.1, "cm"), # optional: spacing between legend and plot
        panel.border = element_blank(), legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = dot_size), nrow = 4))  
dev.off()

```
