---
title: "Cytometric and Integrated biomarker analysis across immunotherapy cohorts"
author: "Tiangen Chang <tiangen.chang@sjtu.edu.cn>"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Integrated biomarker analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "markup", warning = FALSE, message = FALSE)
source("utils.R")
```

####### This script makes the below displays for the paper
## Figure 3C, I-K
## Figure 4F
## Figure 5B-F
## SFigure 9A
## SFigure 11
## SFigure 12
## SFigure 13
## SFigure 14A,B,C (right)
## SFigure 15
## SFigure 16C
## SFigure 17
## STables 9-11



## Load data
```{r}
main_feature <- "CD8T_CM_tripletORp_ratio_pre"

HCI002_data <- read_cohort("HCI002_info_save.csv")
HCI001_data <- read_cohort("HCI001_info_save.csv")
Nicolas_FACS_data <- read_cohort("Nicolas_all_info_FACS_save.csv", melanoma_only = TRUE)
Nicolas_Cytek_data <- read_cohort("Nicolas_all_info_Cytek_save.csv", melanoma_only = TRUE)
Nicolas_CyTOF_data <- read_cohort("Nicolas_all_info_CyTOF_save.csv", melanoma_only = TRUE)

datasets <- list(
  HCI002 = HCI002_data,
  HCI001 = HCI001_data,
  Nicolas_FACS = Nicolas_FACS_data,
  Nicolas_Cytek = Nicolas_Cytek_data,
  Nicolas_CyTOF = Nicolas_CyTOF_data
)

datasets <- lapply(datasets, ensure_feats, feats = all_feats)
datasets <- lapply(datasets, make_severe_irAE, out_col = pheno_feat)

```


# Figure 3C (phenotype_type = "actionable") and SFigure 14A (right; phenotype_type = "conventional"). SFigure 15A.
```{r}

HCI002_data <- datasets$HCI002
HCI001_data <- datasets$HCI001
Nicolas_FACS_data <- datasets$Nicolas_FACS
Nicolas_Cytek_data <- datasets$Nicolas_Cytek
Nicolas_CyTOF_data <- datasets$Nicolas_CyTOF

for (phenotype_type in c('actionable', 'conventional', 'any', 'drop_mild')) {
  HCI002_data <- set_irAE_outcome(HCI002_data, phenotype_type)
  HCI001_data <- set_irAE_outcome(HCI001_data, phenotype_type)

  auc_datasets <- list(
    HCI002 = HCI002_data,
    HCI001 = HCI001_data,
    Nicolas_FACS = Nicolas_FACS_data,
    Nicolas_Cytek = Nicolas_Cytek_data,
    Nicolas_CyTOF = Nicolas_CyTOF_data
  )

  auc_df <- sapply(all_feats, function(feat) {
    sapply(auc_datasets, compute_single_feature_auc, y_col = pheno_feat, feature_col = feat)
  }) %>%
    as.data.frame()

  feature_order <- c(
    "CD8T_CM_tripletORp_ratio_pre",
    "CD8T_KI67p_ratio_pre",
    "CD4T_KI67p_ratio_pre",
    "Treg_KI67p_ratio_pre",
    "CD4TM_tripletORp_ratio_pre",
    "CD8T_EMRA_CD38p_ratio_pre",
    "CD4T_EM_tripletORp_ratio_pre",
    "CD4T_EM_ratio_pre",
    "CD4T_Naive_ratio_pre"
  )

  feature_labels <- c(
    "Activated CD8Tcm",
    "Ki67+ CD8T",
    "Ki67+ CD4T",
    "Ki67+ Treg",
    "Activated CD4Tm",
    "CD38+ CD8Temra",
    "Activated CD4Tem",
    "CD4Tem",
    "Naive CD4T"
  )

  save_auc_comparison(
    auc_df = auc_df,
    selected_features = feature_order,
    labels = feature_labels,
    file_name = paste0("barplot_AUC_compare_AllDatasets_", phenotype_type, ".pdf"),
    width = 4.5 * 1.2,
    height = 3.3 * 1.2
  )
  
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$CD4T_Naive_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$CD8T_EMRA_CD38p_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$CD8T_KI67p_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$CD4TM_tripletORp_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio[1:5], auc_df$CD4T_EM_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$CD4T_KI67p_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$Treg_KI67p_ratio_pre[1:5], paired = T)
  t.test(auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], auc_df$CD4T_EM_tripletORp_ratio_pre[1:5], paired = T)
}

```


# Figure 4F. AUC bar plots for Olink features (fold change of proteins as predictors of severe irAE). SFigure 14B (right; phenotype_type = "conventional"). SFigure 15B.
```{r}

HCI002_data <- datasets$HCI002
HCI001_data <- datasets$HCI001
Nicolas_FACS_data <- datasets$Nicolas_FACS
Nicolas_Cytek_data <- datasets$Nicolas_Cytek
Nicolas_CyTOF_data <- datasets$Nicolas_CyTOF

OLINK_feats <- c(
  "CXCL9_FC",
  "CXCL10_FC",
  "CXCL11_FC",
  "IFNG_FC",
  "CXCL13_FC",
  "IL17A_FC",
  "IL10_FC",
  #"IL12A_FC",
  #"IL21_FC",
  "B4GALT1_FC",
  "CD276_FC",
  "CD70_FC",
  "FLT3LG_FC",
  "FSTL3_FC",
  "IL2RB_FC",
  "KRT19_FC",
  "PGF_FC"
)

for (phenotype_type in c('actionable', 'conventional', 'any', 'drop_mild')) {
  HCI002_data <- set_irAE_outcome(HCI002_data, phenotype_type)
  HCI001_data <- set_irAE_outcome(HCI001_data, phenotype_type)

  auc_datasets <- list(
    HCI002 = HCI002_data,
    HCI001 = HCI001_data,
    Nicolas_FACS = Nicolas_FACS_data,
    Nicolas_Cytek = Nicolas_Cytek_data
  )

  auc_df <- sapply(OLINK_feats, function(feat) {
    sapply(auc_datasets, compute_single_feature_auc, y_col = pheno_feat, feature_col = feat)
  }) %>%
    as.data.frame()

  feature_order <- c(
    "CXCL9_FC",
    "CXCL10_FC",
    "CXCL11_FC",
    "IFNG_FC",
    "CXCL13_FC",
    "IL17A_FC",
    "IL10_FC",
    #"IL12A_FC",
    #"IL21_FC",
    "B4GALT1_FC",
    "CD276_FC",
    "CD70_FC",
    "FLT3LG_FC",
    "FSTL3_FC",
    "IL2RB_FC",
    "KRT19_FC",
    "PGF_FC"
  )

  feature_labels <- c(
    "CXCL9",
    "CXCL10",
    "CXCL11",
    "IFNG",
    "CXCL13",
    "IL17A",
    "IL10",
    #"IL12A",
    #"IL21",
    "B4GALT1",
    "CD276",
    "CD70",
    "FLT3LG",
    "FSTL3",
    "IL2RB",
    "KRT19",
    "PGF"
  )

  save_auc_comparison(
    auc_df = auc_df,
    selected_features = feature_order,
    labels = feature_labels,
    file_name = paste0("barplot_Olink_AUC_compare_AllDatasets_", phenotype_type, ".pdf"),
    width = 4.5 * 1.2,
    height = 3.3 * 1.2
  )
  
}
```


# Figure 3I,J. Odds ratio barplots.
```{r}

external_datasets <- list(
  Lozano = read_cohort("Lozano_save.csv"),
  Gerber = read_cohort("Gerber_save.csv"),
  Switzerland1 = read_cohort("Switzerland1_save.csv"),
  Switzerland2 = read_cohort("Switzerland2_save.csv"),
  Switzerland3 = read_cohort("Switzerland3_save.csv")
)

external_datasets <- lapply(external_datasets, ensure_feats, feats = all_feats)
external_datasets <- lapply(external_datasets, make_severe_irAE, out_col = pheno_feat)

for (dataset_name in names(external_datasets)) {
  plot_data <- prepare_binary_plot_data(external_datasets[[dataset_name]], main_feature)
  or_value <- save_irAE_barplot2(
    plot_data,
    paste0("barplot_", dataset_name, "_OddsRatio.pdf"),
    width = 1.5,
    height = 1.85,
    phenotype_type = ""
  )
  print(paste(dataset_name, "OR:", or_value))
}

## OR for merged data
merged_data <- lapply(names(external_datasets), function(dataset_name) {
  plot_data <- prepare_binary_plot_data(
    external_datasets[[dataset_name]],
    main_feature
  )
  plot_data$Cohort <- dataset_name
  plot_data
}) |>
  dplyr::bind_rows()

or_value <- save_irAE_barplot2(
  merged_data,
  paste0("barplot_5merged_OddsRatio_aCD8Tcm.pdf"),
  width = 1.5,
  height = 1.85,
  phenotype_type = "" # conventional
)
print(paste(dataset_name, "OR:", or_value))

```


# Figure 3K. Forest plot of OS hazard ratios across Switzerland cohorts
```{r}
feature_col <- "CD8T_CM_tripletORp_ratio_pre"

format_p_value <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p >= 0.05 ~ as.character(round(p, 2)),
    p >= 0.0095 ~ as.character(round(p, 3)),
    TRUE ~ format(p, scientific = TRUE, digits = 2)
  )
}

fit_cox_by_score <- function(df, dataset_name, feature_col) {
  model_df <- df %>%
    select(all_of(c(feature_col, "OS_time", "OS_event"))) %>%
    mutate(
      OS_time = suppressWarnings(as.numeric(OS_time)),
      OS_event = suppressWarnings(as.numeric(OS_event)),
      Score = factor(
        ifelse(.data[[feature_col]] >= quantile(.data[[feature_col]], 0.5, na.rm = TRUE), "High", "Low"), 
        levels = c("Low", "High")
      )
    ) %>%
    drop_na()

  if (nrow(model_df) < 5 || length(unique(model_df$Score)) < 2 || length(unique(model_df$OS_event)) < 2) {
    return(data.frame(
      dataset = dataset_name,
      HR = NA_real_,
      HR_low = NA_real_,
      HR_up = NA_real_,
      pval = NA_real_
    ))
  }

  cox_fit <- coxph(Surv(OS_time, OS_event) ~ Score, data = model_df)
  cox_sum <- summary(cox_fit)

  data.frame(
    dataset = dataset_name,
    HR = cox_sum$coefficients[1, "exp(coef)"],
    HR_low = cox_sum$conf.int[1, "lower .95"],
    HR_up = cox_sum$conf.int[1, "upper .95"],
    pval = cox_sum$coefficients[1, "Pr(>|z|)"]
  )
}

switzerland_datasets <- list(
  Switzerland1 = read_cohort("Switzerland1_save.csv"),
  Switzerland2 = read_cohort("Switzerland2_save.csv"),
  Switzerland3 = read_cohort("Switzerland3_save.csv")
)

result_df <- bind_rows(
  lapply(names(switzerland_datasets), function(dataset_name) {
    fit_cox_by_score(switzerland_datasets[[dataset_name]], dataset_name, feature_col)
  })
)

merged_df <- bind_rows(switzerland_datasets)
result_df <- bind_rows(
  result_df,
  fit_cox_by_score(merged_df, "Merged", feature_col)
)

plot_data <- result_df %>%
  filter(!is.na(HR), !is.na(HR_low), !is.na(HR_up)) %>%
  mutate(
    mean = HR,
    lower = HR_low,
    upper = HR_up,
    Variable = dataset,
    effSize = sprintf("%.3f", HR),
    pval = format_p_value(pval)
  )

forest_plot <- forestplot(
  plot_data,
  labeltext = c(effSize, Variable, pval),
  mean = mean,
  lower = lower,
  upper = upper,
  graph.pos = 3,
  boxsize = 0.25,
  vertices = TRUE,
  clip = c(0, 4),
  xlog = FALSE,
  zero = 1,
  xlab = "HR of OS",
  xticks = c(0, 1, 2, 3),
  graphwidth = unit(2.5, "cm"),
  lineheight = unit(0.65, "cm"),
  txt_gp = fpTxtGp(
    ticks = gpar(cex = 1.1),
    xlab = gpar(cex = 1.1),
    label = gpar(cex = 1.1),
    title = gpar(cex = 1.1)
  )
) %>%
  fp_set_style(
    box = "black",
    line = "black",
    align = "llr",
    summary = "black"
  ) %>%
  fp_add_header(
    effSize = "HR",
    Variable = "Cohort",
    pval = "P-value"
  )

pdf(
  file.path(result_figure_dir, "Forest_Switzerland_cohorts_actCD8Tcm_OS.pdf"),
  onefile = FALSE,
  width = 5.5,
  height = 3.2
)
print(forest_plot)
dev.off()

result_df

```


# Figure 5B,D. Feature importance and Odds ratio bar plots of the LR model.
```{r}

HCI002_data <- datasets$HCI002
HCI001_data <- datasets$HCI001
Nicolas_FACS_data <- datasets$Nicolas_FACS
Nicolas_Cytek_data <- datasets$Nicolas_Cytek
Nicolas_CyTOF_data <- datasets$Nicolas_CyTOF

HCI002_data$severe_irAE <- HCI002_data$IRAEactionable01
HCI001_data$severe_irAE <- HCI001_data$IRAEactionable01

model_datasets <- list(
  HCI002 = HCI002_data,
  HCI001 = HCI001_data,
  Nicolas_FACS = Nicolas_FACS_data,
  Nicolas_Cytek = Nicolas_Cytek_data,
  Nicolas_CyTOF = Nicolas_CyTOF_data
)

model_datasets <- lapply(model_datasets, zscore_feats, feats = sig_feats)

HCI002_data <- model_datasets$HCI002
HCI001_data <- model_datasets$HCI001
Nicolas_FACS_data <- model_datasets$Nicolas_FACS
Nicolas_Cytek_data <- model_datasets$Nicolas_Cytek
Nicolas_CyTOF_data <- model_datasets$Nicolas_CyTOF

logistic_model <- fit_logistic(HCI002_data, y_col = pheno_feat, x_cols = sig_feats)

a <- HCI002_data[, c(pheno_feat, sig_feats)]
a <- na.omit(a)
dim(a)
sum(a$severe_irAE)

coef_values <- coef(logistic_model)[-1]

feature_importance <- data.frame(
  Feature = c("CXCL9", "Activated CD8Tcm"),
  Coefficient = c(coef_values[["CXCL9_pre"]], coef_values[["CD8T_CM_tripletORp_ratio_pre"]])
) %>%
  mutate(
    Importance = abs(Coefficient),
    Direction = ifelse(Coefficient >= 0, "Positive", "Negative")
  ) %>%
  arrange(desc(Importance))

p <- ggplot(feature_importance, aes(x = reorder(Feature, Importance), y = Importance, fill = Direction)) +
  geom_bar(stat = "identity", color = "black", width = 0.6) +
  scale_fill_manual(values = c(Negative = "#d62728", Positive = "#1f77b4")) +
  geom_text(aes(label = sprintf("%.2f", Coefficient), y = Importance + 0.04), size = 4) +
  ylim(c(0, 0.8)) +
  labs(x = NULL, y = "Feature importance", fill = NULL) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

pdf(file.path(result_figure_dir, "barplot_featureImportance_IntegratedML_irAE_HCI002.pdf"), onefile = FALSE, width = 3.5 * 0.8, height = 4.5 * 0.8)
print(p)
dev.off()

model_datasets <- list(
  HCI002 = HCI002_data,
  HCI001 = HCI001_data,
  Nicolas_FACS = Nicolas_FACS_data,
  Nicolas_CyTOF = Nicolas_CyTOF_data,
  Nicolas_Cytek = Nicolas_Cytek_data
)
model_datasets <- lapply(model_datasets, add_model_prediction, y_col = pheno_feat, x_cols = sig_feats, model = logistic_model)
phenotype_types5 <- c("actionable", "actionable", "conventional", "conventional", "conventional")

HCI002_data <- model_datasets$HCI002
HCI001_data <- model_datasets$HCI001
Nicolas_FACS_data <- model_datasets$Nicolas_FACS
Nicolas_CyTOF_data <- model_datasets$Nicolas_CyTOF
Nicolas_Cytek_data <- model_datasets$Nicolas_Cytek

for (i in 1:5) {
  dataset_name <- names(model_datasets)[i]
  pt <- phenotype_types5[i]
  plot_data <- prepare_binary_plot_data(model_datasets[[dataset_name]], "integrated")

  save_irAE_barplot2(
    plot_data,
    paste0("barplot_", dataset_name, "_ML.pdf"),
    width = 1.5 * 0.95 * 1.1,
    height = 2.7 * 0.95 * 0.8 * 1.1,
    phenotype_type = ""
  )
}

## OR for merged data
merged_data <- lapply(names(model_datasets), function(dataset_name) {
  plot_data <- prepare_binary_plot_data(
    model_datasets[[dataset_name]],
    "integrated"
  )
  plot_data$Cohort <- dataset_name
  plot_data
}) |>
  dplyr::bind_rows()

or_value <- save_irAE_barplot2(
  merged_data,
  paste0("barplot_5merged_OddsRatio_ML.pdf"),
  width = 1.5 * 0.95 * 1.1,
  height = 2.7 * 0.95 * 0.8 * 1.1,
  phenotype_type = "" # conventional
)
print(paste(dataset_name, "OR:", or_value))


```

# SFigure 11, STable 10
```{r}

phenotype_type <- "actionable"

HCI002_data <- set_irAE_outcome(HCI002_data, phenotype_type)
HCI001_data <- set_irAE_outcome(HCI001_data, phenotype_type)

model_datasets <- list(
  HCI002 = HCI002_data,
  HCI001 = HCI001_data,
  Nunez1_FACS = Nicolas_FACS_data,
  Nunez2_Cytek = Nicolas_Cytek_data
)

dataset_labels <- c(
  HCI002 = "HCI002",
  HCI001 = "HCI001",
  Nunez1_FACS = "Nunez1",
  Nunez2_Cytek = "Nunez2"
)

auc_df <- lapply(names(model_datasets), function(dataset_name) {

  df_roc <- model_datasets[[dataset_name]] %>%
    filter(
      !is.na(severe_irAE),
      !is.na(CXCL9_pre)
    )

  roc_obj <- roc(
    response  = df_roc$severe_irAE,
    predictor = df_roc$CXCL9_pre,
    levels    = c(0, 1),
    direction = "<",
    quiet = TRUE
  )

  auc_ci <- ci.auc(
    roc_obj,
    method = "delong"
  )

  data.frame(
    dataset = dataset_name,
    label = dataset_labels[dataset_name],
    n = nrow(df_roc),
    AUC = as.numeric(auc(roc_obj)),
    CI_low = as.numeric(auc_ci[1]),
    CI_high = as.numeric(auc_ci[3])
  )
}) %>%
  bind_rows()


auc_df$label <- factor(
  auc_df$label,
  levels = unname(dataset_labels)
)

auc_df

auc_df$label <- factor(
  auc_df$label,
  levels = rev(c(
    "HCI002",
    "HCI001",
    "Nunez1",
    "Nunez2"
  ))
)

p_auc <- ggplot(
  auc_df,
  aes(x = AUC, y = label)
) +
  geom_vline(
    xintercept = 0.5,
    linetype = "dashed",
    linewidth = 0.6,
    color = "grey50"
  ) +
  geom_errorbarh(
    aes(
      xmin = CI_low,
      xmax = CI_high
    ),
    height = 0.15,
    linewidth = 0.8
  ) +
  geom_point(
    size = 3
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2)
  ) +
  labs(
    x = "AUC (95% CI)",
    y = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 13),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank()
  )

save_pdf(print(p_auc), file.path(result_figure_dir, paste0("AUC_4datasets_CXCL9_pre.pdf")), width = 2 * 0.6 * 3, height = 2 * 0.4 * 3)


########### train LR model with different datasets
HCI002_data_temp <- datasets$HCI002
HCI001_data_temp <- datasets$HCI001
Nicolas_FACS_data_temp <- datasets$Nicolas_FACS
Nicolas_Cytek_data_temp <- datasets$Nicolas_Cytek
Nicolas_CyTOF_data_temp <- datasets$Nicolas_CyTOF

HCI002_data_temp$severe_irAE <- HCI002_data_temp$IRAEactionable01
HCI001_data_temp$severe_irAE <- HCI001_data_temp$IRAEactionable01

model_datasets <- list(
  HCI002 = HCI002_data_temp,
  HCI001 = HCI001_data_temp,
  Nicolas_FACS = Nicolas_FACS_data_temp,
  Nicolas_Cytek = Nicolas_Cytek_data_temp,
  Nicolas_CyTOF = Nicolas_CyTOF_data_temp
)

model_datasets <- lapply(model_datasets, zscore_feats, feats = sig_feats)

HCI002_data_temp <- model_datasets$HCI002
HCI001_data_temp <- model_datasets$HCI001
Nicolas_FACS_data_temp <- model_datasets$Nicolas_FACS
Nicolas_Cytek_data_temp <- model_datasets$Nicolas_Cytek
Nicolas_CyTOF_data_temp <- model_datasets$Nicolas_CyTOF

logistic_model <- fit_logistic(Nicolas_CyTOF_data_temp, y_col = pheno_feat, x_cols = sig_feats)

coef_values <- coef(logistic_model)[-1]

```

# SFigure 17, STable 11
```{r}

HCI002_info <- read_cohort("HCI002_info_save.csv")
data_selected1 <- HCI002_info[, c("patient_therapy", "Therapy_type", "Response01", "TMB (mutation /mb)")] 
data_selected2 <- HCI002_data[, c("patient_therapy", "integrated", "CD8T_CM_tripletORp_ratio_pre", "CXCL9_FC", "irAE_grade", "IRAEactionable01")] 
data_selected2 <- make_severe_irAE(data_selected2, grade_col = "irAE_grade", out_col = "severe_irAE")
data_selected_HCI002 <- merge(data_selected1, data_selected2, by = "patient_therapy")
colnames(data_selected_HCI002)[colnames(data_selected_HCI002) == "TMB (mutation /mb)"] <- "TMB"

data_selected <- data_selected_HCI002

df <- data_selected %>%
  select(Response01, IRAEactionable01, TMB, integrated) %>%
  na.omit()
## ROC curves with Youden index
HCI002_actionable_plot <- df[, c("integrated", "IRAEactionable01")]
colnames(HCI002_actionable_plot) <- c("feature", "phenotype")
save_roc_plot_with_youden(
  HCI002_actionable_plot,
  "ROC_HCI002_actionable_irAE_Youden_cutoff_integrated.pdf"
)
HCI002_actionable_plot <- df[, c("TMB", "Response01")]
colnames(HCI002_actionable_plot) <- c("feature", "phenotype")
save_roc_plot_with_youden(
  HCI002_actionable_plot,
  "ROC_HCI002_Response01_Youden_cutoff_TMB.pdf"
)

# Youden cutoffs
roc_tmb <- roc(df$Response01, df$TMB,
               levels = c(0, 1), direction = "<", quiet = TRUE)

roc_irae <- roc(df$IRAEactionable01, df$integrated,
                levels = c(0, 1), direction = "<", quiet = TRUE)

cut_tmb <- as.numeric(coords(
  roc_tmb, "best", best.method = "youden", ret = "threshold"
)[[1]][1])

cut_irae <- as.numeric(coords(
  roc_irae, "best", best.method = "youden", ret = "threshold"
)[[1]][1])

df <- df %>%
  na.omit()

# Four regions
df <- df %>%
  mutate(
    TMB_group = ifelse(TMB >= cut_tmb, "TMB-high", "TMB-low"),
    irAE_group = ifelse(integrated >= cut_irae, "irAE-high", "irAE-low"),
    region = paste(TMB_group, irAE_group, sep = " / ")
  )

# Response and irAE rates in each region
region_summary <- df %>%
  group_by(region) %>%
  summarise(
    n = n(),
    response_n = sum(Response01 == 1),
    response_rate = mean(Response01 == 1),
    irAE_n = sum(IRAEactionable01 == 1),
    irAE_rate = mean(IRAEactionable01 == 1),
    .groups = "drop"
  )

region_summary

```

# SFigure 14C (left) (phenotype_type = "conventional"). ROC plots, box plots, bar plots.
```{r}
main_feature <- "integrated"

for (phenotype_type in phenotype_types) {
  HCI002_data <- set_irAE_outcome(HCI002_data, phenotype_type)
  HCI001_data <- set_irAE_outcome(HCI001_data, phenotype_type)

  hci_datasets <- list(HCI001 = HCI001_data, HCI002 = HCI002_data)

  for (dataset_name in names(hci_datasets)) {
    plot_data <- prepare_binary_plot_data(hci_datasets[[dataset_name]], main_feature)

    save_irAE_barplot2(
      plot_data = plot_data,
      paste0("barplot_", dataset_name, "_", main_feature, "_", phenotype_type, ".pdf"),
      width = 1.5 * 1.05,
      phenotype_type = phenotype_type
    )
    
    plot_data_temp <- plot_data
    plot_data_temp$feature <- plot_data_temp$feature/100
    save_feature_boxplot(
      plot_data_temp,
      paste0("boxplot_", dataset_name, "_",  main_feature, "_", phenotype_type, ".pdf"),
      ylabel = "Integrated score", ylim = 0.9
    )

    save_roc_plot(
      plot_data,
      paste0("ROC_", dataset_name, "_",  main_feature, "_", phenotype_type, ".pdf")
    )
  }
}

```

# SFigure 16C. Performance on HCI001/HCI002 with a fixed cutoff of integrated_pre
```{r}

main_feature <- "integrated"

HCI002_actionable <- set_irAE_outcome(
  HCI002_data,
  "actionable"
)

HCI002_actionable_plot <- prepare_binary_plot_data(
  HCI002_actionable,
  feature_col = main_feature
)

HCI002_youden <- get_youden_cutoff(
  HCI002_actionable_plot
)

cutoff_youden <- HCI002_youden$threshold

save_roc_plot_with_youden(
  HCI002_actionable_plot,
  "ROC_HCI002_actionable_integrated_Youden_cutoff.pdf"
)

datasets_cutoff <- list(
  HCI001 = HCI001_data,
  HCI002 = HCI002_data
)

for (phenotype_type in phenotype_types) {
  for (dataset_name in names(datasets_cutoff)) {
    df <- set_irAE_outcome(
      datasets_cutoff[[dataset_name]],
      phenotype_type
    )
    plot_data <- prepare_binary_plot_data(
      df,
      feature_col = main_feature,
      cutoff = cutoff_youden
    )
    cat(
      dataset_name,
      phenotype_type,
      "cutoff =",
      attr(plot_data, "cutoff"),
      "\n"
    )
    save_irAE_barplot2(
      plot_data,
      paste0(
        "barplot_",
        dataset_name,
        "_",main_feature,"_",
        phenotype_type,
        "_Youden_cutoff.pdf"
      ),
      width = 1.5 * 1.2,
      height = 2.7 * 0.95,
      phenotype_type = phenotype_type
    )
  }
}

```


# Figure 5C (phenotype_type = "actionable") and SFigure 14C (right; phenotype_type = "conventional") AUC comparison between Integrated ML vs other signatures (across 5 datasets). SFigure 15C.
```{r}

HCI002_data_temp <- datasets$HCI002
HCI001_data_temp <- datasets$HCI001
Nicolas_FACS_data_temp <- datasets$Nicolas_FACS
Nicolas_Cytek_data_temp <- datasets$Nicolas_Cytek
Nicolas_CyTOF_data_temp <- datasets$Nicolas_CyTOF

HCI002_data_temp$severe_irAE <- HCI002_data_temp$IRAEactionable01
HCI001_data_temp$severe_irAE <- HCI001_data_temp$IRAEactionable01

model_datasets <- list(
  HCI002 = HCI002_data_temp,
  HCI001 = HCI001_data_temp,
  Nicolas_FACS = Nicolas_FACS_data_temp,
  Nicolas_Cytek = Nicolas_Cytek_data_temp,
  Nicolas_CyTOF = Nicolas_CyTOF_data_temp
)

model_datasets <- lapply(model_datasets, zscore_feats, feats = sig_feats)

HCI002_data_temp <- model_datasets$HCI002
HCI001_data_temp <- model_datasets$HCI001
Nicolas_FACS_data_temp <- model_datasets$Nicolas_FACS
Nicolas_Cytek_data_temp <- model_datasets$Nicolas_Cytek
Nicolas_CyTOF_data_temp <- model_datasets$Nicolas_CyTOF

logistic_model <- fit_logistic(HCI002_data, y_col = pheno_feat, x_cols = sig_feats)


for (phenotype_type in c('actionable', 'conventional', 'any', 'drop_mild')) {
  HCI002_temp <- set_irAE_outcome(HCI002_data, phenotype_type)
  HCI001_temp <- set_irAE_outcome(HCI001_data, phenotype_type)
  
  model_datasets <- list(
    HCI002 = HCI002_temp,
    HCI001 = HCI001_temp,
    Nicolas_FACS = Nicolas_FACS_data,
    Nicolas_Cytek = Nicolas_Cytek_data,
    Nicolas_CyTOF = Nicolas_CyTOF_data
  )
  
  auc_df <- matrix(
    NA_real_,
    nrow = length(model_datasets),
    ncol = length(c("MODEL_sig_feats", all_feats)),
    dimnames = list(names(model_datasets), c("MODEL_sig_feats", all_feats))
  ) %>%
    as.data.frame()
  
  for (dataset_name in names(model_datasets)) {
    df <- model_datasets[[dataset_name]]
    auc_df[dataset_name, "MODEL_sig_feats"] <- predict_auc(
      logistic_model,
      df,
      y_col = pheno_feat,
      x_cols = sig_feats
    )$auc
  
    for (feat in all_feats) {
      auc_df[dataset_name, feat] <- compute_single_feature_auc(df, pheno_feat, feat)
    }
  }
  
  ml_features <- c(
    "MODEL_sig_feats",
    "CD8T_CM_tripletORp_ratio_pre",
    "CXCL9_pre",
    "CD8T_KI67p_ratio_pre",
    "CD4T_KI67p_ratio_pre",
    "Treg_KI67p_ratio_pre",
    "CD4TM_tripletORp_ratio_pre",
    "CD8T_EMRA_CD38p_ratio_pre",
    "CD4T_EM_tripletORp_ratio_pre",
    "CD4T_EM_ratio_pre",
    "CD4T_Naive_ratio_pre",
    "CXCL10_pre",
    "CXCL11_pre",
    "IFNG_pre"
  )
  
  ml_labels <- c(
    "Integrated",
    "Activated CD8Tcm",
    "CXCL9",
    "Ki67+ CD8T",
    "Ki67+ CD4T",
    "Ki67+ Treg",
    "Activated CD4Tm",
    "CD38+ CD8Temra",
    "Activated CD4Tem",
    "CD4Tem",
    "Naive CD4T",
    "CXCL10",
    "CXCL11",
    "IFNG"
  )
  
  save_auc_comparison(
    auc_df = auc_df,
    selected_features = ml_features,
    labels = ml_labels,
    file_name = paste0("barplot_AUC_compare_AllDatasets_ML_", phenotype_type, ".pdf"),
    width = 4.5 * 1.35 * 0.8,
    height = 3.3 * 1.2 * 0.8
  )
  
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD8T_CM_tripletORp_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD4T_Naive_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD8T_EMRA_CD38p_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD8T_KI67p_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD4TM_tripletORp_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD4T_EM_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$IFNG_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD4T_KI67p_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$Treg_KI67p_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CD4T_EM_tripletORp_ratio_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CXCL11_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CXCL9_pre[1:5], paired = T)
  t.test(auc_df$MODEL_sig_feats[1:5], auc_df$CXCL10_pre[1:5], paired = T)
}

```


# Figure 5E and SFigure 12F. Survival analysis.
```{r}
make_survival_df <- function(df, score_col = "integrated", high_rule = c(">=", ">")) {
  high_rule <- match.arg(high_rule)

  # Add missing PFS columns as NA
  if (!"PFS_time" %in% colnames(df)) df$PFS_time <- NA_real_
  if (!"PFS_event" %in% colnames(df)) df$PFS_event <- NA_real_

  out <- df[, c(score_col, "OS_time", "OS_event", "PFS_time", "PFS_event"), drop = FALSE]
  colnames(out) <- c("integrated", "OS_time", "OS_event", "PFS_time", "PFS_event")

  med <- median(out$integrated, na.rm = TRUE)
  ok <- !is.na(out$integrated)

  out$score <- NA_character_
  if (high_rule == ">=") {
    out$score[ok & out$integrated >= med] <- "High"
    out$score[ok & out$integrated < med] <- "Low"
  } else {
    out$score[ok & out$integrated > med] <- "High"
    out$score[ok & out$integrated <= med] <- "Low"
  }

  out$score <- factor(out$score, levels = c("Low", "High"))
  out
}

survival_datasets <- list(
  HCI002 = make_survival_df(HCI002_data, high_rule = ">"),
  HCI001 = make_survival_df(HCI001_data, high_rule = ">"),
  Nicolas_FACS = make_survival_df(Nicolas_FACS_data, high_rule = ">"),
  Nicolas_CyTOF = make_survival_df(Nicolas_CyTOF_data, high_rule = ">"),
  Nicolas_Cytek = make_survival_df(Nicolas_Cytek_data, high_rule = ">")
)

survival_labels <- c(
  OS = "Overall \nsurvival probability",
  PFS = "Progression-free \nsurvival probability"
)

for (cohort_type in c("Pooled", "HCI002")) {
  survival_df <- if (cohort_type == "Pooled") {
    bind_rows(
      transform(survival_datasets$HCI002, cohort = "HCI002"),
      transform(survival_datasets$HCI001, cohort = "HCI001"),
      transform(survival_datasets$Nicolas_FACS, cohort = "Nicolas_FACS"),
      transform(survival_datasets$Nicolas_Cytek, cohort = "Nicolas_Cytek")
    )
  } else {
    transform(survival_datasets$HCI002, cohort = "HCI002")
  }

  for (phenotype in names(survival_labels)) {
    save_km_plot2(
      survival_df,
      phenotype = phenotype,
      predictor_name = "Integrated",
      cohort_type = cohort_type,
      y_label = survival_labels[[phenotype]]
    )
  }
}

```


# Figure 5F. Multivariable Cox regression forest plots for OS and PFS in HCI002
```{r}
test_var <- "integrated"
test_var_label <- "Integrated"

control_vars <- c("Age", "Sex", "Stage", "Drug", "prior_ICI", "IRAEactionable01")

variable_labels <- c(
  test_var_label,
  "Age",
  "Sex (Female vs male)",
  "Stage (IV vs I-III)",
  "Drug (anti-PD(L)1 vs\n anti-CTLA4-containing)",
  "Previous ICB therapy",
  "Severe irAE"
)

format_p_value <- function(p) {
  dplyr::case_when(
    is.na(p) ~ NA_character_,
    p >= 0.05 ~ as.character(round(p, 2)),
    p >= 0.0095 ~ as.character(round(p, 3)),
    TRUE ~ format(p, scientific = TRUE, digits = 2)
  )
}

fit_multivariable_cox <- function(df, test_var, control_vars, outcome_prefix) {
  model_df <- df %>%
    select(
      all_of(c(test_var, control_vars)),
      time = all_of(paste0(outcome_prefix, "_time")),
      event = all_of(paste0(outcome_prefix, "_event"))
    ) %>%
    mutate(
      Score = factor(
        ifelse(.data[[test_var]] > median(.data[[test_var]], na.rm = TRUE), "High", "Low"),
        levels = c("Low", "High")
      )
    ) %>%
    select(Score, all_of(control_vars), time, event) %>%
    drop_na()

  cox_fit <- coxph(
    as.formula(paste("Surv(time, event) ~ Score +", paste(control_vars, collapse = " + "))),
    data = model_df
  )

  cox_sum <- summary(cox_fit)

  tibble(
    Variable = variable_labels,
    HR = cox_sum$coefficients[, "exp(coef)"],
    HR_low = cox_sum$conf.int[, "lower .95"],
    HR_up = cox_sum$conf.int[, "upper .95"],
    pval = cox_sum$coefficients[, "Pr(>|z|)"]
  )
}

save_forest_plot <- function(plot_data, file_name, x_label) {
  plot_data <- plot_data %>%
    mutate(
      effSize = sprintf("%.3f", HR),
      pval = format_p_value(pval)
    )

  forest_plot <- forestplot(
    plot_data,
    labeltext = c(effSize, Variable, pval),
    mean = HR,
    lower = HR_low,
    upper = HR_up,
    graph.pos = 3,
    boxsize = 0.25,
    vertices = TRUE,
    clip = c(0, 4),
    xlog = FALSE,
    zero = 1,
    txt_gp = fpTxtGp(
      ticks = gpar(cex = 1.1),
      xlab = gpar(cex = 1.1),
      label = gpar(cex = 1.1),
      title = gpar(cex = 1.1)
    ),
    xlab = x_label,
    xticks = c(0, 1, 2, 3),
    graphwidth = unit(2.5, "cm"),
    lineheight = unit(0.8, "cm")
  ) %>%
    fp_set_style(
      box = "black",
      line = "black",
      align = "llr",
      summary = "black"
    ) %>%
    fp_add_header(
      effSize = "HR",
      Variable = "Variable",
      pval = "P-value"
    )

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = 5.5, height = 3.4)
  print(forest_plot)
  dev.off()
}

os_results <- fit_multivariable_cox(HCI002_data, test_var, control_vars, "OS")
pfs_results <- fit_multivariable_cox(HCI002_data, test_var, control_vars, "PFS")

result_df <- os_results %>%
  select(Variable, HR_OS = HR, HR_OS_low = HR_low, HR_OS_up = HR_up, p_val_OS = pval) %>%
  left_join(
    pfs_results %>%
      select(Variable, HR_PFS = HR, HR_PFS_low = HR_low, HR_PFS_up = HR_up, p_val_PFS = pval),
    by = "Variable"
  )

save_forest_plot(
  os_results,
  "Forest_HCI002_IntegratedML_clinicalVARs_MHR_OS.pdf",
  "HR of OS"
)

save_forest_plot(
  pfs_results,
  "Forest_HCI002_IntegratedML_clinicalVARs_MHR_PFS.pdf",
  "HR of PFS"
)

result_df
```

# SFigure 9A. Pooled analysis (Odds ratio barplot, K-M curves). 
```{r}

main_feature <- "CD8T_CM_tripletORp_ratio_pre" 

clinical_feats <- c(
  "CD8T_CM_tripletORp_ratio_pre",
  "CXCL9_pre",
  "CXCL9_FC",
  "irAE_grade",
  "prior_BRAFi_MEKi",
  "prior_ICI",
  "Drug",
  "Sex",
  "Age",
  "Stage",
  "OS_time",
  "OS_event",
  "PFS_time",
  "PFS_event"
)

pooled_datasets <- list(
  HCI002 = read_cohort("HCI002_info_save.csv"),
  HCI001 = read_cohort("HCI001_info_save.csv"),
  Nicolas_FACS = read_cohort("Nicolas_all_info_FACS_save.csv", melanoma_only = TRUE),
  Nicolas_Cytek = read_cohort("Nicolas_all_info_Cytek_save.csv", melanoma_only = TRUE),
  Nicolas_CyTOF = read_cohort("Nicolas_all_info_CyTOF_save.csv", melanoma_only = TRUE),
  Lozano = read_cohort("Lozano_save.csv"),
  Gerber = read_cohort("Gerber_save.csv"),
  Switzerland1 = read_cohort("Switzerland1_save.csv"),
  Switzerland2 = read_cohort("Switzerland2_save.csv"),
  Switzerland3 = read_cohort("Switzerland3_save.csv")
)

pooled_datasets <- lapply(pooled_datasets, ensure_feats, feats = clinical_feats)
pooled_datasets <- lapply(pooled_datasets, function(df) df[, clinical_feats])
pooled_datasets <- lapply(pooled_datasets, make_severe_irAE, out_col = pheno_feat)

pooled_datasets <- lapply(pooled_datasets, make_binary_score, score_col = main_feature, high_rule = ">")

standardize_clinical_types <- function(df) {
  df %>%
    mutate(
      prior_BRAFi_MEKi = as.character(prior_BRAFi_MEKi),
      prior_ICI = as.character(prior_ICI),
      Drug = as.character(Drug),
      Sex = as.character(Sex),
      Stage = as.character(Stage),
      irAE_grade = suppressWarnings(as.numeric(irAE_grade)),
      severe_irAE = suppressWarnings(as.numeric(severe_irAE)),
      OS_time = suppressWarnings(as.numeric(OS_time)),
      OS_event = suppressWarnings(as.numeric(OS_event)),
      PFS_time = suppressWarnings(as.numeric(PFS_time)),
      PFS_event = suppressWarnings(as.numeric(PFS_event))
    )
}

pooled_datasets <- lapply(pooled_datasets, standardize_clinical_types)

all_merged_survival <- bind_rows(
  transform(pooled_datasets$HCI002, cohort = "HCI002"),
  transform(pooled_datasets$HCI001, cohort = "HCI001"),
  transform(pooled_datasets$Nicolas_FACS, cohort = "Nicolas_FACS"),
  transform(pooled_datasets$Nicolas_Cytek, cohort = "Nicolas_Cytek"),
  transform(pooled_datasets$Switzerland1, cohort = "Switzerland1"),
  transform(pooled_datasets$Switzerland2, cohort = "Switzerland2"),
  transform(pooled_datasets$Switzerland3, cohort = "Switzerland3")
)

all_merged_survival <- all_merged_survival %>%
  mutate(
    Sex = recode(Sex, M = "Male", F = "Female"),
    Stage = ifelse(!is.na(Stage) & Stage != "4", "1-3", Stage),
    prior_ICI = recode(as.character(prior_ICI), `0` = "no", `1` = "yes")
  )

km_input <- all_merged_survival %>%
  select(score = all_of(main_feature), OS_time, OS_event, PFS_time, PFS_event)

save_km_plot2(
  km_input,
  phenotype = "OS",
  predictor_name = main_feature,
  cohort_type = "pooled",
  y_label = "Overall \nsurvival probability"
)

plot_data <- all_merged_survival %>%
  select(group = all_of(main_feature), phenotype = severe_irAE) %>%
  drop_na() %>%
  mutate(group = factor(group, levels = c("Low", "High")))

save_irAE_barplot(plot_data, paste0("barplot_pooled.pdf"), width = 1.5, height = 1.85, phenotype_type = "")

plot_data <- all_merged_survival %>%
  select(group = all_of(main_feature), phenotype = severe_irAE) %>%
  drop_na() %>%
  mutate(group = factor(group, levels = c("Low", "High")))

save_irAE_barplot2(plot_data, paste0("barplot_pooled.pdf"), width = 1.2*1.2, height = 1.85*1.2, phenotype_type = "")

```

# SFigure 13A,B. revision for Pooled analysis (Odds ratio barplot)  for CXCL9 and aCD8Tcm, PD1 vs CTLA4 subgroup analysis
```{r}
########## for revision  (odds ratio pooled analysis)
drug <- "CTLA4-containing" # "CTLA4-containing" "PD1/PDL1"
drug_name <- "CTLA4" # "CTLA4" "PD1"

####for aCD8T_CM
filtered_data <- all_merged_survival # [all_merged_survival$cohort == "HCI002", ] # all or HCI002 only
filtered_data <- filtered_data[filtered_data$Drug == drug, c("CD8T_CM_tripletORp_ratio_pre", "severe_irAE")] # "CTLA4-containing" "PD1/PDL1"
filtered_data <- na.omit(filtered_data)
dim(filtered_data)
table(filtered_data$severe_irAE)
plot_data <- filtered_data %>%
  select(group = "CD8T_CM_tripletORp_ratio_pre", phenotype = severe_irAE) %>%
  drop_na() %>%
  mutate(group = factor(group, levels = c("Low", "High")))
save_irAE_barplot2(plot_data, paste0("barplot_pooled_", drug_name, "_aCD8Tcm.pdf"), width = 1.5, height = 1.85, phenotype_type = "")


########## for CXCL9
filtered_data <- all_merged_survival 
filtered_data <- filtered_data[filtered_data$Drug == drug, c("CXCL9_FC", "severe_irAE")] # "CTLA4-containing" "PD1/PDL1"
filtered_data <- na.omit(filtered_data)
dim(filtered_data)
table(filtered_data$severe_irAE)
plot_data <- filtered_data %>%
  select(group = "CXCL9_FC", phenotype = severe_irAE) %>%
  drop_na() %>%
  mutate(
    group = ifelse(group > median(group), "High", "Low"),
    group = factor(group, levels = c("Low", "High"))
  )
save_irAE_barplot2(plot_data, paste0("barplot_pooled_", drug_name, "_CXCL9FC.pdf"), width = 1.5, height = 1.85, phenotype_type = "")

```


# SFigure 13C. pooled analaysis of integrated sig. include PD1 vs CTLA4 subgroup analysis
```{r}

main_feature <- "integrated"
pooled_datasets_temp <- list(
  HCI002 = HCI002_data[, c("integrated", "irAE_grade", "Drug")],
  HCI001 = HCI001_data[, c("integrated", "irAE_grade", "Drug")],
  Nicolas_FACS = Nicolas_FACS_data[, c("integrated", "irAE_grade", "Drug")],
  Nicolas_Cytek = Nicolas_Cytek_data[, c("integrated", "irAE_grade", "Drug")],
  Nicolas_CyTOF = Nicolas_CyTOF_data[, c("integrated", "irAE_grade", "Drug")]
)

pooled_datasets_temp <- lapply(pooled_datasets_temp, make_severe_irAE, out_col = pheno_feat)
pooled_datasets_temp <- lapply(pooled_datasets_temp, make_binary_score, score_col = main_feature, high_rule = ">")
all_merged_temp <- bind_rows(
  transform(pooled_datasets_temp$HCI002, cohort = "HCI002"),
  transform(pooled_datasets_temp$HCI001, cohort = "HCI001"),
  transform(pooled_datasets_temp$Nicolas_FACS, cohort = "Nicolas_FACS"),
  transform(pooled_datasets_temp$Nicolas_Cytek, cohort = "Nicolas_Cytek")
)

drug <- "CTLA4-containing" # "CTLA4-containing" "PD1/PDL1"
drug_name <- "CTLA4" # "CTLA4" "PD1"
filtered_data <- all_merged_temp 
filtered_data <- filtered_data[filtered_data$Drug == drug, c("integrated", "severe_irAE")] 
filtered_data <- na.omit(filtered_data)
dim(filtered_data)
table(filtered_data$severe_irAE)
plot_data <- filtered_data %>%
  select(group = all_of(main_feature), phenotype = severe_irAE) %>%
  drop_na() %>%
  mutate(group = factor(group, levels = c("Low", "High")))
save_irAE_barplot2(plot_data, paste0("barplot_pooled_", drug_name, "_ML.pdf"), width = 1.5, height = 1.85, phenotype_type = "")


```


# STable9. Subgroup analysis summary table
```{r}

main_feature <- "CD8T_CM_tripletORp_ratio_pre"

subgroup_feats <- c(
  "CD8T_CM_tripletORp_ratio_pre",
  "irAE_grade",
  "prior_BRAFi_MEKi",
  "prior_ICI",
  "Drug",
  "Sex",
  "OS_time",
  "OS_event"
)

subgroup_datasets <- list(
  HCI002 = read_cohort("HCI002_info_save.csv"),
  HCI001 = read_cohort("HCI001_info_save.csv"),
  Nicolas_FACS = read_cohort("Nicolas_all_info_FACS_save.csv", melanoma_only = TRUE),
  Nicolas_Cytek = read_cohort("Nicolas_all_info_Cytek_save.csv", melanoma_only = TRUE),
  Nicolas_CyTOF = read_cohort("Nicolas_all_info_CyTOF_save.csv", melanoma_only = TRUE),
  Lozano = read_cohort("Lozano_save.csv"),
  Gerber = read_cohort("Gerber_save.csv"),
  Switzerland1 = read_cohort("Switzerland1_save.csv"),
  Switzerland2 = read_cohort("Switzerland2_save.csv"),
  Switzerland3 = read_cohort("Switzerland3_save.csv")
)

subgroup_datasets <- lapply(subgroup_datasets, ensure_feats, feats = subgroup_feats)
subgroup_datasets <- lapply(subgroup_datasets, function(df) df[, subgroup_feats])
subgroup_datasets <- lapply(subgroup_datasets, make_severe_irAE, out_col = pheno_feat)

standardize_subgroup_types <- function(df) {
  df %>%
    mutate(
      prior_BRAFi_MEKi = as.character(prior_BRAFi_MEKi),
      prior_ICI = as.character(prior_ICI),
      Drug = as.character(Drug),
      Sex = as.character(Sex),
      irAE_grade = suppressWarnings(as.numeric(irAE_grade)),
      severe_irAE = suppressWarnings(as.numeric(severe_irAE)),
      OS_time = suppressWarnings(as.numeric(OS_time)),
      OS_event = suppressWarnings(as.numeric(OS_event))
    )
}

subgroup_datasets <- lapply(subgroup_datasets, standardize_subgroup_types)

subgroup_df <- bind_rows(
  transform(subgroup_datasets$HCI002, cohort = "HCI002"),
  transform(subgroup_datasets$HCI001, cohort = "HCI001"),
  transform(subgroup_datasets$Nicolas_FACS, cohort = "Nicolas_FACS"),
  transform(subgroup_datasets$Nicolas_Cytek, cohort = "Nicolas_Cytek"),
  transform(subgroup_datasets$Switzerland1, cohort = "Switzerland1"),
  transform(subgroup_datasets$Switzerland2, cohort = "Switzerland2"),
  transform(subgroup_datasets$Switzerland3, cohort = "Switzerland3")
) %>%
  mutate(
    Sex = recode(Sex, M = "Male", F = "Female"),
    prior_ICI = recode(as.character(prior_ICI), `0` = "no", `1` = "yes")
  )

safe_or <- function(df, outcome, pred_bin = "pred_bin") {
  if (nrow(df) < 5) return(tibble(OR = NA_real_, OR_p = NA_real_))
  if (n_distinct(df[[pred_bin]]) < 2 || n_distinct(df[[outcome]]) < 2) {
    return(tibble(OR = NA_real_, OR_p = NA_real_))
  }

  fit <- glm(reformulate(pred_bin, response = outcome), data = df, family = binomial())

  broom::tidy(fit, exponentiate = TRUE, conf.int = FALSE) %>%
    filter(term == paste0(pred_bin, "High")) %>%
    transmute(OR = estimate, OR_p = p.value) %>%
    { if (nrow(.) == 0) tibble(OR = NA_real_, OR_p = NA_real_) else . }
}

safe_hr <- function(df, time_col, event_col, pred_bin = "pred_bin") {
  if (nrow(df) < 5) return(tibble(HR = NA_real_, HR_p = NA_real_))
  if (n_distinct(df[[pred_bin]]) < 2 || n_distinct(df[[event_col]]) < 2) {
    return(tibble(HR = NA_real_, HR_p = NA_real_))
  }

  fit <- coxph(as.formula(paste0("Surv(", time_col, ", ", event_col, ") ~ ", pred_bin)), data = df)

  broom::tidy(fit, exponentiate = TRUE, conf.int = FALSE) %>%
    filter(grepl(paste0("^", pred_bin), term)) %>%
    slice(1) %>%
    transmute(HR = estimate, HR_p = p.value) %>%
    { if (nrow(.) == 0) tibble(HR = NA_real_, HR_p = NA_real_) else . }
}

make_subgroup_table <- function(dat, predictor_num = main_feature) {
  dat0 <- dat %>% filter(!is.na(.data[[predictor_num]]))

  subgroup_specs <- list(
    prior_BRAFi_MEKi = c("no", "yes"),
    prior_ICI = c("no", "yes"),
    Drug = sort(unique(na.omit(dat0$Drug))),
    Sex = sort(unique(na.omit(dat0$Sex)))
  )

  rows <- list()

  for (var in names(subgroup_specs)) {
    for (lvl in subgroup_specs[[var]]) {
      df_sub <- dat0 %>%
        filter(
          !is.na(.data[[var]]),
          .data[[var]] == lvl,
          !is.na(.data[[pheno_feat]])
        )

      subgroup_median <- median(df_sub[[predictor_num]], na.rm = TRUE)

      df_sub <- df_sub %>%
        mutate(
          pred_bin = factor(
            ifelse(.data[[predictor_num]] > subgroup_median, "High", "Low"),
            levels = c("Low", "High")
          )
        )

      rows[[length(rows) + 1]] <- tibble(
        subgroup_var = var,
        subgroup_level = as.character(lvl),
        n = nrow(df_sub)
      ) %>%
        bind_cols(
          safe_or(df_sub, pheno_feat),
          safe_hr(df_sub, "OS_time", "OS_event") %>% rename(OS_HR = HR, OS_p = HR_p)
        )
    }
  }

  bind_rows(rows) %>%
    arrange(
      factor(subgroup_var, levels = c("prior_BRAFi_MEKi", "prior_ICI", "Drug", "Sex")),
      subgroup_level
    )
}

subgroup_results <- make_subgroup_table(subgroup_df)
write.csv(subgroup_results, file.path(data_dir,"merged_cohorts", "Pooled_analysis_subgroup_OR_OS.csv"), row.names = FALSE)

```


# Load data for SFigure 12A-E.
```{r}

HCI002_all_info <- readr::read_csv(
  file.path(data_dir, "HCI002_all_info.csv"),
  name_repair = "minimal"
)

pheno_all_info_HCI002 <- HCI002_all_info[, c(1:19, 25892:25894)] %>%
  merge(
    HCI002_data[, c("patient_therapy", "integrated", "irAE_grade")],
    by = "patient_therapy",
    all.x = TRUE
  ) %>%
  distinct(patient_therapy, .keep_all = TRUE)
irAE_grade_vec <- pheno_all_info_HCI002$irAE_grade
pheno_all_info_HCI002$`irAE Overall Assessment` <- dplyr::case_when(
  is.na(irAE_grade_vec) ~ NA_character_,
  irAE_grade_vec >= 3  ~ "Severe",
  irAE_grade_vec == 2  ~ "Moderate",
  irAE_grade_vec == 1  ~ "Mild",
  irAE_grade_vec == 0  ~ "None"
)

feature_test <- "integrated"
```


# SFigure 12A. Integrated score across irAE severity groups.
```{r}
plot_df <- pheno_all_info_HCI002 %>%
  transmute(
    group = factor(
      `irAE Overall Assessment`,
      levels = c("None", "Mild", "Moderate", "Severe")
    ),
    value = integrated
  ) %>%
  drop_na()

p <- ggplot(plot_df, aes(x = group, y = value, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, color = "black") +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.6) +
  labs(x = NULL, y = "Score") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic() +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 47, hjust = 1),
    legend.position = "none"
  )

pdf(
  file.path(result_figure_dir, "boxplot_AUC_compare_irAE_groups_HCI002_IntegrateML.pdf"),
  onefile = FALSE,
  width = 2,
  height = 3
)
print(p)
dev.off()

wilcox.test(
  plot_df$value[plot_df$group == "Severe"],
  plot_df$value[plot_df$group != "Severe"]
)
```

# SFigure 12B. AUC comparison across HCI002 clinical subgroups.
```{r}
compute_auc_summary <- function(df, predictor_col, outcome_col = "Phenotype") {
  auc_df <- df %>%
    select(outcome = all_of(outcome_col), predictor = all_of(predictor_col)) %>%
    drop_na()

  if (nrow(auc_df) < 5 || length(unique(auc_df$outcome)) < 2) {
    return(tibble(AUC = NA_real_, AUC_low = NA_real_, AUC_up = NA_real_, p_value = NA_real_, n = nrow(auc_df)))
  }

  roc_obj <- pROC::roc(
    response = auc_df$outcome,
    predictor = auc_df$predictor,
    direction = "<",
    levels = c(0, 1),
    quiet = TRUE
  )

  auc_ci <- pROC::ci.auc(roc_obj)

  tibble(
    AUC = as.numeric(auc_ci[2]),
    AUC_low = as.numeric(auc_ci[1]),
    AUC_up = as.numeric(auc_ci[3]),
    p_value = verification::roc.area(auc_df$outcome, auc_df$predictor)$p.value,
    n = nrow(auc_df)
  )
}

plot_auc_bar <- function(plot_df, file_name) {
  p <- ggplot(plot_df, aes(x = label, y = AUC, fill = label)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = AUC_low, ymax = AUC_up), width = 0.2, color = "black", linewidth = 0.5) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey", linewidth = 0.5) +
    labs(x = NULL, y = "AUC") +
    coord_cartesian(ylim = c(0, 1)) +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      axis.text.x = element_text(angle = 47, hjust = 1),
      legend.position = "none"
    )

  pdf(file.path(result_figure_dir, file_name), onefile = FALSE, width = 2, height = 3.5)
  print(p)
  dev.off()
}

subgroup_df <- pheno_all_info_HCI002 %>%
  transmute(
    Feature = .data[[feature_test]],
    Phenotype = IRAEactionable01,
    Drug = Drug,
    ICB_exposure = case_when(
      `ICI Naïve?` == "Yes" ~ "ICB-naive",
      `ICI Naïve?` == "No" ~ "ICB-experienced",
      TRUE ~ NA_character_
    ),
    Stage = ifelse(grepl("IV", `Baseline Stage / TNM`), "Stage IV", "Stage I-III"),
    TherapyType = case_when(
      TherapyType == "Active" ~ "Active treatment",
      TherapyType == "Adjuvant" ~ "Adjuvant",
      TRUE ~ TherapyType
    )
  )

subgroup_specs <- list(
  Drug = c("PD1/PDL1", "CTLA4-containing"),
  ICB_exposure = c("ICB-naive", "ICB-experienced"),
  Stage = c("Stage I-III", "Stage IV"),
  TherapyType = c("Active treatment", "Adjuvant")
)

auc_subgroup_results <- bind_rows(lapply(names(subgroup_specs), function(group_col) {
  bind_rows(lapply(subgroup_specs[[group_col]], function(group_level) {
    subgroup_df %>%
      filter(.data[[group_col]] == group_level) %>%
      compute_auc_summary(predictor_col = "Feature") %>%
      mutate(
        subgroup = group_col,
        label = group_level,
        .before = 1
      )
  }))
}))

for (group_col in names(subgroup_specs)) {
  plot_df <- auc_subgroup_results %>%
    filter(subgroup == group_col) %>%
    mutate(label = paste0(label, "\n(n=", n, ")"))

  plot_auc_bar(
    plot_df,
    paste0("barplot_AUC_compare_subgroups_", group_col, "_HCI002.pdf")
  )
}

```

# SFigure 12C. Integrated model AUC across irAE subtypes.
```{r}
irae_subtypes <- c("IRAE_GI01", "IRAE_HB01", "IRAE_skin01", "IRAE_MSK01")

irae_labels <- c(
  IRAE_GI01 = "Gastrointestinal",
  IRAE_HB01 = "Hepatobiliary",
  IRAE_skin01 = "Skin",
  IRAE_MSK01 = "Musculoskeletal"
)

auc_irae_df <- bind_rows(lapply(irae_subtypes, function(irae_type) {
  pheno_all_info_HCI002 %>%
    transmute(Phenotype = .data[[irae_type]], Feature = .data[[feature_test]]) %>%
    compute_auc_summary(predictor_col = "Feature") %>%
    mutate(irAE_type = irae_labels[[irae_type]], .before = 1)
}))

p <- ggplot(auc_irae_df, aes(x = irAE_type, y = AUC, fill = irAE_type)) +
  geom_col(width = 0.7) +
  geom_errorbar(aes(ymin = AUC_low, ymax = AUC_up), width = 0.2, color = "black", linewidth = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey", linewidth = 0.5) +
  labs(x = NULL, y = "AUC") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic() +
  theme(
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 47, hjust = 1),
    legend.position = "none"
  )

pdf(
  file.path(result_figure_dir, "barplot_AUC_compare_irAE_types_HCI002_IntegrateML.pdf"),
  onefile = FALSE,
  width = 2,
  height = 3.5
)
print(p)
dev.off()
```

# SFigure 12D,E. Association between integrated score and baseline clinical variables in HCI002.
```{r}
plot_df <- pheno_all_info_HCI002 %>%
  transmute(
    Score = .data[[feature_test]],
    Age = `Age at C1`,
    TMB = `TMB (mutation /mb)`,
    PDL1_TPS = `PD-L1 TPS`,
    PDL1_CPS = `PD-L1 CPS`,
    Gender = Gender,
    Stage = case_when(
      grepl("IV", `Baseline Stage / TNM`) ~ "Stage IV",
      grepl("I|II|III", `Baseline Stage / TNM`) ~ "Stage I-III",
      TRUE ~ NA_character_
    ),
    ICB_naive = `ICI Naïve?`,
    Drug = Drug,
    BRAF = case_when(
      is.na(`BRAF Mutation`) ~ "Unknown",
      `BRAF Mutation` == "Wildtype" ~ "Wildtype",
      grepl("V600E", `BRAF Mutation`) ~ "V600E",
      grepl("V600K", `BRAF Mutation`) ~ "V600K",
      TRUE ~ "Other mut."
    )
  )

continuous_vars <- c(
  Age = "Age",
  TMB = "TMB (mut. / Mb)",
  PDL1_TPS = "PD-L1 TPS (%)",
  PDL1_CPS = "PD-L1 CPS (%)"
)

for (var_name in names(continuous_vars)) {
  p <- ggplot(plot_df, aes(x = .data[[var_name]], y = Score)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top") +
    labs(x = continuous_vars[[var_name]], y = "Score") +
    theme_classic() +
    theme(axis.text = element_text(color = "black"))

  ggsave(
    file.path(result_figure_dir, paste0("scatter_HCI002_", feature_test, "_vs_", var_name, ".pdf")),
    plot = p,
    width = 2.1,
    height = 2.4
  )
}

categorical_vars <- c("Gender", "Stage", "ICB_naive", "Drug", "BRAF")

for (var_name in categorical_vars) {
  p <- ggplot(plot_df, aes(x = .data[[var_name]], y = Score)) +
    geom_boxplot(outlier.shape = NA, fill = "lightgray", color = "black") +
    geom_jitter(width = 0.2, size = 1, aes(color = .data[[var_name]])) +
    labs(x = var_name, y = "Score") +
    theme_classic() +
    theme(
      axis.text = element_text(color = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )

  ggsave(
    file.path(result_figure_dir, paste0("barplot_HCI002_", feature_test, "_vs_", var_name, ".pdf")),
    plot = p,
    width = 1.5,
    height = 3
  )
}

wilcox_results <- bind_rows(lapply(categorical_vars, function(var_name) {

  test_df <- plot_df %>%
    select(Score, Group = all_of(var_name)) %>%
    drop_na()

  if (var_name == "BRAF") {
    test_df <- test_df %>%
      mutate(
        Group = ifelse(Group == "Wildtype", "Wildtype", "Non-Wildtype")
      )
  }

  p_value <- if (n_distinct(test_df$Group) == 2) {
    wilcox.test(Score ~ Group, data = test_df)$p.value
  } else {
    NA_real_
  }

  tibble(
    Variable = var_name,
    P_value = p_value
  )
}))

wilcox_results
```
